### 2022-07-22 source/paper.tex
title

\documentclass[11pt, a4paper]{article}
\usepackage{authblk}

\begin{document}

\author[1,*]{Jemma Shipton}
\author[2]{Thomas M. Bendall}
\author[3]{So many others}

\affil[1]{Department of Mathematics and Statistics, University of Exeter}
\affil[2]{Met Office}
\affil[*]{Correspondence to: \texttt{j.shipton@exeter.ac.uk}}

\title{Gusto: a toolkit for compatible finite element dynamical cores}
\date{July 2022}

\maketitle

\end{document}

### 2023-05-31 source/paper.tex
making a start

@@
 \usepackage{authblk}
+
+\usepackage{amsmath}
+\usepackage{amssymb}
+
+\usepackage{natbib}
+\bibliographystyle{plainnat}
+
+\def\MM#1{\boldsymbol{#1}}
 
@@
 \title{Gusto: a toolkit for compatible finite element dynamical cores}
-\date{July 2022}
+\date{}
 
@@
 
+\section{Introduction}
+
+\subsection{Overview}
+
+\subsection{Motivation}
+Compatible finite elements provide a route to stable discretisations
+of the partial differential equations governing geophysical fluids
+which also have the correct wave propagation properties on non-uniform
+grids. While the wave propagation and required conservation properties
+guide the choice of finite element function spaces for the
+discretisation, there are many options for both the spatial and
+temporal discretisation and it is not clear which will be best for any
+particular problem. Gusto provides a flexible, easy-to-use toolkit of
+options for rapid prototyping of different spatial and temporal
+discretisations for geophysical fluid dynamics equations relevant to
+numerical weather and climate prediction.
+
+\section{Governing Equations}
+
+\subsection{Shallow Water Equations}
+The shallow water equations are commonly used for testing numerical
+algorithms and for exploring geophysical fluid dynamics concepts under
+simplified conditions. They are useful in both contexts since they are
+the simplest set of equations that model motion on both the slow,
+geostrophic timescale and the much faster timescale of the
+inertia-gravity waves. This timescale separation poses challenges to
+the numerical discretisation and also provides a rich enough range of
+dynamics to explain many of the phenomena observed in the atmosphere
+and oceans \citep{zeitlin2018geophysical}.
+
+The shallow water equations describe the flow of a shallow layer of
+fluid subject to gravitational and, optionally, rotational forces. The
+prognostic variables are the two horizontal velocity components
+$\MM{u} = (u, v)$ and the fluid depth $D = H + h - b$ where $H$ is the
+undisturbed depth of the fluid layer, $h$ is the perturbation to the
+fluid depth and $b$ is the bottom topography, if present. The
+equations are:
+
+\begin{align*}
+  \frac{\partial\MM{u}}{\partial t} + \MM{u}\cdot\nabla\MM{u} + f\MM{k}\times\MM{u} + g\nabla (D+b) &= 0, \\
+  \frac{\partial D}{\partial t} + \nabla\cdot(\MM{u}D) &= 0,
+\end{align*}
+where $f$ is the Coriolis parameter, $\MM{k}$ the vertical vector normal
+to the domain and $g$ the acceleration due to gravity. The nonlinear velocity
+advection term can be replaced by the sum of a vorticity-based term
+and the gradient of the kinetic energy. This is known as the vector
+invariant form and it enables the construction of vorticity based
+advection schemes that lead to conservation of energy and/or enstrophy
+(**refs). However, this form can exhibit numerical instabilities and
+it is not yet clear which is preferable so in Gusto we offer both
+forms of the equations.
+
+The finite element discretisation of these equations is formed by
+choosing appropriate function spaces for the prognostic variables,
+multiplying by test functions from these function spaces and
+integrating over the domain. In addition to this, we integrate by
+parts when the inter-element continuity of the basis functions is not
+sufficient to define the result of applying the differential operator
+directly. In line with the theory outlined in section \ref{}, we
+choose $\MM{u} \in \mathbb{V}_1$ and $D \in \mathbb{V}_2$ with
+$\mathbb{V}_1 \subset H(\text{div})$ and $\mathbb{V}_2 \subset L^2$. The
+corresponding test functions are denoted by $\MM{w}$ and $\phi$. This gives
+
+\begin{align*}
+  \int_\Omega\MM{w}\cdot\frac{\partial\MM{u}}{\partial t}dV + \MM{u}\cdot\nabla\MM{u} + f\MM{k}\times\MM{u} - \int_\Omega \nabla\cdot\MM{w} g(D+b) dV &= 0, \forall \MM{w}\in\mathbb{V}_1 \\
+  \int_\Omega\phi\frac{\partial D}{\partial t}dV + \int_\Omega\phi\nabla\cdot(\MM{u}D) dV &= 0, \forall \phi\in\mathbb{V}_2
+\end{align*}
+
+\subsection{Compressible Euler Equations}
+
+\subsection{Incompressible Boussinesq Equations}
+
+\section{Results}
+The aim of this section is to present a selection of results that
+demonstrate the tools available in Gusto, such as the choice of grids,
+different discretisation options, timestepping schemes and
+physics. The numerical properties of these configurations
+(convergence, stability, accuracy and efficiency) have either been
+demonstrated elsewhere, or will be investigated more thoroughly in
+future publications.
+
+\begin{itemize}
+\item grids
+\item FE order
+\item spatial discretisation options
+\item timestepping schemes
+\item physics
+\end{itemize}
+
+\subsection{Shallow water simulations}
+The standard suite of shallow water test cases originally proposed in
+\citet{williamson1992standard}
+
+\section{Discussion}
+
+\section{Code availability}
+
+\bibliography{references}
+
 \end{document}

### 2023-07-13 source/paper.tex
a little bit of rejigging and a reference
progress on writing and references

@@
 
-\section{Introduction}
-
-\subsection{Overview}
+\section{Background}
 
 \subsection{Motivation}
-Compatible finite elements provide a route to stable discretisations
-of the partial differential equations governing geophysical fluids
-which also have the correct wave propagation properties on non-uniform
-grids. While the wave propagation and required conservation properties
-guide the choice of finite element function spaces for the
-discretisation, there are many options for both the spatial and
-temporal discretisation and it is not clear which will be best for any
-particular problem. Gusto provides a flexible, easy-to-use toolkit of
-options for rapid prototyping of different spatial and temporal
-discretisations for geophysical fluid dynamics equations relevant to
-numerical weather and climate prediction.
+\begin{itemize}
+\item route to discretisation of eqns with correct properties on non-uniform grids
+\item these properties are considered essential for dynamical cores
+\item non-uniform grids are considered important for parallel scaling
+\item many choices and research to be done therefore flexible implementation is useful
+\end{itemize}
+
+Compatible finite elements provide a route to stable, consistent
+discretisations of the partial differential equations governing
+geophysical fluids which also have the correct wave propagation
+properties on non-uniform grids and avoid spurious computational
+modes. These properties, along with conservation of mass and
+potentially other quantities such as energy and enstrophy, are
+considered essential for the accuracy of the dynamical core of weather
+and climate models. While the wave propagation and required
+conservation properties guide the choice of finite element function
+spaces for the discretisation, there are many options for both the
+spatial and temporal discretisation and it is not clear which will be
+best for any particular problem. Gusto provides a flexible,
+easy-to-use toolkit of options for rapid prototyping of different
+spatial and temporal discretisations for geophysical fluid dynamics
+equations relevant to numerical weather and climate prediction.
+
+In this article we will describe the current capabilities of Gusto and
+the design features that enable our flexible implementation whereby
+different spatial and temporal discretisations can be applied to a
+range of different equations sets relevant for geophysical fluid
+modelling. In section \ref{sec: governing} we will outline three
+different equation sets, along with their finite element
+discretisations. Then in section \ref{sec: design} we will present the
+code structure of Gusto followed by a detailed example (section
+\ref{sec: FML}) of the Form Manipulation Language that facilitates
+automatic manipulation of the finite element forms within our
+timestepping code. Section \ref{sec: results} contains a selection of
+numerical results that showcases the range of discretisations
+available within Gusto. We summarise our current status and future
+directions in section \ref{sec: summary}.
 
 \section{Governing Equations}
+\label{sec: governing}
+
+The aim of this section is to introduce three governing equation sets
+commonly used in the development of weather and climate models, along
+with enough detail about the finite element discretiations provided by
+Gusto to demonstrate the similarities between them. These similarities
+guide the design of the code structure described in the next section
+and motivate the use of Form Manipulation Labelling (FML) that
+underlies the flexibility of Gusto, a detailed example of which will
+be given in section \ref{sec: FML}.
 
@@
 
-\begin{align*}
+\begin{align}
   \frac{\partial\MM{u}}{\partial t} + \MM{u}\cdot\nabla\MM{u} + f\MM{k}\times\MM{u} + g\nabla (D+b) &= 0, \\
   \frac{\partial D}{\partial t} + \nabla\cdot(\MM{u}D) &= 0,
-\end{align*}
-where $f$ is the Coriolis parameter, $\MM{k}$ the vertical vector normal
-to the domain and $g$ the acceleration due to gravity. The nonlinear velocity
-advection term can be replaced by the sum of a vorticity-based term
-and the gradient of the kinetic energy. This is known as the vector
-invariant form and it enables the construction of vorticity based
-advection schemes that lead to conservation of energy and/or enstrophy
-(**refs). However, this form can exhibit numerical instabilities and
-it is not yet clear which is preferable so in Gusto we offer both
-forms of the equations.
+\end{align}
+where $f$ is the Coriolis parameter, $\MM{k}$ is the vertical vector
+normal to the domain and $g$ is the acceleration due to gravity. The
+nonlinear velocity advection term can be replaced by the sum of a
+vorticity-based term and the gradient of the kinetic energy. This is
+known as the vector invariant form and it enables the construction of
+schemes that conservation energy and/or enstrophy
+\citep{mcrae2014energy, bauer2018energy, wimmer2020energy,
+  wimmer2021energy}. However, this form can exhibit numerical
+instabilities \citep{bell2017numerical} and it is not yet clear which
+form is preferable so in Gusto we offer both forms of the equations.
 
@@
 
-\begin{align*}
-  \int_\Omega\MM{w}\cdot\frac{\partial\MM{u}}{\partial t}dV + \MM{u}\cdot\nabla\MM{u} + f\MM{k}\times\MM{u} - \int_\Omega \nabla\cdot\MM{w} g(D+b) dV &= 0, \forall \MM{w}\in\mathbb{V}_1 \\
+\begin{align}
+  \int_\Omega\MM{w}\cdot\frac{\partial\MM{u}}{\partial t}dV + \int_\Omega\MM{w}\cdot(\MM{u}\cdot\nabla)\MM{u} dV + \int_\Omega\MM{w}\cdot f\MM{k}\times\MM{u}dV - \int_\Omega \nabla\cdot\MM{w} g(D+b) dV &= 0, \forall \MM{w}\in\mathbb{V}_1 \\
   \int_\Omega\phi\frac{\partial D}{\partial t}dV + \int_\Omega\phi\nabla\cdot(\MM{u}D) dV &= 0, \forall \phi\in\mathbb{V}_2
-\end{align*}
+\end{align}
 
 \subsection{Compressible Euler Equations}
+The dynamical core at the heart of any weather or climate model solves
+the rotating compressible Euler equations that model atmospheric flow,
+often with additional approximations such as shallow vs deep
+atmosphere or hydrostatic vs non-hydrostatic flow. These equations can
+be written in terms of
 
-\subsection{Incompressible Boussinesq Equations}
+\subsection{Compressible Boussinesq Equations}
+
+\section{Software Design}
+\label{sec: design}
+
+\begin{itemize}
+\item include schematic
+\item generic example - no FML
+\item lead in to next section... facilitated by FML
+\end{itemize}
+
+\section{Form Manipulation Labelling}
+\label{sec: FML}
 
 \section{Results}
+\label{sec: results}
+
 The aim of this section is to present a selection of results that
@@
 
-\section{Discussion}
+\section{Summary}
+\label{sec: summary}
 

### 2023-07-13 to_do.tex
to do list got longer

To Do:

Code:
* compressible Boussinesq equations class and examples
* general mixed solve for the compressible Euler equations
* merge transport wrapper PR and update all examples and scripts
* fix vorticity function space
* check hydrostatic Euler is correct
* better provision of solver options?

Decisions:
* where to store scripts and how to archive them (GitHub repo + Zenodo)
* which simulations to run (see later section)
* will physics be part of this paper?

Website:
* dragon logo still apears in the top tab

Simulations:
* shallow water
 - lowest order and next-to-lowest order
 - cubed sphere and icosahedral sphere meshes
 - range of timestepping schemes: show convergence plot of W2
 - vector invariant vs advective form

Writing:
* describe advantages of building on top of Firedrake
* compressible Euler section
* compressible Boussinesq section
* Gusto design
* FML example
 - generate code
* future directions:
 - structure preserving discretisations / vorticity based discretisations
 - other equation sets
 - PDC
 - ML?
 - coupled models (mention Thetis and / or Icepack?)

Figures:
* cubed sphere and icosahedral sphere grids
* extruded mesh?
* DOFs?
* convergence plot of SW W2 (and W3?) for a range of timestepping schemes
* snapshots of W5 (after 15 and 50 days), W6 and Galewsky
* vertical slice gravity wave final state
* vertical slice flow over orography
* dense bubble
* rising bubble

### 2023-07-13 source/paper.tex
some progress

@@
 \def\MM#1{\boldsymbol{#1}}
+\newcommand{\pp}[2]{\frac{\partial #1}{\partial #2}}
 
@@
 
-\section{Background}
+\section{Introduction}
 
@@
 \begin{itemize}
-\item route to discretisation of eqns with correct properties on non-uniform grids
-\item these properties are considered essential for dynamical cores
-\item non-uniform grids are considered important for parallel scaling
-\item many choices and research to be done therefore flexible implementation is useful
+\item many choices and research to be done - Colin says so in his Acta Numerica paper! - therefore flexible implementation is useful
 \end{itemize}
@@
 discretisations of the partial differential equations governing
-geophysical fluids which also have the correct wave propagation
-properties on non-uniform grids and avoid spurious computational
-modes. These properties, along with conservation of mass and
+geophysical fluids which also have the correct wave propagation speeds
+and avoid spurious computational modes, even on non-uniform
+grids. These properties, along with conservation of mass and
 potentially other quantities such as energy and enstrophy, are
 considered essential for the accuracy of the dynamical core of weather
-and climate models. While the wave propagation and required
-conservation properties guide the choice of finite element function
-spaces for the discretisation, there are many options for both the
+and climate models \citep{staniforth2012horizontal}. Achieving these
+properties on a non-uniform grid is important because of parallel
+scaling. While the required properties of the discretisation guide the
+choice of finite element spaces, there are many options for both the
 spatial and temporal discretisation and it is not clear which will be
 best for any particular problem. Gusto provides a flexible,
-easy-to-use toolkit of options for rapid prototyping of different
-spatial and temporal discretisations for geophysical fluid dynamics
-equations relevant to numerical weather and climate prediction.
+extensible, easy-to-use toolkit of options for rapid prototyping of
+different spatial and temporal discretisations based on the compatible
+finite element framework applied to geophysical fluid dynamics
+equations relevant to numerical weather and climate prediction. In
+this paper we introduce Gusto to the community, summarise its current
+capabilities and outline our vision for future development.
+
+\subsection{Background}
+
+\begin{itemize}
+\item overall theory is described in \citet{gibson2019compatible, cotter2023compatible}
+\item Gusto has grown out of code described in \citet{natale2016compatible, cotter2012mixed, bendall2019recovered, bendall2020compatible, yamazaki2017vertical, shipton2018higher}
+\item links to LFRic - we mention the Gung Ho project on the Gusto website so should describe that link here
+\end{itemize}
 
@@
 different spatial and temporal discretisations can be applied to a
-range of different equations sets relevant for geophysical fluid
+range of different equation sets relevant for weather and climate
 modelling. In section \ref{sec: governing} we will outline three
@@
 code structure of Gusto followed by a detailed example (section
-\ref{sec: FML}) of the Form Manipulation Language that facilitates
-automatic manipulation of the finite element forms within our
+\ref{sec: FML}) of the Form Manipulation Language (FML) that facilitates
+symbolic manipulation of the finite element forms within our
 timestepping code. Section \ref{sec: results} contains a selection of
@@
 \label{sec: governing}
-
 The aim of this section is to introduce three governing equation sets
@@
 underlies the flexibility of Gusto, a detailed example of which will
-be given in section \ref{sec: FML}.
+be given in section \ref{sec: FML}. The convention in Gusto is to
+write the equations in residual form, meaning that all the terms
+appear on the left hand side. We will follow this convention below in
+contrast to many papers which present the equations with the pressure
+gradient term and any forcing terms on the right hand side.
 
@@
 \begin{align}
-  \frac{\partial\MM{u}}{\partial t} + \MM{u}\cdot\nabla\MM{u} + f\MM{k}\times\MM{u} + g\nabla (D+b) &= 0, \\
-  \frac{\partial D}{\partial t} + \nabla\cdot(\MM{u}D) &= 0,
+  \pp{\MM{u}}{t} + \MM{u}\cdot\nabla\MM{u} + f\hat{\MM{k}}\times\MM{u} + g\nabla (D+b) &= 0, \\
+  \pp{D}{t} + \nabla\cdot(D\MM{u}) &= 0,
 \end{align}
-where $f$ is the Coriolis parameter, $\MM{k}$ is the vertical vector
-normal to the domain and $g$ is the acceleration due to gravity. The
+where $f$ is the Coriolis parameter, $\hat{\MM{k}}$ is the unit vector
+pointing upwards and $g$ is the acceleration due to gravity. The
 nonlinear velocity advection term can be replaced by the sum of a
@@
 \begin{align}
-  \int_\Omega\MM{w}\cdot\frac{\partial\MM{u}}{\partial t}dV + \int_\Omega\MM{w}\cdot(\MM{u}\cdot\nabla)\MM{u} dV + \int_\Omega\MM{w}\cdot f\MM{k}\times\MM{u}dV - \int_\Omega \nabla\cdot\MM{w} g(D+b) dV &= 0, \forall \MM{w}\in\mathbb{V}_1 \\
-  \int_\Omega\phi\frac{\partial D}{\partial t}dV + \int_\Omega\phi\nabla\cdot(\MM{u}D) dV &= 0, \forall \phi\in\mathbb{V}_2
+  \int_\Omega\MM{w}\cdot\pp{\MM{u}}{t}dV + \int_\Omega\MM{w}\cdot(\MM{u}\cdot\nabla)\MM{u} dV + \int_\Omega\MM{w}\cdot f\hat{\MM{k}}\times\MM{u}dV - \int_\Omega \nabla\cdot\MM{w} g(D+b) dV &= 0, \forall \MM{w}\in\mathbb{V}_1 \\
+  \int_\Omega\phi\pp{D}{t}dV + \int_\Omega\phi\nabla\cdot(D\MM{u}) dV &= 0, \forall \phi\in\mathbb{V}_2
+\end{align}
+
+\subsection{Compressible Boussinesq Equations}
+The compressible Boussinesq equations are commonly used in ocean
+modelling. They are useful from a numerical discretisation perspective
+because they incorporate compressibility effects without the
+complexity of discretising the nonlinear pressure gradient terms that
+we shall see in the compressible Euler equations. We write these
+equations in terms of the prognostic variables $\MM{u}$, $b$ and
+$p$ where $\MM{u}=(u, v, w)$ is now a three dimensional velocity
+field, $b$ is the fluid buoyancy and $p$ is pressure. In terms of
+these variables, the equations are:
+
+\begin{align}
+  \pp{\MM{u}}{t} + 
+  (\MM{u}\cdot\nabla)\MM{u} +
+  f\hat{\MM{k}}\times \MM{u} + \nabla p + b\hat{\MM{k}} & = 0, \\
+  \pp{p}{t} + c^2\nabla\cdot\MM{u} & = 0, \\
+  \pp{b}{t} + \MM{u}\cdot\nabla(b) & = 0, \\
 \end{align}
@@
 often with additional approximations such as shallow vs deep
-atmosphere or hydrostatic vs non-hydrostatic flow. These equations can
-be written in terms of
+atmosphere or hydrostatic vs non-hydrostatic flow (** mention this
+here or at the end of the section??). We write these equations in
+terms of the prognostic variables $\MM{u}$, $\rho$ and $\theta$ where
+$\MM{u}=(u, v, w)$ is now a three dimensional velocity field, $\rho$
+is the fluid density and $\theta$ is the potential temperature. In
+terms of these variables, the equations are:
 
-\subsection{Compressible Boussinesq Equations}
+\begin{align}
+  \pp{\MM{u}}{t} + 
+  (\MM{u}\cdot\nabla)\MM{u} +
+  f\hat{\MM{k}}\times \MM{u} + c_p\theta\nabla \Pi + g\hat{\MM{k}} & = 0, \\
+  \pp{\rho}{t} + \nabla\cdot(\MM{u}\rho) & = 0, \\
+  \pp{\theta}{t} + \MM{u}\cdot\nabla\theta & = 0, \\
+  \Pi^{(1-\kappa)/\kappa} = \frac{R}{p_0}\rho\theta, & 
+\end{align}
 
@@
 \label{sec: design}
+Gusto is built on top of the Firedrake package, an automated system
+for solving partial differential equations using the finite element
+method. This gives many advantages, not least that it provides the
+finite element machinery required to translate the discretisations
+written in the previous section in to the matrix-vector systems that
+we need to solve. This is done using the Unified Form Language (UFL),
+a domain specific language that enables the definition of the finite
+element forms within Python. The other packages that Firedrake depends
+on are summarised in \citet{davies2022towards} ** mention extruded
+meshes and PETSc?
 
@@
 \label{sec: results}
-
 The aim of this section is to present a selection of results that
 demonstrate the tools available in Gusto, such as the choice of grids,
-different discretisation options, timestepping schemes and
-physics. The numerical properties of these configurations
-(convergence, stability, accuracy and efficiency) have either been
-demonstrated elsewhere, or will be investigated more thoroughly in
-future publications.
-
-\begin{itemize}
-\item grids
-\item FE order
-\item spatial discretisation options
-\item timestepping schemes
-\item physics
-\end{itemize}
+order of finite element fucntion spaces, different spatial
+discretisation options, timestepping schemes and physics. The
+numerical properties of these configurations (convergence, stability,
+accuracy and efficiency) have either been demonstrated elsewhere, or
+will be investigated more thoroughly in future publications.
 
@@
 The standard suite of shallow water test cases originally proposed in
-\citet{williamson1992standard}
+\citet{williamson1992standard} provides a common starting point for
+testing new discretisations. This test suite is usually augmented by
+the unstable barotropic jet described in \citet{galewsky2004initial}
+(**and linear tests?)
 

### 2023-07-25 source/paper.tex
small changes

@@
   (\MM{u}\cdot\nabla)\MM{u} +
-  f\hat{\MM{k}}\times \MM{u} + \nabla p + b\hat{\MM{k}} & = 0, \\
+  2\MM{\Omega}\times \MM{u} + \nabla p + b\hat{\MM{k}} & = 0, \\
   \pp{p}{t} + c^2\nabla\cdot\MM{u} & = 0, \\
-  \pp{b}{t} + \MM{u}\cdot\nabla(b) & = 0, \\
+  \pp{b}{t} + \MM{u}\cdot\nabla b & = 0,
 \end{align}
+where $\MM{\Omega}$ is the rotation vector of the domain,
+$\hat{\MM{k}}$ is the unit vector pointing upwards and $c$ is the
+acoustic wave speed.
 
@@
   (\MM{u}\cdot\nabla)\MM{u} +
-  f\hat{\MM{k}}\times \MM{u} + c_p\theta\nabla \Pi + g\hat{\MM{k}} & = 0, \\
+  2\MM{\Omega}\times \MM{u} + c_p\theta\nabla \Pi + g\hat{\MM{k}} & = 0, \\
   \pp{\rho}{t} + \nabla\cdot(\MM{u}\rho) & = 0, \\
@@
 \end{align}
+
+\subsection{}
 
@@
 demonstrate the tools available in Gusto, such as the choice of grids,
-order of finite element fucntion spaces, different spatial
-discretisation options, timestepping schemes and physics. The
-numerical properties of these configurations (convergence, stability,
-accuracy and efficiency) have either been demonstrated elsewhere, or
-will be investigated more thoroughly in future publications.
+order of finite element spaces, different spatial discretisation
+options, timestepping schemes and physics. The numerical properties of
+these configurations (convergence, stability, accuracy and efficiency)
+have either been demonstrated elsewhere, or will be investigated more
+thoroughly in future publications.
 

### 2023-12-06 to_do.tex
to do list changes from ages ago

@@
 * general mixed solve for the compressible Euler equations
-* merge transport wrapper PR and update all examples and scripts
-* fix vorticity function space
+DONE* merge transport wrapper PR and update all examples and scripts
+DONE?* fix vorticity function space
 * check hydrostatic Euler is correct
+* check SUPG is correct
 * better provision of solver options?
@@
 * convergence plot of SW W2 (and W3?) for a range of timestepping schemes
-* snapshots of W5 (after 15 and 50 days), W6 and Galewsky
+* snapshots of W5 (after 15 and 50 days), W6 and Galewsky [suggest: W5 and W6 at both orders; Galewsky on both icosahedral and cubed spheres]
 * vertical slice gravity wave final state
@@
 * dense bubble
-* rising bubble
+* something in 3D

### 2024-02-20 source/paper.tex
a bit of writing

@@
 and climate models \citep{staniforth2012horizontal}. Achieving these
-properties on a non-uniform grid is important because of parallel
-scaling. While the required properties of the discretisation guide the
+properties on a non-orthogonal grid is important because of parallel
+scaling**. While the required properties of the discretisation guide the
 choice of finite element spaces, there are many options for both the
@@
 \end{itemize}
+
+The finite element method represents fields on a mesh made of
+non-overlapping, conforming cells, or elements. Within each element,
+the field is represented as a sum of coefficients multiplied by
+polynomial basis functions. The finite element space is specified by
+the degree of the polynomial basis functions and the continuity of
+these functions between the elements. The partial differential
+equations used in weather and climate modelling have multiple
+prognostic fields and it is not necessary, or indeed desirable, to use
+the same finite element function space for each field. Using different
+finite element function spaces for each field gives what is known as a
+mixed finite element discretisation. If the finite element function
+spaces are chosen so that
 
@@
 
-The finite element discretisation of these equations is formed by
-choosing appropriate function spaces for the prognostic variables,
-multiplying by test functions from these function spaces and
-integrating over the domain. In addition to this, we integrate by
-parts when the inter-element continuity of the basis functions is not
-sufficient to define the result of applying the differential operator
-directly. In line with the theory outlined in section \ref{}, we
-choose $\MM{u} \in \mathbb{V}_1$ and $D \in \mathbb{V}_2$ with
-$\mathbb{V}_1 \subset H(\text{div})$ and $\mathbb{V}_2 \subset L^2$. The
-corresponding test functions are denoted by $\MM{w}$ and $\phi$. This gives
-
-\begin{align}
-  \int_\Omega\MM{w}\cdot\pp{\MM{u}}{t}dV + \int_\Omega\MM{w}\cdot(\MM{u}\cdot\nabla)\MM{u} dV + \int_\Omega\MM{w}\cdot f\hat{\MM{k}}\times\MM{u}dV - \int_\Omega \nabla\cdot\MM{w} g(D+b) dV &= 0, \forall \MM{w}\in\mathbb{V}_1 \\
-  \int_\Omega\phi\pp{D}{t}dV + \int_\Omega\phi\nabla\cdot(D\MM{u}) dV &= 0, \forall \phi\in\mathbb{V}_2
-\end{align}
-
 \subsection{Compressible Boussinesq Equations}
@@
   \pp{\theta}{t} + \MM{u}\cdot\nabla\theta & = 0, \\
-  \Pi^{(1-\kappa)/\kappa} = \frac{R}{p_0}\rho\theta, & 
+  \Pi^{(1-\kappa)/\kappa} & = \frac{R}{p_0}\rho\theta, & 
 \end{align}
+where $\Pi$ is the Exner pressure, $p_0$ is a reference pressure,
+typically taken to be *, $R$ is ** and $\kappa$ is **.
 
-\subsection{}
+\subsection{Finite Element Discretisation}
+\label{sec: FEM}
+
+Comparing the equation sets given in the previous sections, we see
+several similarities. For example, the prognostic equation for the
+velocity always contains a nonlinear transport term, a term due to
+rotation and a pressure gradient term. The other prognostic fields
+satisfy either an advective or continuity form of the transport
+equation. These similarities remain when we formulate the compatible
+finite element discretisations, as we shall see below.
+
+The finite element discretisation is formed by choosing appropriate
+function spaces for the prognostic variables, multiplying by test
+functions from these function spaces and integrating over the
+domain. In addition to this, we integrate by parts when the
+inter-element continuity of the basis functions is not sufficient to
+define the result of applying the differential operator directly.
+
+** Don't do whole equation set - go term by term and try to write as
+few equations as possible.
+
+In line with the theory outlined in section \ref{}, we
+choose $\MM{u} \in \mathbb{V}_1$ and $D \in \mathbb{V}_2$ with
+$\mathbb{V}_1 \subset H(\text{div})$ and $\mathbb{V}_2 \subset L^2$. The
+corresponding test functions are denoted by $\MM{w}$ and $\phi$. This gives
 
@@
 finite element machinery required to translate the discretisations
-written in the previous section in to the matrix-vector systems that
+written in the previous section in to the matrix-vector equations that
 we need to solve. This is done using the Unified Form Language (UFL),
@@
 element forms within Python. The other packages that Firedrake depends
-on are summarised in \citet{davies2022towards} ** mention extruded
-meshes and PETSc?
+on are summarised in \citet{davies2022towards} but we note that we
+make particular use of higher order mesh representation on curved
+manifolds (**ref), extruded meshes (**ref) and the PETSc library
+solver options (**ref).
+
+Prognostic equation sets are each defined in a python class, making
+use of the common finite element forms for the transport and
+difffusion terms.
+
+** mention physics source terms?
 
@@
 \item generic example - no FML
-\item lead in to next section... facilitated by FML
+\item lead in to next section... facilitated by FML i.e. timesteppers don't care what equations they're given because forms are labelled
 \end{itemize}

### 2024-02-20 source/paper.tex
putting the paper into the GMD template was very motivating

@@
-\documentclass[11pt, a4paper]{article}
-\usepackage{authblk}
+%% Copernicus Publications Manuscript Preparation Template for LaTeX Submissions
+%% ---------------------------------
+%% This template should be used for copernicus.cls
+%% The class file and some style files are bundled in the Copernicus Latex Package, which can be downloaded from the different journal webpages.
+%% For further assistance please contact Copernicus Publications at: production@copernicus.org
+%% https://publications.copernicus.org/for_authors/manuscript_preparation.html
 
-\usepackage{amsmath}
-\usepackage{amssymb}
+%% Please use the following documentclass and journal abbreviations for preprints and final revised papers.
 
-\usepackage{natbib}
-\bibliographystyle{plainnat}
+%% 2-column papers and preprints
+\documentclass[journal abbreviation, manuscript]{copernicus}
 
@@
 
+%% Journal abbreviations (please use the same for preprints and final revised papers)
+
+% Advances in Geosciences (adgeo)
+% Advances in Radio Science (ars)
+% Advances in Science and Research (asr)
+% Advances in Statistical Climatology, Meteorology and Oceanography (ascmo)
+% Aerosol Research (ar)
+% Annales Geophysicae (angeo)
+% Archives Animal Breeding (aab)
+% Atmospheric Chemistry and Physics (acp)
+% Atmospheric Measurement Techniques (amt)
+% Biogeosciences (bg)
+% Climate of the Past (cp)
+% DEUQUA Special Publications (deuquasp)
+% Earth Surface Dynamics (esurf)
+% Earth System Dynamics (esd)
+% Earth System Science Data (essd)
+% E&G Quaternary Science Journal (egqsj)
+% EGUsphere (egusphere) | This is only for EGUsphere preprints submitted without relation to an EGU journal.
+% European Journal of Mineralogy (ejm)
+% Fossil Record (fr)
+% Geochronology (gchron)
+% Geographica Helvetica (gh)
+% Geoscience Communication (gc)
+% Geoscientific Instrumentation, Methods and Data Systems (gi)
+% Geoscientific Model Development (gmd)
+% History of Geo- and Space Sciences (hgss)
+% Hydrology and Earth System Sciences (hess)
+% Journal of Bone and Joint Infection (jbji)
+% Journal of Micropalaeontology (jm)
+% Journal of Sensors and Sensor Systems (jsss)
+% Magnetic Resonance (mr)
+% Mechanical Sciences (ms)
+% Natural Hazards and Earth System Sciences (nhess)
+% Nonlinear Processes in Geophysics (npg)
+% Ocean Science (os)
+% Polarforschung - Journal of the German Society for Polar Research (polf)
+% Primate Biology (pb)
+% Proceedings of the International Association of Hydrological Sciences (piahs)
+% Safety of Nuclear Waste Disposal (sand)
+% Scientific Drilling (sd)
+% SOIL (soil)
+% Solid Earth (se)
+% State of the Planet (sp)
+% The Cryosphere (tc)
+% Weather and Climate Dynamics (wcd)
+% Web Ecology (we)
+% Wind Energy Science (wes)
+
+%% \usepackage commands included in the copernicus.cls:
+%\usepackage[german, english]{babel}
+%\usepackage{tabularx}
+%\usepackage{cancel}
+%\usepackage{multirow}
+%\usepackage{supertabular}
+%\usepackage{algorithmic}
+%\usepackage{algorithm}
+%\usepackage{amsthm}
+%\usepackage{float}
+%\usepackage{subfig}
+%\usepackage{rotating}
+
 \begin{document}
 
-\author[1,*]{Jemma Shipton}
-\author[2]{Thomas M. Bendall}
-\author[3]{So many others}
+\title{Gusto: a toolkit for compatible finite element dynamical cores}
+
+% \Author[affil]{given_name}{surname}
+
+\Author[1,*]{Jemma}{Shipton}
+\Author[2]{Thomas M.}{Bendall}
+\Author[3]{So many}{Others}
 
@@
 
-\title{Gusto: a toolkit for compatible finite element dynamical cores}
-\date{}
+%% The [] brackets identify the author with the corresponding affiliation. 1, 2, 3, etc. should be inserted.
+
+%% If an author is deceased, please mark the respective author name(s) with a dagger, e.g. "\Author[2,$\dag$]{Anton}{Smith}", and add a further "\affil[$\dag$]{deceased, 1 July 2019}".
+
+%% If authors contributed equally, please mark the respective author names with an asterisk, e.g.
+"\Author[2,*]{Anton}{Smith}" and "\Author[3,*]{Bradley}{Miller}" and add a further affiliation: "\affil[*]{These authors contributed equally to this work.}".
+
+\correspondence{NAME (EMAIL)}
+
+\runningtitle{TEXT}
+
+\runningauthor{TEXT}
+
+\received{}
+\pubdiscuss{} %% only important for two-stage journals
+\revised{}
+\accepted{}
+\published{}
+
+%% These dates will be inserted by Copernicus Publications during the typesetting process.
+
+\firstpage{1}
 
@@
 
-\section{Introduction}
+\begin{abstract}
+TEXT
+\end{abstract}
+
+\introduction   %% \introduction[modified heading if necessary]
 
@@
 
-\section{Code availability}
+\conclusions  %% \conclusions[modified heading if necessary]
+TEXT
 
-\bibliography{references}
+%% The following commands are for the statements about the availability of data sets and/or software code corresponding to the manuscript.
+%% It is strongly recommended to make use of these sections in case data sets and/or software code have been part of your research the article is based on.
+
+\codeavailability{TEXT} %% use this section when having only software code available
+
+\dataavailability{TEXT} %% use this section when having only data sets available
+
+\codedataavailability{TEXT} %% use this section when having data sets and software code available
+
+\sampleavailability{TEXT} %% use this section when having geoscientific samples available
+
+\videosupplement{TEXT} %% use this section when having video supplements available
+
+\appendix
+\section{}    %% Appendix A
+
+\subsection{}     %% Appendix A1, A2, etc.
+
+\noappendix       %% use this to mark the end of the appendix section. Otherwise the figures might be numbered incorrectly (e.g. 10 instead of 1).
+
+%% Regarding figures and tables in appendices, the following two options are possible depending on your general handling of figures and tables in the manuscript environment:
+
+%% Option 1: If you sorted all figures and tables into the sections of the text, please also sort the appendix figures and appendix tables into the respective appendix sections.
+%% They will be correctly named automatically.
+
+%% Option 2: If you put all figures after the reference list, please insert appendix tables and figures after the normal tables and figures.
+%% To rename them correctly to A1, A2, etc., please add the following commands in front of them:
+
+\appendixfigures  %% needs to be added in front of appendix figures
+
+\appendixtables   %% needs to be added in front of appendix tables
+
+%% Please add \clearpage between each table and/or figure. Further guidelines on figures and tables can be found below.
+
+\authorcontribution{TEXT} %% this section is mandatory
+
+\competinginterests{TEXT} %% this section is mandatory even if you declare that no competing interests are present
+
+\disclaimer{TEXT} %% optional section
+
+\begin{acknowledgements}
+TEXT
+\end{acknowledgements}
+
+%% REFERENCES
+
+%% The reference list is compiled as follows:
+
+%% \begin{thebibliography}{}
+
+%% \bibitem[AUTHOR(YEAR)]{LABEL1}
+%% REFERENCE 1
+
+%% \bibitem[AUTHOR(YEAR)]{LABEL2}
+%% REFERENCE 2
+
+%% \end{thebibliography}
+
+%% Since the Copernicus LaTeX package includes the BibTeX style file copernicus.bst,
+%% authors experienced with BibTeX only have to include the following two lines:
+%%
+\bibliographystyle{copernicus}
+\bibliography{references.bib}
+%%
+%% URLs and DOIs can be entered in your BibTeX file as:
+%%
+%% URL = {http://www.xyz.org/~jones/idx_g.htm}
+%% DOI = {10.5194/xyz}
+
+%% LITERATURE CITATIONS
+%%
+%% command                        & example result
+%% \citet{jones90}|               & Jones et al. (1990)
+%% \citep{jones90}|               & (Jones et al., 1990)
+%% \citep{jones90,jones93}|       & (Jones et al., 1990, 1993)
+%% \citep[p.~32]{jones90}|        & (Jones et al., 1990, p.~32)
+%% \citep[e.g.,][]{jones90}|      & (e.g., Jones et al., 1990)
+%% \citep[e.g.,][p.~32]{jones90}| & (e.g., Jones et al., 1990, p.~32)
+%% \citeauthor{jones90}|          & Jones et al.
+%% \citeyear{jones90}|            & 1990
+
+%% FIGURES
+
+%% When figures and tables are placed at the end of the MS (article in one-column style), please add \clearpage
+%% between bibliography and first table and/or figure as well as between each table and/or figure.
+
+% The figure files should be labelled correctly with Arabic numerals (e.g. fig01.jpg, fig02.png).
+
+%% ONE-COLUMN FIGURES
+
+%%f
+%\begin{figure}[t]
+%\includegraphics[width=8.3cm]{FILE NAME}
+%\caption{TEXT}
+%\end{figure}
+%
+%%% TWO-COLUMN FIGURES
+%
+%%f
+%\begin{figure*}[t]
+%\includegraphics[width=12cm]{FILE NAME}
+%\caption{TEXT}
+%\end{figure*}
+%
+%
+%%% TABLES
+%%%
+%%% The different columns must be seperated with a & command and should
+%%% end with \\ to identify the column brake.
+%
+%%% ONE-COLUMN TABLE
+%
+%%t
+%\begin{table}[t]
+%\caption{TEXT}
+%\begin{tabular}{column = lcr}
+%\tophline
+%
+%\middlehline
+%
+%\bottomhline
+%\end{tabular}
+%\belowtable{} % Table Footnotes
+%\end{table}
+%
+%%% TWO-COLUMN TABLE
+%
+%%t
+%\begin{table*}[t]
+%\caption{TEXT}
+%\begin{tabular}{column = lcr}
+%\tophline
+%
+%\middlehline
+%
+%\bottomhline
+%\end{tabular}
+%\belowtable{} % Table Footnotes
+%\end{table*}
+%
+%%% LANDSCAPE TABLE
+%
+%%t
+%\begin{sidewaystable*}[t]
+%\caption{TEXT}
+%\begin{tabular}{column = lcr}
+%\tophline
+%
+%\middlehline
+%
+%\bottomhline
+%\end{tabular}
+%\belowtable{} % Table Footnotes
+%\end{sidewaystable*}
+%
+%
+%%% MATHEMATICAL EXPRESSIONS
+%
+%%% All papers typeset by Copernicus Publications follow the math typesetting regulations
+%%% given by the IUPAC Green Book (IUPAC: Quantities, Units and Symbols in Physical Chemistry,
+%%% 2nd Edn., Blackwell Science, available at: http://old.iupac.org/publications/books/gbook/green_book_2ed.pdf, 1993).
+%%%
+%%% Physical quantities/variables are typeset in italic font (t for time, T for Temperature)
+%%% Indices which are not defined are typeset in italic font (x, y, z, a, b, c)
+%%% Items/objects which are defined are typeset in roman font (Car A, Car B)
+%%% Descriptions/specifications which are defined by itself are typeset in roman font (abs, rel, ref, tot, net, ice)
+%%% Abbreviations from 2 letters are typeset in roman font (RH, LAI)
+%%% Vectors are identified in bold italic font using \vec{x}
+%%% Matrices are identified in bold roman font
+%%% Multiplication signs are typeset using the LaTeX commands \times (for vector products, grids, and exponential notations) or \cdot
+%%% The character * should not be applied as mutliplication sign
+%
+%
+%%% EQUATIONS
+%
+%%% Single-row equation
+%
+%\begin{equation}
+%
+%\end{equation}
+%
+%%% Multiline equation
+%
+%\begin{align}
+%& 3 + 5 = 8\\
+%& 3 + 5 = 8\\
+%& 3 + 5 = 8
+%\end{align}
+%
+%
+%%% MATRICES
+%
+%\begin{matrix}
+%x & y & z\\
+%x & y & z\\
+%x & y & z\\
+%\end{matrix}
+%
+%
+%%% ALGORITHM
+%
+%\begin{algorithm}
+%\caption{...}
+%\label{a1}
+%\begin{algorithmic}
+%...
+%\end{algorithmic}
+%\end{algorithm}
+%
+%
+%%% CHEMICAL FORMULAS AND REACTIONS
+%
+%%% For formulas embedded in the text, please use \chem{}
+%
+%%% The reaction environment creates labels including the letter R, i.e. (R1), (R2), etc.
+%
+%\begin{reaction}
+%%% \rightarrow should be used for normal (one-way) chemical reactions
+%%% \rightleftharpoons should be used for equilibria
+%%% \leftrightarrow should be used for resonance structures
+%\end{reaction}
+%
+%
+%%% PHYSICAL UNITS
+%%%
+%%% Please use \unit{} and apply the exponential notation
 

### 2024-02-20 source/paper.tex
a bit more text on the software and FML

@@
 \newcommand{\pp}[2]{\frac{\partial #1}{\partial #2}} 
+\newcommand{\JScomment}[1]{\textit{\textbf{Jemma says: #1}}}
 
@@
 
+\JScomment{Text here depends on what the schematic looks like - this
+  paragraph should refer to the schematic and explain all aspects}
 Prognostic equation sets are each defined in a python class, making
 use of the common finite element forms for the transport and
-difffusion terms.
-
-** mention physics source terms?
+difffusion terms. The timeloop class contains the definition of the
+timestepping procedure and the \texttt{run} method that loops over the
+timesteps. It also sets up diagnostic fields and calls the IO as
+required. The simplest timeloop class will just apply one time
+discretisation method to the equation set but it is common in weather
+and climate prediction models to split the different types of terms
+and apply a time discretisation separately to each type of term. For
+example, the transport and forcing terms might be solved for
+separately, such as in **, which enables explicit transport schemes to
+be applied to each field. ** something about SIQN structure. Terms
+related to the physics are solved for in the appropriate part of the
+iteration loop, with physics processes that occur on `fast' timescales
+evaluated more frequently and `slower' physics processes evaluated
+just once at the end of the timestep. The choice of where to evaluate
+the physics processes in the iteration is also based on the
+computational cost and the bast way to do this is still an open
+research question (**ref).
 
@@
 \label{sec: FML}
+The discussion of the time discretisation so far has been general i.e.
+not specific to any particular equation set. This is also true of the
+implementation of the timestepping schemes. This enables flexibility
+and reproducibility (**elaborate). This functionality is enabled
+through form manipulation labelling (FML) which is used to associate a
+label to each term in the residual form defined in the equation
+class. The timestepping methods are then able to reason about which
+manipulations to apply to which forms by applying filters. A simple
+example would be a splitting method that evaluates transport terms
+separately. The full residual is passed to the transport scheme which
+then applies a filter to drop all terms that are not labelled as
+transport.
 
@@
 demonstrate the tools available in Gusto, such as the choice of grids,
-order of finite element spaces, different spatial discretisation
-options, timestepping schemes and physics. The numerical properties of
-these configurations (convergence, stability, accuracy and efficiency)
-have either been demonstrated elsewhere, or will be investigated more
-thoroughly in future publications.
+HDiv element families, order of finite element spaces, different
+spatial discretisation options, timestepping schemes and physics. The
+numerical properties of these configurations (convergence, stability,
+accuracy and efficiency) have either been demonstrated elsewhere, or
+will be investigated more thoroughly in future publications.
 

### 2024-03-07 source/paper.tex
some progress

@@
 \affil[2]{Met Office}
-\affil[*]{Correspondence to: \texttt{j.shipton@exeter.ac.uk}}
 
@@
 finite element function spaces for each field gives what is known as a
-mixed finite element discretisation. If the finite element function
-spaces are chosen so that
+mixed finite element discretisation. If these different finite element
+function spaces are chosen so that discrete versions of the vector
+calculus identities ** hold then the discretisation is called
+`mimetic' or `compatible'.
 
@@
 velocity always contains a nonlinear transport term, a term due to
-rotation and a pressure gradient term. The other prognostic fields
-satisfy either an advective or continuity form of the transport
-equation. These similarities remain when we formulate the compatible
-finite element discretisations, as we shall see below.
+rotation and a `pressure' gradient term (in shallow water this is the
+gradient of the depth but the term plays the same role in the dynamics
+and is derived from the pressure gradient term in the full equations
+(**cite Zeitlin book)). The other prognostic fields satisfy either an
+advective or continuity form of the transport equation. These
+similarities remain when we formulate the compatible finite element
+discretisations, as we shall see below, and importantly this means
+that the discretisation of these terms can be written in a general way
+that means it can be applied to any of the equation sets.
 
@@
 
+In line with the theory outlined in section \ref{}, we always choose
+the velocity $\MM{u} \in \mathbb{V}_u$ with $\mathbb{V}_u \subset
+H(\text{div})$. The field whose gradient appears in the `pressure'
+gradient term is chosen to be in the space $\mathbb{V}_p$ with
+$\mathbb{V}_p \subset L_2$. The compressible Boussinesq and
+compressible Euler equations have a third prognostic variable, either
+buoyancy or potential temperature, which is chosen to be in the space
+$\mathbb{V}_\theta$ with $\mathbb{V}_\theta \subset ??$ These choices
+are governed by the properties required of discretisation of the
+linearised equation sets (as discussed in section ** and other
+refs). To solve the nonlinear equation sets we need to introduce
+stable and accurate discretisations of the transport terms and the
+nonlinear pressure gradient term $c_p\theta\nabla \Pi$ in the
+compressible Euler velocity equation. Since the prognostic variables
+are discretised using different finite element function spaces
+$\mathbb{V}_u$, $\mathbb{V}_p$ and $\mathbb{V}_\theta$, each with
+different continuity properties between the elements, the choice of
+stable transport discretisation is different for each prognostic
+variable.
+
 ** Don't do whole equation set - go term by term and try to write as
 few equations as possible.
-
-In line with the theory outlined in section \ref{}, we
-choose $\MM{u} \in \mathbb{V}_1$ and $D \in \mathbb{V}_2$ with
-$\mathbb{V}_1 \subset H(\text{div})$ and $\mathbb{V}_2 \subset L^2$. The
-corresponding test functions are denoted by $\MM{w}$ and $\phi$. This gives
 
@@
 the physics processes in the iteration is also based on the
-computational cost and the bast way to do this is still an open
+computational cost and the best way to do this is still an open
 research question (**ref).
@@
 manipulations to apply to which forms by applying filters. A simple
-example would be a splitting method that evaluates transport terms
-separately. The full residual is passed to the transport scheme which
-then applies a filter to drop all terms that are not labelled as
-transport.
+example would be a splitting method that evaluates transport terms and
+diffusion terms separately. The full residual is passed to the
+transport scheme which then applies a filter to drop all terms that
+are not labelled as transport; similarly for diffusion. We illustrate
+this procedure in figure \ref{fig:advdiff} where we show the
+definition of the residual for the advection diffusion equation for a
+scalar variable $q$ with a prescribed transport velocity $\MM{c}$ and
+diffusivity $\kappa$
+
+\begin{equation}
+  q_t + \MM{c}\cdot\nabla q - \kappa\nabla^2 q = 0,
+\end{equation}
+with the terms labelled either \texttt{time\_derivative},
+\texttt{transport} or \texttt{diffusion}. The timestepping loop then
+applies a step of forward Euler to the transport term and backwards
+Euler to the diffusive term, i.e. the time discretisation is
+
+\begin{align}
+  q^* &= q^n - \Delta t (\MM{c}\cdot\nabla) q^n, \\
+  q^{n+1} &= q^* + \Delta t \kappa \nabla^2 q^{n+1}.
+\end{align}
 
@@
 will be investigated more thoroughly in future publications.
+
+To include:
+\begin{itemize}
+\item terminator toy (Tim)
+\item shallow water on different grids and with different HDiv families (Nell and Dan)
+\item convergence with different RK schemes / multistage schemes (Alex)
+\item vertical slice Boussinesq / Euler (+ moisture?)
+\item moist baroclinic wave in channel?? (Alex)
+\item spherical gravity wave, Boussinesq (Jemma)
+\item spherical baroclinic wave, Euler (Dan)
+\end{itemize}
 

### 2024-03-07 source/paper.tex
better start for this paragraph describing the software structure

@@
 
+\begin{figure}
+  \centering{\includegraphics[width=0.8\textwidth]{gusto_schematic.png}}
+\end{figure}
+
 \JScomment{Text here depends on what the schematic looks like - this
-  paragraph should refer to the schematic and explain all aspects}
-Prognostic equation sets are each defined in a python class, making
-use of the common finite element forms for the transport and
-difffusion terms. The timeloop class contains the definition of the
-timestepping procedure and the \texttt{run} method that loops over the
-timesteps. It also sets up diagnostic fields and calls the IO as
+  paragraph should refer to the schematic and explain all aspects} The
+Gusto model comprises a domain and a prognostic equation set. The
+domain holds the mesh and the finite element function spaces along
+with any other information about the geometry such as the rotation
+vector. The Prognostic equation sets are each defined in a python
+class, making use of the common finite element forms for the transport
+and difffusion terms. The timeloop class contains the definition of
+the timestepping procedure and the \texttt{run} method that loops over
+the timesteps. It also sets up diagnostic fields and calls the IO as
 required. The simplest timeloop class will just apply one time
@@
 
-\section{Form Manipulation Labelling}
+\section{Form Manipulation Language}
 \label{sec: FML}
@@
 and reproducibility (**elaborate). This functionality is enabled
-through form manipulation labelling (FML) which is used to associate a
-label to each term in the residual form defined in the equation
-class. The timestepping methods are then able to reason about which
-manipulations to apply to which forms by applying filters. A simple
-example would be a splitting method that evaluates transport terms and
-diffusion terms separately. The full residual is passed to the
-transport scheme which then applies a filter to drop all terms that
-are not labelled as transport; similarly for diffusion. We illustrate
-this procedure in figure \ref{fig:advdiff} where we show the
-definition of the residual for the advection diffusion equation for a
-scalar variable $q$ with a prescribed transport velocity $\MM{c}$ and
-diffusivity $\kappa$
+through the form manipulation language (FML) which is used to
+associate a label to each term in the residual form defined in the
+equation class. The timestepping methods are then able to reason about
+which manipulations to apply to which forms by applying filters. A
+simple example would be a splitting method that evaluates transport
+terms and diffusion terms separately. The full residual is passed to
+the transport scheme which then applies a filter to drop all terms
+that are not labelled as transport; similarly for diffusion. We
+illustrate this procedure in figure \ref{fig:advdiff} where we show
+the definition of the residual for the advection diffusion equation
+for a scalar variable $q$ with a prescribed transport velocity
+$\MM{c}$ and diffusivity $\kappa$
 

### 2024-03-08 source/paper.tex
some edits and most of the pressure gradient section

@@
 \def\MM#1{\boldsymbol{#1}}
+\DeclareMathOperator{\diff}{d}
+\newcommand{\avg}[1]{\{#1\}}
+\newcommand{\jump}[1]{[\![#1]\!]}
 \newcommand{\pp}[2]{\frac{\partial #1}{\partial #2}} 
@@
 \end{abstract}
+
+\begin{itemize}
+\item Section 1: why compatible FEM
+\item Section 2: why Gusto
+\item Sections 3 and 4: this is Gusto
+\item Section 5: and here are some examples of what we can currently do
+\end{itemize}
 
@@
 The dynamical core at the heart of any weather or climate model solves
-the rotating compressible Euler equations that model atmospheric flow,
-often with additional approximations such as shallow vs deep
-atmosphere or hydrostatic vs non-hydrostatic flow (** mention this
-here or at the end of the section??). We write these equations in
-terms of the prognostic variables $\MM{u}$, $\rho$ and $\theta$ where
-$\MM{u}=(u, v, w)$ is now a three dimensional velocity field, $\rho$
-is the fluid density and $\theta$ is the potential temperature. In
-terms of these variables, the equations are:
+the rotating compressible Euler equations that model atmospheric
+flow. We write these equations in terms of the prognostic variables
+$\MM{u}$, $\rho$ and $\theta$ where $\MM{u}=(u, v, w)$ is now a three
+dimensional velocity field, $\rho$ is the fluid density and $\theta$
+is the potential temperature. The pressure gradient term is written in
+terms of the Exner pressure $\Pi$ which is calculated through an
+equation of state. In terms of these variables, the equations are:
 
@@
 
-\subsection{Finite Element Discretisation}
+\section{Finite Element Discretisation}
 \label{sec: FEM}
@@
 buoyancy or potential temperature, which is chosen to be in the space
-$\mathbb{V}_\theta$ with $\mathbb{V}_\theta \subset ??$ These choices
-are governed by the properties required of discretisation of the
-linearised equation sets (as discussed in section ** and other
-refs). To solve the nonlinear equation sets we need to introduce
-stable and accurate discretisations of the transport terms and the
-nonlinear pressure gradient term $c_p\theta\nabla \Pi$ in the
-compressible Euler velocity equation. Since the prognostic variables
-are discretised using different finite element function spaces
-$\mathbb{V}_u$, $\mathbb{V}_p$ and $\mathbb{V}_\theta$, each with
-different continuity properties between the elements, the choice of
-stable transport discretisation is different for each prognostic
-variable.
+$\mathbb{V}_\theta$ which has degrees of freedom that are collocated
+with those of the vertical velocity, mimicking the Charney-Phillips
+staggering as discussed in **. These choices are governed by the
+properties required of discretisation of the linearised equation sets
+(as discussed in section ** and other refs). To solve the nonlinear
+equation sets we need to introduce stable and accurate discretisations
+of the transport terms and the nonlinear pressure gradient term in the
+compressible Euler equation set. In addition, some test cases require
+the addition of viscous and diffusive terms to generate converged
+solutions \ref{}. In the following subsections we illustrate a
+non-exhaustive (**probably!) selection of the options available in
+Gusto.
+
+\subsection{Transport terms}
+Since the prognostic variables are discretised using different finite
+element function spaces $\mathbb{V}_u$, $\mathbb{V}_p$ and
+$\mathbb{V}_\theta$, each with different continuity properties between
+the elements, the choice of stable transport discretisation is
+different for each prognostic variable.
+
+\subsection{Pressure gradient term}
+The pressure gradient term is discretised in a similar way for each
+equation set, with some additional considerations required for the
+nonlinear form of this term in the compressible Euler equations. Since
+the pressure (or depth, for shallow water) field is discretised in
+$\mathbb{V}_p \subset L_2$, the gradient is not defined globally and
+the term must be integrated by parts. In the linear case, using the
+notation for the compressible Boussinesq equations, the weak form is
+
+\begin{equation}
+\int_\Omega\MM{w}\cdot\nabla p \diff x = -\int_\Omega\nabla\cdot\MM{w}p\diff x,
+\end{equation}
+where $\Omega$ denotes the domain and there are no surface terms from
+the integration by parts because these vanish on interior element
+boundaries since the normal component of $\MM{w}$ is continuous across
+element faces and on exterior boundaries we set no normal flow
+boundary conditions. However, the pressure gradient term in the
+compressible Euler equations, $c_p\theta\nabla\Pi$, is nonlinear and
+this introduces an additional complication. Integrating by parts is
+still required since $\Pi$ is evaluated as a function of the density
+$\rho \in L_2$ and so is discontinuous, but this means that we now
+have a differential operator applied to $\theta \in \mathbb{V}_\theta$
+which is discontinuous in the vertical. Following \ref{} we integrate
+by parts in each element $e$, giving
+
+\begin{equation}
+  c_p\int_e\MM{w}\cdot\theta\nabla\Pi\diff x = -c_p\int_e\nabla\cdot(\MM{w}\theta)\Pi\diff x + c_p\int_{\partial e}\theta^-\MM{w}\cdot\MM{n}\Pi^- \diff S,
+\end{equation}
+where $\partial e$ denotes the element faces, $\MM{n}$ is the outward
+pointing normal to the face and $\cdot^-$ denotes the value of a
+variable on the interior side of the boundary. Summing over the domain
+gives
+
+\begin{equation}
+  c_p\int_\Omega\MM{w}\cdot\theta\nabla\Pi\diff x = -c_p\int_\Omega\nabla\cdot(\MM{w}\theta)\Pi\diff x + c_p\int_{\Gamma_v}\jump{\theta\MM{w}\cdot\MM{n}}\avg{\Pi} \diff S,
+\end{equation}
+where $\jump{\cdot}$ and $\avg{\cdot}$.
+
+\subsection{Viscous and diffusive terms}
 
 ** Don't do whole equation set - go term by term and try to write as
-few equations as possible.
+few equations as possible: transport, pressure gradient, diffusion
 
@@
 domain holds the mesh and the finite element function spaces along
-with any other information about the geometry such as the rotation
-vector. The Prognostic equation sets are each defined in a python
-class, making use of the common finite element forms for the transport
-and difffusion terms. The timeloop class contains the definition of
-the timestepping procedure and the \texttt{run} method that loops over
-the timesteps. It also sets up diagnostic fields and calls the IO as
-required. The simplest timeloop class will just apply one time
-discretisation method to the equation set but it is common in weather
-and climate prediction models to split the different types of terms
-and apply a time discretisation separately to each type of term. For
-example, the transport and forcing terms might be solved for
-separately, such as in **, which enables explicit transport schemes to
-be applied to each field. ** something about SIQN structure. Terms
-related to the physics are solved for in the appropriate part of the
-iteration loop, with physics processes that occur on `fast' timescales
-evaluated more frequently and `slower' physics processes evaluated
-just once at the end of the timestep. The choice of where to evaluate
-the physics processes in the iteration is also based on the
-computational cost and the best way to do this is still an open
-research question (**ref).
+with any other information about the space-time domain such as the
+coordinates and the timestep. The Prognostic equation sets are each
+defined in a python class, making use of the common finite element
+forms for the transport and difffusion terms. The timeloop class
+contains the definition of the timestepping procedure and the
+\texttt{run} method that loops over the timesteps. It also sets up
+diagnostic fields and calls the IO as required. The simplest timeloop
+class will just apply one time discretisation method to the equation
+set but it is common in weather and climate prediction models to split
+the different types of terms and apply a time discretisation
+separately to each type of term. For example, the transport and
+forcing terms might be solved for separately, such as in **, which
+enables explicit transport schemes to be applied to each field. **
+something about SIQN structure. Terms related to the physics are
+solved for in the appropriate part of the iteration loop, with physics
+processes that occur on `fast' timescales evaluated more frequently
+and `slower' physics processes evaluated just once at the end of the
+timestep. The choice of where to evaluate the physics processes in the
+iteration is also based on the computational cost and the best way to
+do this is still an open research question (**ref).
 
@@
 \item terminator toy (Tim)
-\item shallow water on different grids and with different HDiv families (Nell and Dan)
-\item convergence with different RK schemes / multistage schemes (Alex)
-\item vertical slice Boussinesq / Euler (+ moisture?)
-\item moist baroclinic wave in channel?? (Alex)
+\item shallow water on different grids and with different HDiv
+  families (Nell) e.g. Williamson 5 cubed sphere PV and Galewsky
+  icosahedral sphere PV, high res (ref 6)
+\item convergence with different RK schemes / multistage schemes using
+  Williamson 2 or 6, L2 and Linf errors for D and u (2 plots, one L2,
+  one Linf) (Alex)
+\item vertical slice Euler - Straka, different orders, illustrates
+  recovery for diffusion (Tom)
+\item 3D moist bubble with rain (Tom/Alex)
 \item spherical gravity wave, Boussinesq (Jemma)
-\item spherical baroclinic wave, Euler (Dan)
+\item spherical baroclinic wave, Euler - whatever people usually plot
+  (horizontal slice pressure contours, temperature shadng) (Dan)
+\item Held Suarez??
 \end{itemize}

### 2024-03-10 source/paper.tex
some edits

@@
 %\usepackage{supertabular}
-%\usepackage{algorithmic}
-%\usepackage{algorithm}
+\usepackage{algorithmic}
+\usepackage{algorithm}
 %\usepackage{amsthm}
@@
 code structure of Gusto followed by a detailed example (section
-\ref{sec: FML}) of the Form Manipulation Language (FML) that facilitates
-symbolic manipulation of the finite element forms within our
-timestepping code. Section \ref{sec: results} contains a selection of
-numerical results that showcases the range of discretisations
-available within Gusto. We summarise our current status and future
+\ref{sec: FML}) of the Form Manipulation Language (FML) that
+facilitates symbolic manipulation of the finite element forms within
+our timestepping code. Section \ref{sec: results} contains a selection
+of numerical results that showcases the range of discretisations
+available within Gusto. **mention here new functionality facilitated
+by the flexibility of Gusto? We summarise our current status and future
 directions in section \ref{sec: summary}.
@@
 guide the design of the code structure described in the next section
-and motivate the use of Form Manipulation Labelling (FML) that
+and motivate the use of Form Manipulation Language (FML) that
 underlies the flexibility of Gusto, a detailed example of which will
@@
 and is derived from the pressure gradient term in the full equations
-(**cite Zeitlin book)). The other prognostic fields satisfy either an
-advective or continuity form of the transport equation. These
-similarities remain when we formulate the compatible finite element
-discretisations, as we shall see below, and importantly this means
-that the discretisation of these terms can be written in a general way
-that means it can be applied to any of the equation sets.
+\citep{zeitlin2018geophysical}). The other prognostic fields satisfy
+either an advective or continuity form of the transport
+equation. These similarities remain when we formulate the compatible
+finite element discretisations, as we shall see below, and importantly
+this means that the discretisation of these terms can be written in a
+general way that means it can be applied to any of the equation sets.
 
@@
 boundaries since the normal component of $\MM{w}$ is continuous across
-element faces and on exterior boundaries we set no normal flow
+element facets and on exterior boundaries we set no normal flow
 boundary conditions. However, the pressure gradient term in the
@@
 \begin{equation}
-  c_p\int_e\MM{w}\cdot\theta\nabla\Pi\diff x = -c_p\int_e\nabla\cdot(\MM{w}\theta)\Pi\diff x + c_p\int_{\partial e}\theta^-\MM{w}\cdot\MM{n}\Pi^- \diff S,
+  c_p\int_e\MM{w}\cdot\theta\nabla\Pi\diff x = -c_p\int_e\nabla\cdot(\MM{w}\theta)\Pi\diff x + c_p\int_{\partial e}\hat{\theta}\MM{w}\cdot\MM{n}\hat{\Pi}\diff S,
 \end{equation}
-where $\partial e$ denotes the element faces, $\MM{n}$ is the outward
-pointing normal to the face and $\cdot^-$ denotes the value of a
-variable on the interior side of the boundary. Summing over the domain
-gives
+where $\partial e$ denotes the element facets, $\MM{n}$ is the outward
+pointing normal to the facet and $\hat{q}$ denotes the value of a
+variable $q$ on the facet, which for discontinuous variables does not
+have a unique definition. Summing over the domain gives
 
@@
 \end{equation}
-where $\jump{\cdot}$ and $\avg{\cdot}$.
+where $\Gamma_v$ denotes the vertical element facets, across which
+$\theta$ is discontinuous. Here we have chosen, as in \ref{}, to take
+the average value of $\Pi$ across the facets, $\avg{\Pi} =
+(\Pi^++\Pi^-)/2$, where the $+$ and $-$ indicate each side fo the
+facet. The brackets $\jump{q} = q^+-q^-$ denote the jump in a value
+across the facet and the integral over the horizontal facets vanishes
+since both $\theta$ and $\MM{w}\cdot\MM{n}$ are continuous across
+these facets so the jump is zero.
 
@@
 
+\begin{algorithm}
+  \caption{Pseudocode for SIQN timestep loop}\label{alg:SIQN}
+  \begin{algorithmic}
+    \STATE
+  \end{algorithmic}
+\end{algorithm}
+
 \begin{itemize}
-\item include schematic
-\item generic example - no FML
+\item generic example(s) - no FML, just pseudocode for basic
+  timestepper, SIQN and split physics
 \item lead in to next section... facilitated by FML i.e. timesteppers don't care what equations they're given because forms are labelled
@@
 
-\subsection{Shallow water simulations}
+\subsection{Terminator `toy' chemistry}
+This test combines transport and a `toy' chemistry model to
+demonstrate physics-dynamics coupling in Gusto (where here chemistry
+is the `physics').
+
+\subsection{Shallow water}
 The standard suite of shallow water test cases originally proposed in
@@
 the unstable barotropic jet described in \citet{galewsky2004initial}
-(**and linear tests?)
 

### 2024-07-24 source/paper.tex
some progress on FML section

@@
 The aim of this section is to introduce three governing equation sets
-commonly used in the development of weather and climate models, along
-with enough detail about the finite element discretiations provided by
-Gusto to demonstrate the similarities between them. These similarities
-guide the design of the code structure described in the next section
-and motivate the use of Form Manipulation Language (FML) that
-underlies the flexibility of Gusto, a detailed example of which will
-be given in section \ref{sec: FML}. The convention in Gusto is to
-write the equations in residual form, meaning that all the terms
-appear on the left hand side. We will follow this convention below in
-contrast to many papers which present the equations with the pressure
-gradient term and any forcing terms on the right hand side.
+commonly used in the development of discretisation methods for weather
+and climate models, along with enough detail about the finite element
+discretiations provided by Gusto to demonstrate the similarities
+between them. These similarities guide the design of the code
+structure described in the next section and motivate the use of Form
+Manipulation Language (FML) that underlies the flexibility of Gusto, a
+detailed example of which will be given in section \ref{sec: FML}. The
+convention in Gusto is to write the equations in residual form,
+meaning that all the terms appear on the left hand side. We will
+follow this convention below in contrast to many papers which present
+the equations with the pressure gradient term and any forcing terms on
+the right hand side.
 
@@
 $\MM{u} = (u, v)$ and the fluid depth $D = H + h - b$ where $H$ is the
-undisturbed depth of the fluid layer, $h$ is the perturbation to the
-fluid depth and $b$ is the bottom topography, if present. The
+undisturbed depth of the fluid layer, $h$ is the free surface
+deviation from $H$, and $b$ is the bottom topography, if present. The
 equations are:
@@
 \begin{align}
-  \pp{\MM{u}}{t} + \MM{u}\cdot\nabla\MM{u} + f\hat{\MM{k}}\times\MM{u} + g\nabla (D+b) &= 0, \\
+  \pp{\MM{u}}{t} + (\MM{u}\cdot\nabla)\MM{u} + f\hat{\MM{k}}\times\MM{u} + g\nabla (D+b) &= 0, \\
   \pp{D}{t} + \nabla\cdot(D\MM{u}) &= 0,
@@
 where $f$ is the Coriolis parameter, $\hat{\MM{k}}$ is the unit vector
-pointing upwards and $g$ is the acceleration due to gravity. The
-nonlinear velocity advection term can be replaced by the sum of a
-vorticity-based term and the gradient of the kinetic energy. This is
-known as the vector invariant form and it enables the construction of
-schemes that conservation energy and/or enstrophy
-\citep{mcrae2014energy, bauer2018energy, wimmer2020energy,
-  wimmer2021energy}. However, this form can exhibit numerical
-instabilities \citep{bell2017numerical} and it is not yet clear which
-form is preferable so in Gusto we offer both forms of the equations.
+pointing upwards and $g$ is the acceleration due to gravity.
 
@@
 we shall see in the compressible Euler equations. We write these
-equations in terms of the prognostic variables $\MM{u}$, $b$ and
-$p$ where $\MM{u}=(u, v, w)$ is now a three dimensional velocity
-field, $b$ is the fluid buoyancy and $p$ is pressure. In terms of
-these variables, the equations are:
+equations in terms of the prognostic variables $\MM{u}$, $b$ and $p$
+where $\MM{u}$ is a three dimensional velocity field, $b$ is the fluid
+buoyancy and $p$ is pressure. In terms of these variables, the
+equations are:
 
@@
   \pp{p}{t} + c^2\nabla\cdot\MM{u} & = 0, \\
-  \pp{b}{t} + \MM{u}\cdot\nabla b & = 0,
+  \pp{b}{t} + (\MM{u}\cdot\nabla) b & = 0,
 \end{align}
@@
 $\hat{\MM{k}}$ is the unit vector pointing upwards and $c$ is the
-acoustic wave speed.
+acoustic wave speed. The velocity $\MM{u}$ can be two or three
+dimensional, depending on whether the domain is a vertical slice
+described by $(x, z)$ coordinates, of a three dimensional domain
+described by $(x, y, z)$ coordinates.
 
@@
 flow. We write these equations in terms of the prognostic variables
-$\MM{u}$, $\rho$ and $\theta$ where $\MM{u}=(u, v, w)$ is now a three
-dimensional velocity field, $\rho$ is the fluid density and $\theta$
-is the potential temperature. The pressure gradient term is written in
-terms of the Exner pressure $\Pi$ which is calculated through an
-equation of state. In terms of these variables, the equations are:
+$\MM{u}$, $\rho$ and $\theta$ where $\MM{u}$ is the velocity field,
+$\rho$ is the fluid density and $\theta$ is the potential
+temperature. The pressure gradient term is written in terms of the
+Exner pressure $\Pi$ which is calculated through an equation of
+state. In terms of these variables, the equations are:
 
@@
   \pp{\rho}{t} + \nabla\cdot(\MM{u}\rho) & = 0, \\
-  \pp{\theta}{t} + \MM{u}\cdot\nabla\theta & = 0, \\
+  \pp{\theta}{t} + (\MM{u}\cdot\nabla)\theta & = 0, \\
   \Pi^{(1-\kappa)/\kappa} & = \frac{R}{p_0}\rho\theta, & 
@@
 where $\Pi$ is the Exner pressure, $p_0$ is a reference pressure,
-typically taken to be *, $R$ is ** and $\kappa$ is **.
+typically taken to be *, $R$ is ** and $\kappa$ is **. The velocity
+$\MM{u}$ can be two or three dimensional, depending on whether the
+domain is a vertical slice described by $(x, z)$ coordinates, of a
+three dimensional domain described by $(x, y, z)$ coordinates.
+
+In all of the equation sets described here, the nonlinear velocity
+advection term $(\MM{u}\cdot\nabla)\MM{u}$ can be replaced by the sum
+of a vorticity-based term and the gradient of the kinetic energy. This
+is known as the vector invariant form and it enables the construction
+of schemes that conservation energy and/or enstrophy
+\citep{mcrae2014energy, bauer2018energy, wimmer2020energy,
+  wimmer2021energy}. However, this form can exhibit numerical
+instabilities \citep{bell2017numerical} and it is not yet clear which
+form is preferable so in Gusto we offer both forms of the equations.
 
@@
 
-\subsection{Transport terms}
-Since the prognostic variables are discretised using different finite
-element function spaces $\mathbb{V}_u$, $\mathbb{V}_p$ and
-$\mathbb{V}_\theta$, each with different continuity properties between
-the elements, the choice of stable transport discretisation is
-different for each prognostic variable.
-
 \subsection{Pressure gradient term}
@@
 the pressure (or depth, for shallow water) field is discretised in
-$\mathbb{V}_p \subset L_2$, the gradient is not defined globally and
-the term must be integrated by parts. In the linear case, using the
-notation for the compressible Boussinesq equations, the weak form is
+$\mathbb{V}_p \subset L_2$, meaning that it is discontinuous between
+cells, the gradient is not defined globally and the term must be
+integrated by parts. In the linear case, using the notation for the
+compressible Boussinesq equations, the weak form is
 
@@
 where $\Gamma_v$ denotes the vertical element facets, across which
-$\theta$ is discontinuous. Here we have chosen, as in \ref{}, to take
+$\theta$ is discontinuous. Here we have chosen, as in \citet{}, to take
 the average value of $\Pi$ across the facets, $\avg{\Pi} =
-(\Pi^++\Pi^-)/2$, where the $+$ and $-$ indicate each side fo the
+(\Pi^++\Pi^-)/2$, where the $+$ and $-$ indicate each side of the
 facet. The brackets $\jump{q} = q^+-q^-$ denote the jump in a value
@@
 
+\subsection{Transport terms}
+Since the prognostic variables are discretised using different finite
+element function spaces $\mathbb{V}_u$, $\mathbb{V}_p$ and
+$\mathbb{V}_\theta$, each with different continuity properties between
+the elements, the choice of stable transport discretisation is
+different for each prognostic variable.
+
 \subsection{Viscous and diffusive terms}
 
-** Don't do whole equation set - go term by term and try to write as
-few equations as possible: transport, pressure gradient, diffusion
+For the simulations presented here we do not require viscous or
+diffusive terms for stability, but sometimes they are required to
+produce converged solutions, ie. so that refining the grid does not
+lead to smaller and smaller scale motions. Here we present the symmetric
+interior penalty discretisation method used for these terms in Gusto.
+
+\begin{equation}
+  \nu\int_\Omega\MM{w}\cdot\nabla^2\MM{u}\diff x = \nu\int_\Omega\nabla\MM{w}:\nabla\MM{u}\diff x
+\end{equation}
 
@@
 
-\JScomment{Text here depends on what the schematic looks like - this
-  paragraph should refer to the schematic and explain all aspects} The
-Gusto model comprises a domain and a prognostic equation set. The
-domain holds the mesh and the finite element function spaces along
-with any other information about the space-time domain such as the
-coordinates and the timestep. The Prognostic equation sets are each
-defined in a python class, making use of the common finite element
-forms for the transport and difffusion terms. The timeloop class
-contains the definition of the timestepping procedure and the
-\texttt{run} method that loops over the timesteps. It also sets up
-diagnostic fields and calls the IO as required. The simplest timeloop
-class will just apply one time discretisation method to the equation
-set but it is common in weather and climate prediction models to split
-the different types of terms and apply a time discretisation
-separately to each type of term. For example, the transport and
-forcing terms might be solved for separately, such as in **, which
-enables explicit transport schemes to be applied to each field. **
-something about SIQN structure. Terms related to the physics are
-solved for in the appropriate part of the iteration loop, with physics
-processes that occur on `fast' timescales evaluated more frequently
-and `slower' physics processes evaluated just once at the end of the
-timestep. The choice of where to evaluate the physics processes in the
-iteration is also based on the computational cost and the best way to
-do this is still an open research question (**ref).
+\JScomment{This paragraph should refer to the schematic and explain
+  all aspects} The Gusto model comprises a domain and a prognostic
+equation set. The domain holds the mesh and the finite element
+function spaces along with any other information about the space-time
+domain such as the coordinates and the timestep. The prognostic
+equation sets are each defined in a python class, making use of the
+common finite element forms for the transport and difffusion terms to
+set up the residual (defined in section \ref{}) in weak form. At this
+point, no particular discretisation of the transport or diffusion
+forms is specified. This happens in the \texttt{SpatialMethod} class
+which can be used to apply the methods described in section \ref{} to
+the relevant terms. Any physics source terms, such as **, are added by
+calling the \texttt{Physics} class that defines them. The
+\texttt{TimeStepper} class contains the definition of the timestepping
+procedure and the \texttt{run} method that loops over the
+timesteps. It also sets up diagnostic fields and calls the IO as
+required. The simplest timeloop class will just apply one time
+discretisation method to the equation set but it is common in weather
+and climate prediction models to split the different types of terms
+and apply a time discretisation separately to each type of term. For
+example, the transport and forcing terms might be solved for
+separately, such as in **, which enables explicit transport schemes to
+be applied to each field. ** something about SIQN structure. Terms
+related to the physics are solved for in the appropriate part of the
+iteration loop, with physics processes that occur on `fast' timescales
+evaluated more frequently and `slower' physics processes evaluated
+just once at the end of the timestep. The choice of where to evaluate
+the physics processes in the iteration is also based on the
+computational cost and the best way to do this is still an open
+research question (**ref).
+
+\begin{itemize}
+\item what about initialisation and diagnostics?
+\end{itemize}
 
@@
 equation class. The timestepping methods are then able to reason about
-which manipulations to apply to which forms by applying filters. A
-simple example would be a splitting method that evaluates transport
-terms and diffusion terms separately. The full residual is passed to
-the transport scheme which then applies a filter to drop all terms
-that are not labelled as transport; similarly for diffusion. We
-illustrate this procedure in figure \ref{fig:advdiff} where we show
-the definition of the residual for the advection diffusion equation
-for a scalar variable $q$ with a prescribed transport velocity
-$\MM{c}$ and diffusivity $\kappa$
+which manipulations to apply to which forms by applying filters. Here
+we will describe the objects required to define labels and filters
+before illustrating this with a simple example(s?).
 
-\begin{equation}
-  q_t + \MM{c}\cdot\nabla q - \kappa\nabla^2 q = 0,
+\subsection{Form Manipulation Language Objects}
+
+The finite element discretisation of each term in an equation is
+represented in Firedrake by a \texttt{ufl.Form} object. FML defines a
+\texttt{Term} object that comprises the \texttt{ufl.Form} defining the
+finite element discretisation of the term and a dictionary of
+key-value pairs corresponding to the FML \texttt{Label} objects that
+apply to this term. An FML \texttt{Label} object has a name and an
+optional value. When called on a \texttt{ufl.Form} object, the
+\texttt{Label} ** Labels can also be used to tag parts of the
+form. For example, the \texttt{subject} label is used to tag the
+\texttt{Function} in the \texttt{ufl.Form} that represents the
+prognostic variable. These operations are composable, so more than one
+label can be attached to a term.
+
+The end result is a \texttt{LabelledForm} object that holds a list of
+terms, pairing individual \texttt{ufl.Form} objects with an FML
+\texttt{label}. Labelled forms can be added together, multiplied by
+floats and have additional terms added. More complex manipulations are
+enabled by the \texttt{label\_map} method that provides a way of
+filtering the terms in the form based on their labels, applying
+different mappings to terms that pass or fail the conditions defined
+in the filter. A common operation might be to filter for terms that
+have a certain label, such as \texttt{transport}, drop terms that do
+not have this label and multiply those that do by the timestep. We
+shall demonstrate this in the following example.
+
+**describe methods has_label and get
+
+\subsection{Example}
+
+We demonstrate the use of FML by applying a split timestepping scheme
+to the 1D advection diffusion equation
+\begin{equation}\label{eq:1dadvdiff}
+  q_t + cq_x - \kappa q_{xx} = 0,
 \end{equation}
-with the terms labelled either \texttt{time\_derivative},
-\texttt{transport} or \texttt{diffusion}. The timestepping loop then
-applies a step of forward Euler to the transport term and backwards
-Euler to the diffusive term, i.e. the time discretisation is
+where $q$ is a scalar field, $c$ is the constant transporting velocity
+and $\kappa$ is the constant diffusivity. The time discretisation
+applies an explicit step of forward Euler to the transport term and
+an implicit step of backward Euler to the diffusion term:
+\begin{align}
+  q^* &= q^n - c \Delta t q_x^n, \label{eq:explicit-advdiff} \\
+  q^{n+1} - \kappa \Delta t q_{xx}^{n+1} &= q^*, \label{eq:implicit-advdiff}
+\end{align}
+where $q^n$ is the value of $q$ at time step $n$, $q^*$ is an
+intermediate value for $q$ and $\Delta t$ is the timestep.
 
-\begin{align}
-  q^* &= q^n - \Delta t (\MM{c}\cdot\nabla) q^n, \\
-  q^{n+1} &= q^* + \Delta t \kappa \nabla^2 q^{n+1}.
-\end{align}
+We illustrate the use of FML for this example in three steps: 1)
+define and label the residual; 2) build the explicit and implicit
+solvers; 3) run the timestepping loop. Since the intention of this
+example is to demonstrate FML rather than the compatible finite
+element discretisation, we use piecewise quadratic continuous Lagrange
+polynomials, $P_2$, for the finite element discretisation as this
+greatly simplifies the forms. The finite element discretisation of
+equation \ref{eq:1dadvdiff} is then
+\begin{equation}\label{eq:weak-advdiff}
+  \int_\Omega vq \diff x + c \int_\Omega v q_x \diff x + \kappa \int_\Omega v_x q_x \diff x = 0 \quad \forall v \in P_2,
+\end{equation}
+where $v\in P_2$ is the test function and the diffusion term has been
+integrated by parts.
+
+%\begin{align}\label{eq:weak-advdiff}
+%  \int_\Omega vq^* \diff x &= \int_\Omega vq^n \diff x - \Delta t c \int_\Omega v q_x^n \diff x, &\quad \forall v \in P^2, \\
+%  \int_\Omega vq^{n+1} \diff x  + \Delta t \kappa \int_\Omega v_x q_x^{n+1} \diff x &= \int_\Omega vq^* \diff x, &\quad \forall v \in P^2,
+%\end{align}
+%where $v$ is the test function and the diffusion term has been
+%integrated by parts.
+
+Listing \ref{lst:define-residual} presents the code to define this
+spatial discretisation of the residual with the $P_2$ function space
+denoted by \texttt{V}. Here we omit the code that sets up the mesh,
+function spaces and parameters $c$ and $\kappa$, although this can be
+found in the supplementary materials. The first two lines define the
+test function \texttt{v} and function \texttt{q}, corresponding to $v$
+and $q$ in equation \ref{eq:weak-advdiff}. The next four lines define
+the residual, i.e. the left hand side of equation
+\ref{eq:weak-advdiff}, using the FML labels \texttt{time\_derivative},
+\texttt{transport} and \texttt{diffusion} to tag the corresponding
+terms. The final line labels the function \texttt{q} as the subject of
+this residual. Note that the labels used for tagging just take the UFL
+form as an argument and their value is simply the default, \texttt{True}, in
+contrast to the \texttt{subject} label that has the value \texttt{q}.
+
+\lstinputlisting[caption=blah, label={lst:define-residual}, language=python, firstline=24, lastline=31]{code/advection_diffusion_splitting_example.py}
+
+To define the solvers for the explicit and implicit steps, we make use
+of the predefined mappings \texttt{drop} and \texttt{keep}
+
+\lstinputlisting[caption=blah, label={lst:define-explicit}, language=python, firstline=38, lastline=53]{code/advection_diffusion_splitting_example.py}
+
+\lstinputlisting[caption=blah, label={lst:define-implicit}, language=python, firstline=56, lastline=70]{code/advection_diffusion_splitting_example.py}
+
+\lstinputlisting[caption=blah, label={lst:timeloop}, language=python, firstline=56, lastline=70]{code/advection_diffusion_splitting_example.py}
 
@@
 \item spherical baroclinic wave, Euler - whatever people usually plot
-  (horizontal slice pressure contours, temperature shadng) (Dan)
+  (horizontal slice pressure contours, temperature shading) (Dan)
 \item Held Suarez??
@@
 testing new discretisations. This test suite is usually augmented by
-the unstable barotropic jet described in \citet{galewsky2004initial}
+the unstable barotropic jet described in
+\citet{galewsky2004initial}. Here we present convergence results for a
+variety of timestepping schemes applied to the steady state
+geostrophic balance flow (test case 2 from
+\citet{williamson1992standard}) and a demonstration of two different
+grids and HDiv element families for shallow water flow over a mountain
+(test case 5 from \citet{williamson1992standard}) and the unstable
+barotropic jet \citep{galewsky2004initial}.
+
+\subsubsection{Convergence results}
+
+\subsubsection{Grids and HDiv elements}
+
+\subsection{Compressible Boussinesq equations}
+
+\subsection{Compressible Euler equations}
+
+\subsubsection{Dense bubble in a vertical slice}
+
+\subsubsection{3D moist rising bubble}
+
+\subsubsection{Baroclinic wave on the sphere}
+
+\subsubsection{Held Suarez}
 

### 2024-07-24 source/code/advection_diffusion_splitting_example.py

@@ -1,0 +1,81 @@
+from firedrake import (PeriodicIntervalMesh, SpatialCoordinate, Constant,
+                       FunctionSpace, Function, TestFunction, TrialFunction,
+                       LinearVariationalProblem, LinearVariationalSolver,
+                       exp, dx)
+from firedrake.fml import all_terms, drop, keep, replace_subject, subject
+from firedrake.pyplot import plot
+from gusto.core import time_derivative, transport, diffusion
+import matplotlib.pyplot as plt
+
+L = 1
+nx = 20
+mesh = PeriodicIntervalMesh(nx, L)
+x = SpatialCoordinate(mesh)[0]
+
+V = FunctionSpace(mesh, "Lagrange", 2)
+d = 0.05
+qn = Function(V).interpolate(exp(-((x-0.5*L)/d)**2))
+plot(qn)
+plt.show()
+c = Constant(0.5)
+kappa = Constant(0.01)
+
+# Equation
+q = Function(V)
+v = TestFunction(V)
+residual = (
+    time_derivative(q * v * dx) +
+    transport(c * q.dx(0) * v * dx) +
+    diffusion(kappa * q.dx(0) * v.dx(0) * dx)
+    )
+residual = subject(residual, q)
+
+# Set up solvers
+q_trial = TrialFunction(V)
+dt = 0.01
+
+# Set up solver for explicit transport step
+lhs_explicit = residual.label_map(lambda t: t.has_label(time_derivative),
+                                  map_if_true=keep,
+                                  map_if_false=drop)
+lhs_explicit = lhs_explicit.label_map(all_terms,
+                                      map_if_true=replace_subject(q_trial))
+rhs_explicit = residual.label_map(lambda t: t.has_label(diffusion),
+                                  map_if_true=drop)
+rhs_explicit = rhs_explicit.label_map(lambda t: t.has_label(time_derivative),
+                                      map_if_false=lambda t: -dt*t)
+rhs_explicit = rhs_explicit.label_map(all_terms,
+                                      map_if_true=replace_subject(qn))
+qstar = Function(V)
+prob_explicit = LinearVariationalProblem(lhs_explicit.form,
+                                         rhs_explicit.form,
+                                         qstar)
+solver_explicit = LinearVariationalSolver(prob_explicit)
+
+# Set up solver for implicit transport step
+lhs_implicit = residual.label_map(lambda t: t.has_label(transport),
+                                  map_if_true=drop)
[…]

### 2024-07-24 source/paper.tex
tiny bit more progress

@@
 We illustrate the use of FML for this example in three steps: 1)
-define and label the residual; 2) build the explicit and implicit
-solvers; 3) run the timestepping loop. Since the intention of this
-example is to demonstrate FML rather than the compatible finite
-element discretisation, we use piecewise quadratic continuous Lagrange
-polynomials, $P_2$, for the finite element discretisation as this
-greatly simplifies the forms. The finite element discretisation of
-equation \ref{eq:1dadvdiff} is then
+define and label the weak form of the residual; 2) build the solvers
+for the explicit and implicit systems; 3) run the timestepping
+loop. Since the intention of this example is to demonstrate FML rather
+than the compatible finite element discretisation, we use space of piecewise
+quadratic continuous Lagrange polynomials, denoted by $P_2$, for the finite
+element discretisation as this greatly simplifies the forms. The
+finite element discretisation of equation \ref{eq:1dadvdiff} is then
 \begin{equation}\label{eq:weak-advdiff}
-  \int_\Omega vq \diff x + c \int_\Omega v q_x \diff x + \kappa \int_\Omega v_x q_x \diff x = 0 \quad \forall v \in P_2,
+  \int_\Omega vq_t \diff x + c \int_\Omega v q_x \diff x + \kappa \int_\Omega v_x q_x \diff x = 0, \quad \forall v \in P_2,
 \end{equation}
@@
 integrated by parts.
-
-%\begin{align}\label{eq:weak-advdiff}
-%  \int_\Omega vq^* \diff x &= \int_\Omega vq^n \diff x - \Delta t c \int_\Omega v q_x^n \diff x, &\quad \forall v \in P^2, \\
-%  \int_\Omega vq^{n+1} \diff x  + \Delta t \kappa \int_\Omega v_x q_x^{n+1} \diff x &= \int_\Omega vq^* \diff x, &\quad \forall v \in P^2,
-%\end{align}
-%where $v$ is the test function and the diffusion term has been
-%integrated by parts.
 
@@
 terms. The final line labels the function \texttt{q} as the subject of
-this residual. Note that the labels used for tagging just take the UFL
-form as an argument and their value is simply the default, \texttt{True}, in
-contrast to the \texttt{subject} label that has the value \texttt{q}.
+this residual, i.e. the prognostic field. Note that the labels used
+for tagging just take the UFL form as their argument and their value is
+simply the default, \texttt{True}, in contrast to the \texttt{subject}
+label that has the value \texttt{q}.
 
@@
 
-To define the solvers for the explicit and implicit steps, we make use
-of the predefined mappings \texttt{drop} and \texttt{keep}
+To define the solvers for the explicit and implicit steps we need to create UFL forms for the left and right sides, given by the following equations
+\begin{align}
+  \label{eq:explicit-advdiff}
+  \int_\Omega vq^* \diff x &= \int_\Omega vq^n \diff x - \Delta t c \int_\Omega v q_x^n \diff x, &\quad \forall v \in P^2, \\
+  \label{eq:implicit-advdiff}
+  \int_\Omega vq^{n+1} \diff x  + \Delta t \kappa \int_\Omega v_x q_x^{n+1} \diff x &= \int_\Omega vq^* \diff x, &\quad \forall v \in P^2.
+\end{align}
+To do this we use the \texttt{label\_map} method to loop over the
+terms in the residual making use of the predefined mappings
+\texttt{drop} and \texttt{keep} to either remove or retain the term
+based on the value of the filter. This code is given in listings
+\ref{lst:define-explicit} and \ref{lst:define-implicit}. In listing
+\ref{lst:define-explicit} we create the left and right hand sides of
+equation \ref{eq:explicit-advdiff}. The left hand side should only
+contain the time derivative term so the filter checks if each term has
+the label \texttt{time\_derivative}, keeping the one that does and
+dropping all others. To make this the left hand side of a
+\texttt{LinearVariationalProblem}, we replace the subject with a trial
+function using the predefined \texttt{replace\_subject} function, a
+lightweight wrapper around the UFL replace method which extracts the
+subject of the term and replaces it with the new expression, in this
+case simply a trial function \texttt{q\_trial}. \JScomment{Is the
+  labelling for the rhs the clearest? maybe we should keep the terms
+  that we want rather than assuming that dropping diffusion gives the
+  right thing?}
 

### 2024-07-25 source/paper.tex
minor updates

@@
 equation class. The timestepping methods are then able to reason about
-which manipulations to apply to which forms by applying filters. Here
-we will describe the objects required to define labels and filters
-before illustrating this with a simple example(s?).
-
-\subsection{Form Manipulation Language Objects}
+which manipulations to apply to which forms by applying filters that
+check for the presence of a label and, optionally, its value. Here we
+will describe the objects required to define labels and filters before
+illustrating this with a simple example(s?).
 
@@
 apply to this term. An FML \texttt{Label} object has a name and an
-optional value. When called on a \texttt{ufl.Form} object, the
-\texttt{Label} ** Labels can also be used to tag parts of the
+optional value, defaulting to \texttt{True} which indicates the
+presence of the label. Labels can also be used to tag parts of the
 form. For example, the \texttt{subject} label is used to tag the
@@
 prognostic variable. These operations are composable, so more than one
-label can be attached to a term.
-
-The end result is a \texttt{LabelledForm} object that holds a list of
+label can be attached to a term. The end result of labelling the terms
+in the residual is a \texttt{LabelledForm} object that holds a list of
 terms, pairing individual \texttt{ufl.Form} objects with an FML

### 2024-07-25 source/paper.tex
draft of FML example finished

@@
 convention in Gusto is to write the equations in residual form,
-meaning that all the terms appear on the left hand side. We will
-follow this convention below in contrast to many papers which present
-the equations with the pressure gradient term and any forcing terms on
-the right hand side.
+meaning that all the terms appear on the left side. We will follow
+this convention below in contrast to many papers which present the
+equations with the pressure gradient term and any forcing terms on the
+right side.
 
@@
 apply to this term. An FML \texttt{Label} object has a name and an
-optional value, defaulting to \texttt{True} which indicates the
-presence of the label. Labels can also be used to tag parts of the
-form. For example, the \texttt{subject} label is used to tag the
+optional value, defaulting to \texttt{True} to indicate the presence
+of the label. Labels can also be used to tag parts of the form. For
+example, the \texttt{subject} label is used to tag the
 \texttt{Function} in the \texttt{ufl.Form} that represents the
@@
 in the residual is a \texttt{LabelledForm} object that holds a list of
-terms, pairing individual \texttt{ufl.Form} objects with an FML
-\texttt{label}. Labelled forms can be added together, multiplied by
-floats and have additional terms added. More complex manipulations are
-enabled by the \texttt{label\_map} method that provides a way of
-filtering the terms in the form based on their labels, applying
-different mappings to terms that pass or fail the conditions defined
-in the filter. A common operation might be to filter for terms that
-have a certain label, such as \texttt{transport}, drop terms that do
-not have this label and multiply those that do by the timestep. We
-shall demonstrate this in the following example.
+terms, pairing individual \texttt{ufl.Form} objects with FML
+labels. Labelled forms can be added together, multiplied by floats and
+have additional terms added. More complex manipulations are enabled by
+the \texttt{label\_map} method that provides a way of filtering the
+terms in the form based on their labels, applying different mappings
+to terms that pass or fail the conditions defined in the filter. A
+common operation might be to filter for terms that have a certain
+label, such as \texttt{transport}, drop terms that do not have this
+label and multiply those that do by the timestep. We shall demonstrate
+this in the following example.
 
-**describe methods has_label and get
+\JScomment{describe methods has label and get?}
 
@@
 We illustrate the use of FML for this example in three steps: 1)
-define and label the weak form of the residual; 2) build the solvers
-for the explicit and implicit systems; 3) run the timestepping
-loop. Since the intention of this example is to demonstrate FML rather
-than the compatible finite element discretisation, we use space of piecewise
-quadratic continuous Lagrange polynomials, denoted by $P_2$, for the finite
-element discretisation as this greatly simplifies the forms. The
-finite element discretisation of equation \ref{eq:1dadvdiff} is then
+define and label the weak form of the residual; 2) build solvers for
+the explicit and implicit systems; 3) run the timestepping loop. Since
+the intention of this example is to demonstrate FML rather than the
+compatible finite element discretisation, we use space of piecewise
+quadratic continuous Lagrange polynomials, denoted by $P_2$, for the
+finite element discretisation as this greatly simplifies the
+forms. The finite element discretisation of equation
+\ref{eq:1dadvdiff} is then
 \begin{equation}\label{eq:weak-advdiff}
@@
 and $q$ in equation \ref{eq:weak-advdiff}. The next four lines define
-the residual, i.e. the left hand side of equation
-\ref{eq:weak-advdiff}, using the FML labels \texttt{time\_derivative},
-\texttt{transport} and \texttt{diffusion} to tag the corresponding
-terms. The final line labels the function \texttt{q} as the subject of
-this residual, i.e. the prognostic field. Note that the labels used
-for tagging just take the UFL form as their argument and their value is
-simply the default, \texttt{True}, in contrast to the \texttt{subject}
-label that has the value \texttt{q}.
+the residual, i.e. the left side of equation \ref{eq:weak-advdiff},
+using the FML labels \texttt{time\_derivative}, \texttt{transport} and
+\texttt{diffusion} to tag the corresponding terms. The final line
+labels the function \texttt{q} as the subject of this residual,
+i.e. the prognostic field. Note that the labels used for tagging just
+take the UFL form as their argument and their value is simply the
+default, \texttt{True}, in contrast to the \texttt{subject} label that
+has the value \texttt{q}.
 
@@
 \ref{lst:define-explicit} and \ref{lst:define-implicit}. In listing
-\ref{lst:define-explicit} we create the left and right hand sides of
-equation \ref{eq:explicit-advdiff}. The left hand side should only
-contain the time derivative term so the filter checks if each term has
-the label \texttt{time\_derivative}, keeping the one that does and
-dropping all others. To make this the left hand side of a
+\ref{lst:define-explicit} we create the left and right sides of
+equation \ref{eq:explicit-advdiff}. The left side should only contain
+the time derivative term so the filter checks if each term has the
+label \texttt{time\_derivative}, keeping the one that does and
+dropping all others. To make this the left side of a
 \texttt{LinearVariationalProblem}, we replace the subject with a trial
@@
 subject of the term and replaces it with the new expression, in this
-case simply a trial function \texttt{q\_trial}. \JScomment{Is the
+case simply a trial function \texttt{q\_trial}. The right side of the
+explicit solver has two forms, one coming from the discretisation of
+the time derivative and the other from the transport term. Both are
+evaluated using the values of $q$ at time step $n$. We build this by
+first dropping the diffusion term (line 6) and then multiplying the
+terms that do not have the \texttt{time\_derivative} label by $-\Delta
+t$ (line 8), where the minus sign is required to move the term from
+the left side to the right side. The final step is to replace the
+subject with the function \texttt{qn} that holds the values of $q$ at
+time step $n$. We can then extract the UFL forms from
+\texttt{lhs\_explicit} and \texttt{lhs\_implicit} to create a
+\texttt{LinearVariationalProblem} and corresponding
+\texttt{LinearVariationalSolver} that, when solved, will put the
+result into the function \texttt{qstar}.
+
+\JScomment{Is the
   labelling for the rhs the clearest? maybe we should keep the terms
@@
 
+The implicit solver is defined in a similar way (see listing
+\ref{lst:define-implicit}). To create the left side of equation
+\ref{eq:implicit-advdiff} we begin with the original residual, drop
+the transport term (line 1), multiply all remaining terms that are not
+the time derivative by $\Delta t$ (line 3) and replace the subject
+with a trail function (line 5). The right side only contains the form
+coming from the discretisation of the time derivative (line 7),
+evaluated using the result of the explicit solve, \texttt{qstar} (line
+9).
+
 \lstinputlisting[caption=blah, label={lst:define-implicit}, language=python, firstline=56, lastline=70]{code/advection_diffusion_splitting_example.py}
 
-\lstinputlisting[caption=blah, label={lst:timeloop}, language=python, firstline=56, lastline=70]{code/advection_diffusion_splitting_example.py}
+The final part of the code is the time loop which calls the solve
+method of the explicit and then the impicit solvers defined above, see
+listing \ref{lst:timeloop}
+
+\lstinputlisting[caption=blah, label={lst:timeloop}, language=python, firstline=73, lastline=79]{code/advection_diffusion_splitting_example.py}
 

### 2024-07-25 source/paper.tex
add captions to listings and make a few changes

@@
 not specific to any particular equation set. This is also true of the
-implementation of the timestepping schemes. This enables flexibility
-and reproducibility (**elaborate). This functionality is enabled
+implementation of the timestepping schemes, resulting in modular code
+where the equations are defined separately from the timestepping
+scheme enabling flexibility in combining different equations with
+different timestepping schemes. This powerful functionality is enabled
 through the form manipulation language (FML) which is used to
@@
 represented in Firedrake by a \texttt{ufl.Form} object. FML defines a
+Label object which tags a ufl.Form to become a Term, with the
+intention that the Term represents a term in the equation.
+
+FML defines a
 \texttt{Term} object that comprises the \texttt{ufl.Form} defining the
@@
 this in the following example.
-
-\JScomment{describe methods has label and get?}
 
@@
 
-\lstinputlisting[caption=blah, label={lst:define-residual}, language=python, firstline=24, lastline=31]{code/advection_diffusion_splitting_example.py}
+\lstinputlisting[caption={This code defines and labels the residual corresponding to equation \ref{eq:weak-advdiff}}, label={lst:define-residual}, language=python, firstline=24, lastline=31]{code/advection_diffusion_splitting_example.py}
 
@@
 
-\lstinputlisting[caption=blah, label={lst:define-explicit}, language=python, firstline=38, lastline=53]{code/advection_diffusion_splitting_example.py}
+\lstinputlisting[caption={This code defines the solver for the explicit step of the timestepping method i.e.
+equation \ref{eq:explicit-advdiff}.
+\texttt{qn} and \texttt{qstar} are Firedrake \texttt{Functions} with \texttt{qn} holding the values of $q$ at time step $n$ and \texttt{qstar} holding the values after the explicit solve.}, label={lst:define-explicit}, language=python, firstline=40, lastline=55]{code/advection_diffusion_splitting_example.py}
 
@@
 
-\lstinputlisting[caption=blah, label={lst:define-implicit}, language=python, firstline=56, lastline=70]{code/advection_diffusion_splitting_example.py}
+\lstinputlisting[caption={This code defines the solver for the implicit step of the timestepping method i.e.
+equation \ref{eq:implicit-advdiff}.
+\texttt{qstar} and \texttt{qnp1} are Firedrake \texttt{Functions} with \texttt{qstar} holding the values after the explicit solve and \texttt{qnp1} holding the values of $q$ at after the implicit solve.}, label={lst:define-implicit}, language=python, firstline=57, lastline=71]{code/advection_diffusion_splitting_example.py}
 
 The final part of the code is the time loop which calls the solve
-method of the explicit and then the impicit solvers defined above, see
-listing \ref{lst:timeloop}
+method of first the explicit solver and then the implicit solver, see
+listing \ref{lst:timeloop}. In practice, the code in listings
+\ref{lst:define-explicit}-\ref{lst:define-implicit} would be in a
+function that could be applied to any residual labeled with the
+appropriate labels.
 

### 2024-08-09 source/paper.tex
some edits

@@
 
-The finite element method represents fields on a mesh made of
-non-overlapping, conforming cells, or elements. Within each element,
-the field is represented as a sum of coefficients multiplied by
-polynomial basis functions. The finite element space is specified by
-the degree of the polynomial basis functions and the continuity of
-these functions between the elements. The partial differential
-equations used in weather and climate modelling have multiple
-prognostic fields and it is not necessary, or indeed desirable, to use
-the same finite element function space for each field. Using different
-finite element function spaces for each field gives what is known as a
-mixed finite element discretisation. If these different finite element
-function spaces are chosen so that discrete versions of the vector
-calculus identities ** hold then the discretisation is called
-`mimetic' or `compatible'.
+The finite element method represents fields (e.g. velocity, pressure,
+density or temperature) on a mesh made of non-overlapping, conforming
+cells, or elements. Within each element, the field is represented as a
+sum of coefficients multiplied by polynomial basis functions. The
+finite element space is specified by the degree of the polynomial
+basis functions and the continuity of these functions between the
+elements. The partial differential equations used in weather and
+climate modelling have multiple prognostic fields and it is not
+necessary, or indeed desirable, to use the same finite element
+function space for each field. Using different finite element function
+spaces for each field gives what is known as a mixed finite element
+discretisation. If these different finite element function spaces are
+chosen so that discrete versions of vector calculus identities hold
+then the discretisation is called `mimetic' or `compatible'.
 
@@
 modelling. In section \ref{sec: governing} we will outline three
-different equation sets, along with their finite element
-discretisations. Then in section \ref{sec: design} we will present the
-code structure of Gusto followed by a detailed example (section
-\ref{sec: FML}) of the Form Manipulation Language (FML) that
-facilitates symbolic manipulation of the finite element forms within
-our timestepping code. Section \ref{sec: results} contains a selection
-of numerical results that showcases the range of discretisations
+different equations commonly used for modelling geophysical fluids in
+order to demonstrate the similaries between the equations and justify
+the use of a unified codebase. The advantages of this unified approach
+become even clearer in section \ref{sec: FEM} where we present the
+compatible finite element discretisation of the different types of
+terms (e.g. pressure gradient, transport) seen in section \ref{sec:
+  governing}. In section \ref{sec: design} we describe the code
+structure of Gusto followed by a detailed example (section \ref{sec:
+  FML}) of the Form Manipulation Language (FML) that facilitates
+symbolic manipulation of the finite element forms within our
+timestepping code. Section \ref{sec: results} contains a selection of
+numerical results that showcases the range of discretisations
 available within Gusto. **mention here new functionality facilitated
-by the flexibility of Gusto? We summarise our current status and future
-directions in section \ref{sec: summary}.
+by the flexibility of Gusto? We summarise our current status and
+future directions in section \ref{sec: summary}.
 
@@
 equations in terms of the prognostic variables $\MM{u}$, $b$ and $p$
-where $\MM{u}$ is a three dimensional velocity field, $b$ is the fluid
-buoyancy and $p$ is pressure. In terms of these variables, the
-equations are:
+where $\MM{u}$ is the velocity field, $b$ is the fluid buoyancy and
+$p$ is pressure. In terms of these variables, the equations are:
 
@@
 $\hat{\MM{k}}$ is the unit vector pointing upwards and $c$ is the
-acoustic wave speed. The velocity $\MM{u}$ can be two or three
-dimensional, depending on whether the domain is a vertical slice
-described by $(x, z)$ coordinates, of a three dimensional domain
-described by $(x, y, z)$ coordinates.
+acoustic wave speed. The equations have been written in
+three-dimensional form but they can also be solved in a
+two-dimensional vertical slice domain described by $(x, z)$
+coordinates. In this case the velocity $\MM{u}$ can either be
+two-dimensional as well (if there is no Coriolis force), or fully
+three-dimensional but with each cartensian component only varying in
+the $x-z$ plane. In the latter case this is implemented using a
+three-dimensional computational domain that is only one element wide
+in the $y$-direction and with periodic boundary conditions also in
+that direction.
 
@@
 \end{align}
-where $\Pi$ is the Exner pressure, $p_0$ is a reference pressure,
-typically taken to be *, $R$ is ** and $\kappa$ is **. The velocity
-$\MM{u}$ can be two or three dimensional, depending on whether the
-domain is a vertical slice described by $(x, z)$ coordinates, of a
-three dimensional domain described by $(x, y, z)$ coordinates.
+where $\Pi$ is the Exner pressure, $p_0$ is a reference pressure, $R$
+is the gas constant, $c_p$ is the specific heat at constant pressure
+and $\kappa=R/c_p$. The equations have been written in
+three-dimensional form but they can also be solved in a
+two-dimensional vertical slice domain described by $(x, z)$
+coordinates. In this case the velocity $\MM{u}$ can either be two
+dimensional as well (if there is no Coriolis force), or fully three
+dimensional but with each cartensian component only varying in the
+$x-z$ plane. In the latter case this is implemented using a three
+dimensional computational domain that is only one element wide in the
+$y$-direction and with periodic boundary conditions also in that
+direction. \JScomment{I know this repeats what is said above but I
+  like that each section is self-contained - I hope that is ok but
+  also happy to change it!}
 
@@
 finite element discretisations, as we shall see below, and importantly
-this means that the discretisation of these terms can be written in a
-general way that means it can be applied to any of the equation sets.
+this means that the discretisation of these terms can be written and
+implemented in software so that it can be applied to any of the
+equation sets.
 
@@
 gradient term is chosen to be in the space $\mathbb{V}_p$ with
-$\mathbb{V}_p \subset L_2$. The compressible Boussinesq and
-compressible Euler equations have a third prognostic variable, either
-buoyancy or potential temperature, which is chosen to be in the space
-$\mathbb{V}_\theta$ which has degrees of freedom that are collocated
-with those of the vertical velocity, mimicking the Charney-Phillips
-staggering as discussed in **. These choices are governed by the
-properties required of discretisation of the linearised equation sets
-(as discussed in section ** and other refs). To solve the nonlinear
-equation sets we need to introduce stable and accurate discretisations
-of the transport terms and the nonlinear pressure gradient term in the
-compressible Euler equation set. In addition, some test cases require
-the addition of viscous and diffusive terms to generate converged
-solutions \ref{}. In the following subsections we illustrate a
-non-exhaustive (**probably!) selection of the options available in
-Gusto.
+$\mathbb{V}_p \subset L_2$, or, in the case of the more complex
+nonlinear pressure gradient term for the compressible Euler equations,
+it is a function of a variable in this space. The compressible
+Boussinesq and compressible Euler equations have a third prognostic
+variable, either buoyancy or potential temperature, which is chosen to
+be in the space $\mathbb{V}_\theta$ which has degrees of freedom that
+are collocated with those of the vertical velocity, mimicking the
+Charney-Phillips staggering as discussed in **. These choices are
+governed by the properties required of discretisation of the
+linearised equation sets (as discussed in section ** and other
+refs). To solve the nonlinear equation sets we need to introduce
+stable and accurate discretisations of the transport terms and the
+nonlinear pressure gradient term in the compressible Euler equation
+set. In addition, some test cases require viscous and diffusive terms
+to generate converged solutions \ref{}. In the following subsections
+we illustrate a non-exhaustive (**probably!) selection of the options
+available in Gusto.
 
@@
 \begin{equation}
-\int_\Omega\MM{w}\cdot\nabla p \diff x = -\int_\Omega\nabla\cdot\MM{w}p\diff x,
+\int_\Omega\MM{w}\cdot\nabla p \diff x = -\int_\Omega\nabla\cdot\MM{w}p\diff x, \quad \forall \MM{w} \in \mathbb{V}_u,
 \end{equation}
@@
 compressible Euler equations, $c_p\theta\nabla\Pi$, is nonlinear and
-this introduces an additional complication. Integrating by parts is
+this introduces an additional complication. Integration by parts is
 still required since $\Pi$ is evaluated as a function of the density
-$\rho \in L_2$ and so is discontinuous, but this means that we now
-have a differential operator applied to $\theta \in \mathbb{V}_\theta$
-which is discontinuous in the vertical. Following \ref{} we integrate
-by parts in each element $e$, giving
+$\rho \in \mathbb{V}_\rho$ and $\theta \in \mathbb{V}_\theta$ so is
+discontinuous, but this means that we now have a differential operator
+applied to $\theta \in \mathbb{V}_\theta$ which is discontinuous in
+the vertical. Following \ref{} we integrate by parts in each element
+$e$, giving
 
@@
 The finite element discretisation of each term in an equation is
-represented in Firedrake by a \texttt{ufl.Form} object. FML defines a
-Label object which tags a ufl.Form to become a Term, with the
-intention that the Term represents a term in the equation.
+represented in Firedrake by a UFL form object. FML defines a
+\texttt{Label} object which is used to tag a UFL form to create a
+\texttt{Term}, with the intention that the \texttt{Term} represents a
+term in the residual.
 
-FML defines a
-\texttt{Term} object that comprises the \texttt{ufl.Form} defining the
-finite element discretisation of the term and a dictionary of
-key-value pairs corresponding to the FML \texttt{Label} objects that
-apply to this term. An FML \texttt{Label} object has a name and an
-optional value, defaulting to \texttt{True} to indicate the presence
-of the label. Labels can also be used to tag parts of the form. For
-example, the \texttt{subject} label is used to tag the
-\texttt{Function} in the \texttt{ufl.Form} that represents the
-prognostic variable. These operations are composable, so more than one
-label can be attached to a term. The end result of labelling the terms
-in the residual is a \texttt{LabelledForm} object that holds a list of
-terms, pairing individual \texttt{ufl.Form} objects with FML
-labels. Labelled forms can be added together, multiplied by floats and
-have additional terms added. More complex manipulations are enabled by
-the \texttt{label\_map} method that provides a way of filtering the
-terms in the form based on their labels, applying different mappings
-to terms that pass or fail the conditions defined in the filter. A
-common operation might be to filter for terms that have a certain
-label, such as \texttt{transport}, drop terms that do not have this
-label and multiply those that do by the timestep. We shall demonstrate
-this in the following example.
+FML defines a \texttt{Term} object that comprises the
+\texttt{ufl.Form} defining the finite element discretisation of the
+term and a dictionary of key-value pairs corresponding to the FML
+\texttt{Label} objects that apply to this term. An FML \texttt{Label}
+object has a name and an optional value, defaulting to \texttt{True}
+to indicate the presence of the label. Labels can also be used to tag
+parts of the form. For example, the \texttt{subject} label is used to
+tag the \texttt{Function} in the \texttt{ufl.Form} that represents the
+prognostic variable, or `subject' of the equation. These operations
+are composable, so more than one label can be attached to a term. The
+end result of labelling the terms in the residual is a
+\texttt{LabelledForm} object that holds a list of terms, pairing
+individual \texttt{ufl.Form} objects with FML labels. Labelled forms
+can be added together, multiplied by floats and have additional terms
+added. More complex manipulations are enabled by the
+\texttt{label\_map} method that provides a way of filtering the terms
+in the form based on their labels, applying different mappings to
+terms that pass or fail the conditions defined in the filter. A common
+operation might be to filter for terms that have a certain label, such
+as \texttt{transport}, drop terms that do not have this label and
+multiply those that do by the timestep. We shall demonstrate this in
+the following example.
 
@@
 and $\kappa$ is the constant diffusivity. The time discretisation
-applies an explicit step of forward Euler to the transport term and
-an implicit step of backward Euler to the diffusion term:
+applies an explicit timestep of forward Euler to the transport term
+$cq_x$ and an implicit timestep of backward Euler to the diffusion
+term $\kappa q_{xx}$:
 \begin{align}
@@
 We illustrate the use of FML for this example in three steps: 1)
-define and label the weak form of the residual; 2) build solvers for
-the explicit and implicit systems; 3) run the timestepping loop. Since
+define and label the weak form of the residual, 2) build solvers for
+the explicit and implicit systems, 3) run the timestepping loop. Since
 the intention of this example is to demonstrate FML rather than the
-compatible finite element discretisation, we use space of piecewise
+compatible finite element discretisation, we use the space of piecewise
 quadratic continuous Lagrange polynomials, denoted by $P_2$, for the
@@
 equation \ref{eq:explicit-advdiff}.
-\texttt{qn} and \texttt{qstar} are Firedrake \texttt{Functions} with \texttt{qn} holding the values of $q$ at time step $n$ and \texttt{qstar} holding the values after the explicit solve.}, label={lst:define-explicit}, language=python, firstline=40, lastline=55]{code/advection_diffusion_splitting_example.py}
+\texttt{qn} and \texttt{qstar} are Firedrake \texttt{Functions} with \texttt{qn} holding the values of $q$ at time step $n$ and \texttt{qstar} holding the values of $q$ after the explicit solve.}, label={lst:define-explicit}, language=python, firstline=40, lastline=55]{code/advection_diffusion_splitting_example.py}
 
@@
 the time derivative by $\Delta t$ (line 3) and replace the subject
-with a trail function (line 5). The right side only contains the form
+with a trial function (line 5). The right side only contains the form
 coming from the discretisation of the time derivative (line 7),
@@
 
-\lstinputlisting[caption=blah, label={lst:timeloop}, language=python, firstline=73, lastline=79]{code/advection_diffusion_splitting_example.py}
+\lstinputlisting[caption={This code is the timeloop implementing the scheme defined in equations \ref{eq:explicit-advdiff}-\ref{eq:implicit-advdiff} using the solvers defined in listings \ref{lst:define-explicit}-\ref{lst:define-implicit}.}, label={lst:timeloop}, language=python, firstline=73, lastline=79]{code/advection_diffusion_splitting_example.py}
+
+This simple example demonstrates the power of manipulating the terms
+in the equation based on the presence or absence of certain
+labels. The full strength of FML becomes even more apparant when we
+exploit information stored in the label value. We use this in two
+important situations in Gusto. The first situation is when we require
+a linearisation of the residual, for example when applying a Newton
+method to the full nonlinear equations \JScomment{should we clarify
+  that we mean a specific linearisation because Firedrake can do
+  Newton methods without FML?}, or to test a new algorithm or model
+setup on linear equations. The second situation is when we need to
+evaluate source terms that come from physics schemes. We can
+illustrate both situations with simple extensions to the above
+example.
+
+First consider the 1D viscous Burgers' equation
+\begin{equation}
+  u_t + uu_x + \nu u_{xx} = 0.
+\end{equation}
 

### 2024-08-09 source/code/advection_diffusion_splitting_example.py

@@ -32,6 +32,8 @@ residual = subject(residual, q)
 
 # Set up solvers
 q_trial = TrialFunction(V)
+qstar = Function(V)
+qnp1 = Function(V)
 dt = 0.01
 
 # Set up solver for explicit transport step
@@ -46,7 +48,6 @@ rhs_explicit = rhs_explicit.label_map(lambda t: t.has_label(time_derivative),
                                       map_if_false=lambda t: -dt*t)
 rhs_explicit = rhs_explicit.label_map(all_terms,
                                       map_if_true=replace_subject(qn))
-qstar = Function(V)
 prob_explicit = LinearVariationalProblem(lhs_explicit.form,
                                          rhs_explicit.form,
                                          qstar)
@@ -63,7 +64,6 @@ rhs_implicit = residual.label_map(lambda t: t.has_label(time_derivative),
                                       map_if_false=drop)
 rhs_implicit = rhs_implicit.label_map(all_terms,
                                       map_if_true=replace_subject(qstar))
-qnp1 = Function(V)
 prob_implicit = LinearVariationalProblem(lhs_implicit.form,
                                          rhs_implicit.form,
                                          qnp1)

### 2024-08-16 source/paper.tex
tweaks

@@
 $\rho \in \mathbb{V}_\rho$ and $\theta \in \mathbb{V}_\theta$ so is
-discontinuous, but this means that we now have a differential operator
-applied to $\theta \in \mathbb{V}_\theta$ which is discontinuous in
-the vertical. Following \ref{} we integrate by parts in each element
-$e$, giving
+discontinuous between elements, but then integrating by parts
+transfers the differential operator to $\MM{w}\theta$ which is
+discontinuous in the horizontal. Following \ref{} we integrate by parts
+in each element $e$, giving
 
@@
 (\Pi^++\Pi^-)/2$, where the $+$ and $-$ indicate each side of the
-facet. The brackets $\jump{q} = q^+-q^-$ denote the jump in a value
+facet. The brackets $\jump{q} = q^+-q^-$ denote the jump in $q$
 across the facet and the integral over the horizontal facets vanishes
@@
 where the equations are defined separately from the timestepping
-scheme enabling flexibility in combining different equations with
-different timestepping schemes. This powerful functionality is enabled
+scheme enabling flexibility in applying different timestepping schemes
+to different equations. This powerful functionality is facilitated
 through the form manipulation language (FML) which is used to

### 2024-08-16 source/paper.tex
plan for intro paragraphs

@@
 spaces for each field gives what is known as a mixed finite element
-discretisation. If these different finite element function spaces are
-chosen so that discrete versions of vector calculus identities hold
-then the discretisation is called `mimetic' or `compatible'.
+discretisation.
+
+If the different finite element function spaces are chosen so that
+discrete versions of vector calculus identities hold then the
+discretisation is called `mimetic' or `compatible'.
+
+* These methods are useful in a variety of applications
+* In the context of geophysical fluid dynamics they are useful because...
+* These properties rely on the underpinning structure of the de Rham complex
+* In 3D the function spaces have a tensor product structure
+* The function spaces are hard to construct especially on the sphere - piola transform
+
+* Gusto has grown out of code developed in (see above refs)
+* The idea is to bring together methods using compatible finite element discretisations in a well-tested and documented code base that can be extended and applied to different problems of interest to the community.
+* X did Y - quick history of the development of the methods, plus others that are not actually in Gusto at the moment.
 
@@
 variety of timestepping schemes applied to the steady state
-geostrophic balance flow (test case 2 from
+geostrophic balanced flow (test case 2 from
 \citet{williamson1992standard}) and a demonstration of two different

### 2024-08-19 source/paper.tex
work on the intro

@@
 discrete versions of vector calculus identities hold then the
-discretisation is called `mimetic' or `compatible'.
-
-* These methods are useful in a variety of applications
-* In the context of geophysical fluid dynamics they are useful because...
-* These properties rely on the underpinning structure of the de Rham complex
+discretisation is called `mimetic' or `compatible'. These methods are
+useful in a variety of applications due to their ability to capture
+the underlying geometric structure of the governing equations. In the
+context of geophysical fluid dynamics, these methods are useful
+because they represent geostrophic balance exactly which is crucial
+for reproducing accurate simulations of the large-scale flow. In
+addition to this, there is flexibility within the compatible finite
+element framework to choose function spaces that also give other
+advantages such as energy conservation, local mass conservation and
+discrete dispersion properties that lead to no spurious pressure
+modes.
+
+The compatible finite element spaces relevant for geophysical fluid
+dynamics are constructed from a discrete differential complex called
+the de Rham complex. This is a sequence of spaces with differential
+operators between them that map onto the kernal of the next operator
+in the sequence. This is often summarised in a commutation diagram as
+in figures \ref{fig:3D de Rham} and \ref{fig:2D de Rham}. In these
+diagrams the infinite dimensional spaces are along the top row and the
+$\pi_i$ denote projections into the corresponding finite dimensional
+subspace used in the compatible finite element discretisation. The
+differential operators are denoted by $\diff^k$ and these commute with
+the projections, meaning that $\diff^{k+1}\pi_k(\cdot) =
+\diff^{k+1}\pi_{k+1}(\cdot)$. This guarantees that, in 3D, $\nabla$
+maps onto the kernal of $\nabla\times$, ensuring
+$\nabla\times\nabla(\cdot)=0$, and $\nabla\times$ maps onto the kernal
+of $\nabla\cdot$, ensuring that $\nabla\cdot\nabla\times(\cdot)=0$;
+i.e. the spaces are compatible. In 2D we have that $\nabla^\perp$ maps
+onto the kernal of $\nabla\cdot$, ensuring that
+$\nabla\cdot\nabla^\perp(\cdot)=0$. It is possible to write this de
+Rham complex such that instead we have that $\nabla$ maps onto the
+kernal of $\nabla^\perp$, ensuring that
+$\nabla^\perp\cdot\nabla(\cdot)=0$, but the former is more useful in
+constructing discretisations for geophysical fluid dynamics since it
+enables schemes with local mass conservation \citep{}.
+
 * In 3D the function spaces have a tensor product structure
-* The function spaces are hard to construct especially on the sphere - piola transform
+* The function spaces are hard to construct especially on the sphere - piola transform, theta space
 

### 2024-08-22 source/paper.tex
latest progress - not happy with it

@@
 
+Given that there are many promising avenues for research in the area
+of compatible finite element discretisations for geophysical fluid
+dynamics (**where to say what these are?), a flexible, extensible
+implementation of the methods is strongly desirable. This is
+especially required because, while it is feasible for someone to write
+a code implementing a specific finite element discretisation of a PDE,
+the code quickly becomes complicated and unsustainable when
+investigating different orders or families of finite elements, or
+different formulations of the equation. For geophysical fluid dynamics
+applications there are additional considerations relating to the
+geometry and types of waves present in the flow. For example, the
+scales of motion in the horizontal and vertical directions differ by
+several orders of magnitude. (also: boundaries, vertical staggering
+for theta space)
+
 * In 3D the function spaces have a tensor product structure
@@
 compatible finite element discretisation of the different types of
-terms (e.g. pressure gradient, transport) seen in section \ref{sec:
-  governing}. In section \ref{sec: design} we describe the code
-structure of Gusto followed by a detailed example (section \ref{sec:
-  FML}) of the Form Manipulation Language (FML) that facilitates
-symbolic manipulation of the finite element forms within our
-timestepping code. Section \ref{sec: results} contains a selection of
-numerical results that showcases the range of discretisations
+terms (e.g. pressure gradient, transport, diffusion) seen in section
+\ref{sec: governing}. In section \ref{sec: design} we describe the
+code structure of Gusto followed by a detailed example (section
+\ref{sec: FML}) of the Form Manipulation Language (FML) that
+facilitates symbolic manipulation of the finite element forms within
+our timestepping code. Section \ref{sec: results} contains a selection
+of numerical results that showcases the range of discretisations
 available within Gusto. **mention here new functionality facilitated
@@
 
-\subsubsection{Grids and HDiv elements}
+\subsubsection{Grids and finite element families}
 

### 2025-02-06 source/paper.tex
fix FML intro

@@
 \texttt{Term}, with the intention that the \texttt{Term} represents a
-term in the residual.
-
-FML defines a \texttt{Term} object that comprises the
-\texttt{ufl.Form} defining the finite element discretisation of the
-term and a dictionary of key-value pairs corresponding to the FML
-\texttt{Label} objects that apply to this term. An FML \texttt{Label}
-object has a name and an optional value, defaulting to \texttt{True}
-to indicate the presence of the label. Labels can also be used to tag
-parts of the form. For example, the \texttt{subject} label is used to
-tag the \texttt{Function} in the \texttt{ufl.Form} that represents the
-prognostic variable, or `subject' of the equation. These operations
-are composable, so more than one label can be attached to a term. The
-end result of labelling the terms in the residual is a
-\texttt{LabelledForm} object that holds a list of terms, pairing
-individual \texttt{ufl.Form} objects with FML labels. Labelled forms
-can be added together, multiplied by floats and have additional terms
-added. More complex manipulations are enabled by the
-\texttt{label\_map} method that provides a way of filtering the terms
-in the form based on their labels, applying different mappings to
-terms that pass or fail the conditions defined in the filter. A common
-operation might be to filter for terms that have a certain label, such
-as \texttt{transport}, drop terms that do not have this label and
-multiply those that do by the timestep. We shall demonstrate this in
-the following example.
+term in the residual. The \texttt{Label} object consists of a name and
+an optional value, defaulting to \texttt{True} to indicate the
+presence of the label. Labels can be used to tag terms to identify
+them as representing particular processes, such as transport, or they
+can tag parts of the form. For example, the \texttt{subject} label is
+used to tag the \texttt{Function} in the \texttt{ufl.Form} that
+represents the prognostic variable, or `subject' of the
+equation. These operations are composable, so more than one label can
+be attached to a term. The end result of labelling the terms in the
+residual is a \texttt{LabelledForm} object that holds a list of terms,
+pairing individual \texttt{ufl.Form} objects with FML labels. Labelled
+forms can be added together, multiplied by floating point numbers and
+extended by adding additional terms. More complex manipulations are
+enabled by the \texttt{label\_map} method that provides a way of
+filtering the terms in the form based on their labels, applying
+different mappings to terms that pass or fail the conditions defined
+in the filter. A common operation might be to filter for terms that
+have a certain label, such as \texttt{transport}, drop terms that do
+not have this label and multiply those that do by the timestep. We
+shall demonstrate this in the following example.
 

### 2025-04-03 source/paper.tex
more text

@@
 
-If the different finite element function spaces are chosen so that
-discrete versions of vector calculus identities hold then the
-discretisation is called `mimetic' or `compatible'. These methods are
-useful in a variety of applications due to their ability to capture
-the underlying geometric structure of the governing equations. In the
-context of geophysical fluid dynamics, these methods are useful
-because they represent geostrophic balance exactly which is crucial
-for reproducing accurate simulations of the large-scale flow. In
-addition to this, there is flexibility within the compatible finite
-element framework to choose function spaces that also give other
-advantages such as energy conservation, local mass conservation and
-discrete dispersion properties that lead to no spurious pressure
-modes.
+`Compatible' or `mimetic' finite element discretisations are
+constructed by choosing the different finite element function spaces
+so that discrete versions of continuous vector calculus identities
+hold. These methods are useful in a variety of applications due to
+their ability to capture the underlying geometric structure of the
+governing equations. In the context of geophysical fluid dynamics,
+these methods are useful because they represent geostrophic balance
+exactly which is crucial for reproducing accurate simulations of the
+large-scale flow. In addition to this, there is flexibility within the
+compatible finite element framework to choose function spaces that
+also give other advantages such as energy conservation, local mass
+conservation and discrete dispersion properties that lead to no
+spurious pressure modes.
 
@@
 in the sequence. This is often summarised in a commutation diagram as
-in figures \ref{fig:3D de Rham} and \ref{fig:2D de Rham}. In these
+in figures \ref{fig:2D de Rham} and \ref{fig:3D de Rham}. In these
 diagrams the infinite dimensional spaces are along the top row and the
@@
 the projections, meaning that $\diff^{k+1}\pi_k(\cdot) =
-\diff^{k+1}\pi_{k+1}(\cdot)$. This guarantees that, in 3D, $\nabla$
+\pi_{k+1}\diff^{k+1}(\cdot)$. This guarantees that, in 3D, $\nabla$
 maps onto the kernal of $\nabla\times$, ensuring
@@
 enables schemes with local mass conservation \citep{}.
+
+\begin{figure}\label{fig:2D de Rham}
+  \centering{\includegraphics{de_rham_fig_2d.pdf}}
+\end{figure}
+
+\begin{figure}\label{fig:3D de Rham}
+  \centering{\includegraphics{de_rham_fig_3d.pdf}}
+\end{figure}
 
@@
 
-To define the solvers for the explicit and implicit steps we need to create UFL forms for the left and right sides, given by the following equations
+To define the solvers for the explicit and implicit timesteps
+(equations \ref{eq:explicit-advdiff} and \ref{eq:implicit-advdiff}),
+we need to create UFL forms given by the following equations
 \begin{align}

### 2025-04-03 source/paper.tex
add some comments to sec1

@@
 
-Compatible finite elements provide a route to stable, consistent
+Compatible finite elements %!TB methods?
+provide a route to stable, consistent
 discretisations of the partial differential equations governing
-geophysical fluids which also have the correct wave propagation speeds
+geophysical fluids %!TB cite acta numerica
+%!TB Suggest rewriting and combining the following two sentences:
+which also have the correct wave propagation speeds
 and avoid spurious computational modes, even on non-uniform
-grids. These properties, along with conservation of mass and
+grids.
+These properties, along with conservation of mass and
 potentially other quantities such as energy and enstrophy, are
 considered essential for the accuracy of the dynamical core of weather
-and climate models \citep{staniforth2012horizontal}. Achieving these
+and climate models \citep{staniforth2012horizontal}.
+Achieving these
 properties on a non-orthogonal grid is important because of parallel
-scaling**. While the required properties of the discretisation guide the
+scaling**. %!TB citation and expand on this. Reference Met Office and GungHo
+While the required properties of the discretisation guide the
 choice of finite element spaces, there are many options for both the
 spatial and temporal discretisation and it is not clear which will be
-best for any particular problem. Gusto provides a flexible,
+best for any particular problem.
+%!TB Maybe a bit more direct about what Gusto is and why?
+Gusto provides a flexible,
 extensible, easy-to-use toolkit of options for rapid prototyping of
@@
 density or temperature) on a mesh made of non-overlapping, conforming
-cells, or elements. Within each element, the field is represented as a
+cells, or elements. %!TB: what does conforming mean here? Can we clarify what we mean by element? Do we want to say something about the elements being the same across the mesh?
+Within each element, the field is represented as a
 sum of coefficients multiplied by polynomial basis functions. The
@@
 necessary, or indeed desirable, to use the same finite element
-function space for each field. Using different finite element function
+function space for each field. %!TB: should we distinguish between finite elements and function spaces?
+Using different finite element function
 spaces for each field gives what is known as a mixed finite element
@@
 the de Rham complex. This is a sequence of spaces with differential
-operators between them that map onto the kernal of the next operator
+operators between them that map onto the kernel of the next operator
 in the sequence. This is often summarised in a commutation diagram as
@@
 \pi_{k+1}\diff^{k+1}(\cdot)$. This guarantees that, in 3D, $\nabla$
-maps onto the kernal of $\nabla\times$, ensuring
-$\nabla\times\nabla(\cdot)=0$, and $\nabla\times$ maps onto the kernal
+maps onto the kernel of $\nabla\times$, ensuring
+$\nabla\times\nabla(\cdot)=0$, and $\nabla\times$ maps onto the kernel
 of $\nabla\cdot$, ensuring that $\nabla\cdot\nabla\times(\cdot)=0$;
-i.e. the spaces are compatible. In 2D we have that $\nabla^\perp$ maps
-onto the kernal of $\nabla\cdot$, ensuring that
+i.e.\, the spaces are compatible. In 2D we have that $\nabla^\perp$ maps
+onto the kernel of $\nabla\cdot$, ensuring that
 $\nabla\cdot\nabla^\perp(\cdot)=0$. It is possible to write this de
-Rham complex such that instead we have that $\nabla$ maps onto the
-kernal of $\nabla^\perp$, ensuring that
+Rham complex such that instead $\nabla$ maps onto the
+kernel of $\nabla^\perp$, ensuring that
 $\nabla^\perp\cdot\nabla(\cdot)=0$, but the former is more useful in
 constructing discretisations for geophysical fluid dynamics since it
-enables schemes with local mass conservation \citep{}.
+enables schemes with local mass conservation \citep{}. %!TB: citation
 
@@
 dynamics (**where to say what these are?), a flexible, extensible
-implementation of the methods is strongly desirable. This is
+implementation of the methods is strongly desirable.
+%!TB: should we change this sentence to start with the Met Office link?
+"Because of these promising properties, gungho.
+However there are still many important avenues for research and unanswered questions
+%! - time stepping methods
+%! - solver techniques
+%! - aspects of physics-dynamics coupling
+This is
 especially required because, while it is feasible for someone to write
@@
 investigating different orders or families of finite elements, or
-different formulations of the equation. For geophysical fluid dynamics
+different formulations of the equation.
+%!TB: this sentence felt a bit clunky.
+I guess we want to say that "The research of these questions can often be difficult in operational models, which are generally designed to execute specific configurations very efficiently.
+So the purpose of Gusto is to facilitate research by providing a flexible, extensible framework, in which the implementation of new methods is easier.
+For geophysical fluid dynamics
 applications there are additional considerations relating to the
@@
 several orders of magnitude. (also: boundaries, vertical staggering
-for theta space)
+for theta space) %!TB: this point feels a bit out of place here, should we move it above?
 
@@
 different equations commonly used for modelling geophysical fluids in
-order to demonstrate the similaries between the equations and justify
+order to demonstrate the similarities between the equations and justify
 the use of a unified codebase. The advantages of this unified approach

### 2025-04-04 source/paper.tex
some work towards addressing comments from Tom and starting to fill in the transport section

@@
 
-Compatible finite elements %!TB methods?
-provide a route to stable, consistent
-discretisations of the partial differential equations governing
-geophysical fluids %!TB cite acta numerica
-%!TB Suggest rewriting and combining the following two sentences:
-which also have the correct wave propagation speeds
-and avoid spurious computational modes, even on non-uniform
-grids.
-These properties, along with conservation of mass and
-potentially other quantities such as energy and enstrophy, are
-considered essential for the accuracy of the dynamical core of weather
-and climate models \citep{staniforth2012horizontal}.
-Achieving these
-properties on a non-orthogonal grid is important because of parallel
-scaling**. %!TB citation and expand on this. Reference Met Office and GungHo
-While the required properties of the discretisation guide the
-choice of finite element spaces, there are many options for both the
-spatial and temporal discretisation and it is not clear which will be
-best for any particular problem.
-%!TB Maybe a bit more direct about what Gusto is and why?
-Gusto provides a flexible,
-extensible, easy-to-use toolkit of options for rapid prototyping of
-different spatial and temporal discretisations based on the compatible
-finite element framework applied to geophysical fluid dynamics
-equations relevant to numerical weather and climate prediction. In
-this paper we introduce Gusto to the community, summarise its current
-capabilities and outline our vision for future development.
+Compatible finite element methods provide a route to stable,
+consistent discretisations of the partial differential equations
+governing geophysical fluids \citep{cotter2023compatible}. They
+possess many of the desirable properties described by
+\citep{staniforth2012horizontal} as essential for the accuracy of the
+dynamical core of weather and climate models, such as conservation of
+mass (and potentially other quantities such as energy and enstrophy),
+correct wave propagation speeds and avoidance of spurious
+computational modes, even on non-orthogonal grids. Obtaining these
+properties on a non-orthogonal grid is important because weather and
+climate models achieve optimal performance on supercomputers by
+performing calculations concurrently, or in parallel. In practice,
+this means that the spatial domain is partitioned into subdomains that
+are shared across the processors and data on the boundaries of the
+subdomains needs to be shared between the processors to compute a
+global solution. Current trends in supercomputer architectures
+
+are run on massivly parallel supercomputers that require the co Spatial domain decomposition methods share the computation
+across processors by While the required properties of the
+discretisation guide the choice of finite element spaces, there are
+many options for both the spatial and temporal discretisation and it
+is not clear which will be best for any particular problem. Gusto
+provides a flexible, extensible, easy-to-use toolkit of options for
+rapid prototyping of different spatial and temporal discretisations
+based on the compatible finite element framework applied to
+geophysical fluid dynamics equations relevant to numerical weather and
+climate prediction. In this paper we introduce Gusto to the community,
+summarise its current capabilities and outline our vision for future
+development.
 
@@
 density or temperature) on a mesh made of non-overlapping, conforming
-cells, or elements. %!TB: what does conforming mean here? Can we clarify what we mean by element? Do we want to say something about the elements being the same across the mesh?
-Within each element, the field is represented as a
+cells, or elements. Within each element, the field is represented as a
 sum of coefficients multiplied by polynomial basis functions. The
@@
 necessary, or indeed desirable, to use the same finite element
-function space for each field. %!TB: should we distinguish between finite elements and function spaces?
-Using different finite element function
-spaces for each field gives what is known as a mixed finite element
-discretisation.
+representation for each field. Using different finite element
+representations for each field gives what is known as a mixed finite
+element discretisation.
 
 `Compatible' or `mimetic' finite element discretisations are
-constructed by choosing the different finite element function spaces
+constructed by choosing the different finite element representations
 so that discrete versions of continuous vector calculus identities
@@
 large-scale flow. In addition to this, there is flexibility within the
-compatible finite element framework to choose function spaces that
+compatible finite element framework to choose representations that
 also give other advantages such as energy conservation, local mass
@@
 of $\nabla\cdot$, ensuring that $\nabla\cdot\nabla\times(\cdot)=0$;
-i.e.\, the spaces are compatible. In 2D we have that $\nabla^\perp$ maps
+i.e. the spaces are compatible. In 2D we have that $\nabla^\perp$ maps
 onto the kernel of $\nabla\cdot$, ensuring that
 $\nabla\cdot\nabla^\perp(\cdot)=0$. It is possible to write this de
-Rham complex such that instead $\nabla$ maps onto the
-kernel of $\nabla^\perp$, ensuring that
-$\nabla^\perp\cdot\nabla(\cdot)=0$, but the former is more useful in
-constructing discretisations for geophysical fluid dynamics since it
-enables schemes with local mass conservation \citep{}. %!TB: citation
+Rham complex such that instead $\nabla$ maps onto the kernel of
+$\nabla^\perp$, ensuring that $\nabla^\perp\cdot\nabla(\cdot)=0$, but
+the former is more useful in constructing discretisations for
+geophysical fluid dynamics since it enables schemes with local mass
+conservation \citep{}.
 
@@
 
-Given that there are many promising avenues for research in the area
-of compatible finite element discretisations for geophysical fluid
-dynamics (**where to say what these are?), a flexible, extensible
-implementation of the methods is strongly desirable.
-%!TB: should we change this sentence to start with the Met Office link?
-"Because of these promising properties, gungho.
-However there are still many important avenues for research and unanswered questions
-%! - time stepping methods
-%! - solver techniques
-%! - aspects of physics-dynamics coupling
-This is
-especially required because, while it is feasible for someone to write
-a code implementing a specific finite element discretisation of a PDE,
-the code quickly becomes complicated and unsustainable when
-investigating different orders or families of finite elements, or
-different formulations of the equation.
-%!TB: this sentence felt a bit clunky.
-I guess we want to say that "The research of these questions can often be difficult in operational models, which are generally designed to execute specific configurations very efficiently.
-So the purpose of Gusto is to facilitate research by providing a flexible, extensible framework, in which the implementation of new methods is easier.
-For geophysical fluid dynamics
-applications there are additional considerations relating to the
-geometry and types of waves present in the flow. For example, the
-scales of motion in the horizontal and vertical directions differ by
-several orders of magnitude. (also: boundaries, vertical staggering
-for theta space) %!TB: this point feels a bit out of place here, should we move it above?
+Due to these promising properties, the next Met Office dynamical core
+will use the lowest order compatible finite element discretisation on
+the cubed sphere \citep{}. However, there are many important avenues
+for research and unanswered questions relating to, for example,
+timestepping methods, linear solver techniques and how to couple
+unresolved physics processes to the resolved dynamics. This type of
+research is difficult in operational models, which are generally
+designed to execute specific configurations very efficiently. The
+purpose of Gusto is to facilitate this research by providing an open
+source, flexible, extensible framework
 
@@
 The finite element discretisation is formed by choosing appropriate
-function spaces for the prognostic variables, multiplying by test
-functions from these function spaces and integrating over the
-domain. In addition to this, we integrate by parts when the
-inter-element continuity of the basis functions is not sufficient to
-define the result of applying the differential operator directly.
+representations, or function spaces, for the prognostic variables,
+multiplying by test functions from these function spaces and
+integrating over the domain. In addition to this, we integrate by
+parts when the inter-element continuity of the basis functions is not
+sufficient to define the result of applying the differential operator
+directly.
 
@@
 the elements, the choice of stable transport discretisation is
-different for each prognostic variable.
+different for each prognostic variable. The spatial order of the
+finite element spaces also imposes requirements on the choice of
+transport scheme. In particular, when the lowest order space
+$\mathbb{V}_p$ is chosen to be DG0, we need to use the recovered
+transport scheme described in \citet{bendall2019recovered} to achieve
+second order convergence overall. Here we describe the default
+transport schemes used in Gusto for higher order function spaces and
+refer the reader to \citet{bendall2019recovered} for details of the
+recovered schemes, which are also implemented in Gusto.
+
+These schemes do not preserve the monotonicity of the continuous
+equation, generating spurious over and undershoots. When monotonicity
+is important, for example, to avoid generating spurious and unphysical
+negative values of moisture variables, we can apply a limiter as part
+of the transport scheme.
+
+\JScomment{I think that we should at least mention that we can do horizonal and vertical splitting for stability - maybe don't need to write out all of the forms though!}
 

### 2025-08-04 source/paper.tex
tidy figures
tidy some text

@@
 possess many of the desirable properties described by
-\citep{staniforth2012horizontal} as essential for the accuracy of the
+\citet{staniforth2012horizontal} as essential for the accuracy of the
 dynamical core of weather and climate models, such as conservation of
@@
 climate models achieve optimal performance on supercomputers by
-performing calculations concurrently, or in parallel. In practice,
-this means that the spatial domain is partitioned into subdomains that
-are shared across the processors and data on the boundaries of the
-subdomains needs to be shared between the processors to compute a
-global solution. Current trends in supercomputer architectures
-
-are run on massivly parallel supercomputers that require the co Spatial domain decomposition methods share the computation
-across processors by While the required properties of the
-discretisation guide the choice of finite element spaces, there are
-many options for both the spatial and temporal discretisation and it
-is not clear which will be best for any particular problem. Gusto
-provides a flexible, extensible, easy-to-use toolkit of options for
-rapid prototyping of different spatial and temporal discretisations
-based on the compatible finite element framework applied to
-geophysical fluid dynamics equations relevant to numerical weather and
-climate prediction. In this paper we introduce Gusto to the community,
-summarise its current capabilities and outline our vision for future
-development.
+performing calculations concurrently, or in parallel, and orthogonal
+grids of a spherical manifold will necessarily have singular points
+that lead to a degredation in parallel scaling. An example of this is
+the commonly-used latitude-longitude grid where the singular points
+are the poles.
+
+% In practice, this means that the spatial domain is partitioned into
+% subdomains that are shared across the processors and data on the
+% boundaries of the subdomains needs to be shared between the
+% processors to compute a global solution. Current trends in
+% supercomputer architectures
+% are run on massivly parallel supercomputers that require the co
+% Spatial domain decomposition methods share the computation
+% across processors by
+
+While the required properties of the discretisation guide the choice
+of finite element spaces, there are many options for both the spatial
+and temporal discretisation and it is not clear which will be best for
+any particular problem. Gusto provides a flexible, extensible,
+easy-to-use toolkit of options for rapid prototyping of different
+spatial and temporal discretisations based on the compatible finite
+element framework applied to geophysical fluid dynamics equations
+relevant to numerical weather and climate prediction. In this paper we
+introduce Gusto to the community, summarise its current capabilities
+and outline our vision for future development.
 
@@
 \begin{figure}\label{fig:2D de Rham}
-  \centering{\includegraphics{de_rham_fig_2d.pdf}}
+  \centering{\includegraphics{figures/de_rham_fig_2d.pdf}}
 \end{figure}
@@
 \begin{figure}\label{fig:3D de Rham}
-  \centering{\includegraphics{de_rham_fig_3d.pdf}}
+  \centering{\includegraphics{figures/de_rham_fig_3d.pdf}}
 \end{figure}
@@
 \begin{figure}
-  \centering{\includegraphics[width=0.8\textwidth]{gusto_schematic.png}}
+  \centering{\includegraphics[width=0.8\textwidth]{figures/gusto_schematic.png}}
 \end{figure}

### 2025-08-04 source/paper.tex
more work on intro

@@
 in the sequence. This is often summarised in a commutation diagram as
-in figures \ref{fig:2D de Rham} and \ref{fig:3D de Rham}. In these
+in figures \ref{fig:2DdeRham} and \ref{fig:3DdeRham}. In these
 diagrams the infinite dimensional spaces are along the top row and the
@@
 of $\nabla\cdot$, ensuring that $\nabla\cdot\nabla\times(\cdot)=0$;
-i.e. the spaces are compatible. In 2D we have that $\nabla^\perp$ maps
-onto the kernel of $\nabla\cdot$, ensuring that
-$\nabla\cdot\nabla^\perp(\cdot)=0$. It is possible to write this de
-Rham complex such that instead $\nabla$ maps onto the kernel of
-$\nabla^\perp$, ensuring that $\nabla^\perp\cdot\nabla(\cdot)=0$, but
-the former is more useful in constructing discretisations for
-geophysical fluid dynamics since it enables schemes with local mass
-conservation \citep{}.
-
-\begin{figure}\label{fig:2D de Rham}
-  \centering{\includegraphics{figures/de_rham_fig_2d.pdf}}
+i.e. the spaces are compatible. In 2D we have that $\nabla^\perp$
+(the 2D curl operator) maps onto the kernel of $\nabla\cdot$,
+ensuring that $\nabla\cdot\nabla^\perp(\cdot)=0$. It is possible to
+write this de Rham complex such that instead $\nabla$ maps onto the
+kernel of $\nabla^\perp$, ensuring that
+$\nabla^\perp\cdot\nabla(\cdot)=0$, but the former is more useful in
+constructing discretisations for geophysical fluid dynamics since it
+enables schemes with local mass conservation \citep{}.
+
+\begin{figure}
+  \includegraphics{figures/de_rham_fig_2d.pdf}
+  \caption{De Rham complex for two dimensional spaces. The
+    interelement continuity of the functions decreases from left to
+    right, with functions in $\mathbb{V}^{H^1}$ being fully
+    continuous, those in $\mathbb{V}^{HDiv}$ having continuous normal
+    components and those in $\mathbb{V}^{L^2}$ being fully
+  discontinuous. The polynomial order also decreases from left to
+  right.}
+  \label{fig:2DdeRham}
 \end{figure}
 
-\begin{figure}\label{fig:3D de Rham}
-  \centering{\includegraphics{figures/de_rham_fig_3d.pdf}}
+\begin{figure}
+  \includegraphics{figures/de_rham_fig_3d.pdf}
+  \caption{De Rham complex for three dimensional spaces. The
+    interelement continuity of the functions decreases from left to
+    right, with functions in $\mathbb{V}^{H^1}$ being fully
+    continuous, those in $\mathbb{V}^{HCurl}$ having continuous
+    tangential components, those in $\mathbb{V}^{HDiv}$ having
+    continuous normal components and those in $\mathbb{V}^{L^2}$ being
+    fully discontinuous. The polynomial order also decreases from left
+    to right.}
+  \label{fig:3DdeRham}
 \end{figure}
@@
 the cubed sphere \citep{}. However, there are many important avenues
-for research and unanswered questions relating to, for example,
-timestepping methods, linear solver techniques and how to couple
-unresolved physics processes to the resolved dynamics. This type of
-research is difficult in operational models, which are generally
-designed to execute specific configurations very efficiently. The
-purpose of Gusto is to facilitate this research by providing an open
-source, flexible, extensible framework
+for research and unanswered questions relating to, for example, the
+choice of grid and $\mathbb{V}^{HDiv}_h$; the design of transport
+schemes; stable and efficient timestepping algorithms; linear solver
+techniques and how to couple unresolved physics processes to the
+resolved dynamics. This type of research is difficult in operational
+models, which are generally designed to execute specific
+configurations very efficiently. The purpose of Gusto is to facilitate
+this research by providing an open source, flexible, extensible
+framework that provides the key components required when designing
+algorithms for a dynamical core. These include: finite element
+function spaces defined on the sphere (where the Piola transform is
+required to... **); meshes that have a columnar structure (because**),
+stable transport and diffusion schemes for variables with different
+interelement continuity,
 
@@
 modelling. In section \ref{sec: governing} we will outline three
-different equations commonly used for modelling geophysical fluids in
-order to demonstrate the similarities between the equations and justify
-the use of a unified codebase. The advantages of this unified approach
-become even clearer in section \ref{sec: FEM} where we present the
-compatible finite element discretisation of the different types of
-terms (e.g. pressure gradient, transport, diffusion) seen in section
-\ref{sec: governing}. In section \ref{sec: design} we describe the
-code structure of Gusto followed by a detailed example (section
+different equation sets commonly used for modelling geophysical fluids
+in order to demonstrate the similarities between the equation sets and
+justify the use of a unified codebase. The advantages of this unified
+approach become even clearer in section \ref{sec: FEM} where we
+present the compatible finite element discretisation of the different
+types of terms (e.g. pressure gradient, transport, diffusion) seen in
+section \ref{sec: governing}. In section \ref{sec: design} we describe
+the code structure of Gusto followed by a detailed example (section
 \ref{sec: FML}) of the Form Manipulation Language (FML) that
@@
 of numerical results that showcases the range of discretisations
-available within Gusto. **mention here new functionality facilitated
-by the flexibility of Gusto? We summarise our current status and
-future directions in section \ref{sec: summary}.
+available within Gusto. We summarise our current status and future
+directions in section \ref{sec: summary}.
 
@@
 equations in terms of the prognostic variables $\MM{u}$, $b$ and $p$
-where $\MM{u}$ is the velocity field, $b$ is the fluid buoyancy and
-$p$ is pressure. In terms of these variables, the equations are:
+where $\MM{u}$ is the velocity field, $p$ is pressure and $b$ is the
+fluid buoyancy. In terms of these variables, the equations are:
 
@@
 $\hat{\MM{k}}$ is the unit vector pointing upwards and $c$ is the
-acoustic wave speed. The equations have been written in
-three-dimensional form but they can also be solved in a
-two-dimensional vertical slice domain described by $(x, z)$
-coordinates. In this case the velocity $\MM{u}$ can either be
-two-dimensional as well (if there is no Coriolis force), or fully
-three-dimensional but with each cartensian component only varying in
-the $x-z$ plane. In the latter case this is implemented using a
-three-dimensional computational domain that is only one element wide
+acoustic wave speed. The equations describe the motion of a
+three dimensional stratified fluid but can also be solved in a
+two dimensional vertical slice domain described by $(x, z)$
+coordinates. In this case, the velocity $\MM{u}$ can either be
+two dimensional as well (if there is no Coriolis force), or fully
+three dimensional but with each Cartesian component only varying in
+the $x-z$ plane. In the latter case, this is implemented using a
+three dimensional computational domain that is only one element wide
 in the $y$-direction and with periodic boundary conditions also in
@@
 is the gas constant, $c_p$ is the specific heat at constant pressure
-and $\kappa=R/c_p$. The equations have been written in
-three-dimensional form but they can also be solved in a
-two-dimensional vertical slice domain described by $(x, z)$
-coordinates. In this case the velocity $\MM{u}$ can either be two
+and $\kappa=R/c_p$. Again, these equations describe the motion of a
+three dimensional stratified fluid but can also be solved in a
+two dimensional vertical slice domain described by $(x, z)$
+coordinates. In this case, the velocity $\MM{u}$ can either be two
 dimensional as well (if there is no Coriolis force), or fully three
-dimensional but with each cartensian component only varying in the
-$x-z$ plane. In the latter case this is implemented using a three
+dimensional but with each Cartesian component only varying in the
+$x-z$ plane. In the latter case, this is implemented using a three
 dimensional computational domain that is only one element wide in the

### 2025-08-04 source/paper.tex
proper terminator toy

--- 2025-04-04 source/paper.tex
@@
 
-Compatible finite element methods provide a route to stable,
-consistent discretisations of the partial differential equations
-governing geophysical fluids \citep{cotter2023compatible}. They
-possess many of the desirable properties described by
-\citep{staniforth2012horizontal} as essential for the accuracy of the
-dynamical core of weather and climate models, such as conservation of
-mass (and potentially other quantities such as energy and enstrophy),
-correct wave propagation speeds and avoidance of spurious
-computational modes, even on non-orthogonal grids. Obtaining these
-properties on a non-orthogonal grid is important because weather and
-climate models achieve optimal performance on supercomputers by
-performing calculations concurrently, or in parallel. In practice,
-this means that the spatial domain is partitioned into subdomains that
-are shared across the processors and data on the boundaries of the
-subdomains needs to be shared between the processors to compute a
-global solution. Current trends in supercomputer architectures
-
-are run on massivly parallel supercomputers that require the co Spatial domain decomposition methods share the computation
-across processors by While the required properties of the
-discretisation guide the choice of finite element spaces, there are
-many options for both the spatial and temporal discretisation and it
-is not clear which will be best for any particular problem. Gusto
-provides a flexible, extensible, easy-to-use toolkit of options for
-rapid prototyping of different spatial and temporal discretisations
-based on the compatible finite element framework applied to
-geophysical fluid dynamics equations relevant to numerical weather and
-climate prediction. In this paper we introduce Gusto to the community,
-summarise its current capabilities and outline our vision for future
-development.
+Compatible finite elements provide a route to stable, consistent
+discretisations of the partial differential equations governing
+geophysical fluids which also have the correct wave propagation speeds
+and avoid spurious computational modes, even on non-uniform
+grids. These properties, along with conservation of mass and
+potentially other quantities such as energy and enstrophy, are
+considered essential for the accuracy of the dynamical core of weather
+and climate models \citep{staniforth2012horizontal}. Achieving these
+properties on a non-orthogonal grid is important because of parallel
+scaling**. While the required properties of the discretisation guide the
+choice of finite element spaces, there are many options for both the
+spatial and temporal discretisation and it is not clear which will be
+best for any particular problem. Gusto provides a flexible,
+extensible, easy-to-use toolkit of options for rapid prototyping of
+different spatial and temporal discretisations based on the compatible
+finite element framework applied to geophysical fluid dynamics
+equations relevant to numerical weather and climate prediction. In
+this paper we introduce Gusto to the community, summarise its current
+capabilities and outline our vision for future development.
 
@@
 necessary, or indeed desirable, to use the same finite element
-representation for each field. Using different finite element
-representations for each field gives what is known as a mixed finite
-element discretisation.
+function space for each field. Using different finite element function
+spaces for each field gives what is known as a mixed finite element
+discretisation.
 
 `Compatible' or `mimetic' finite element discretisations are
-constructed by choosing the different finite element representations
+constructed by choosing the different finite element function spaces
 so that discrete versions of continuous vector calculus identities
@@
 large-scale flow. In addition to this, there is flexibility within the
-compatible finite element framework to choose representations that
+compatible finite element framework to choose function spaces that
 also give other advantages such as energy conservation, local mass
@@
 the de Rham complex. This is a sequence of spaces with differential
-operators between them that map onto the kernel of the next operator
+operators between them that map onto the kernal of the next operator
 in the sequence. This is often summarised in a commutation diagram as
@@
 \pi_{k+1}\diff^{k+1}(\cdot)$. This guarantees that, in 3D, $\nabla$
-maps onto the kernel of $\nabla\times$, ensuring
-$\nabla\times\nabla(\cdot)=0$, and $\nabla\times$ maps onto the kernel
+maps onto the kernal of $\nabla\times$, ensuring
+$\nabla\times\nabla(\cdot)=0$, and $\nabla\times$ maps onto the kernal
 of $\nabla\cdot$, ensuring that $\nabla\cdot\nabla\times(\cdot)=0$;
 i.e. the spaces are compatible. In 2D we have that $\nabla^\perp$ maps
-onto the kernel of $\nabla\cdot$, ensuring that
+onto the kernal of $\nabla\cdot$, ensuring that
 $\nabla\cdot\nabla^\perp(\cdot)=0$. It is possible to write this de
-Rham complex such that instead $\nabla$ maps onto the kernel of
-$\nabla^\perp$, ensuring that $\nabla^\perp\cdot\nabla(\cdot)=0$, but
-the former is more useful in constructing discretisations for
-geophysical fluid dynamics since it enables schemes with local mass
-conservation \citep{}.
+Rham complex such that instead we have that $\nabla$ maps onto the
+kernal of $\nabla^\perp$, ensuring that
+$\nabla^\perp\cdot\nabla(\cdot)=0$, but the former is more useful in
+constructing discretisations for geophysical fluid dynamics since it
+enables schemes with local mass conservation \citep{}.
 
@@
 
-Due to these promising properties, the next Met Office dynamical core
-will use the lowest order compatible finite element discretisation on
-the cubed sphere \citep{}. However, there are many important avenues
-for research and unanswered questions relating to, for example,
-timestepping methods, linear solver techniques and how to couple
-unresolved physics processes to the resolved dynamics. This type of
-research is difficult in operational models, which are generally
-designed to execute specific configurations very efficiently. The
-purpose of Gusto is to facilitate this research by providing an open
-source, flexible, extensible framework
+Given that there are many promising avenues for research in the area
+of compatible finite element discretisations for geophysical fluid
+dynamics (**where to say what these are?), a flexible, extensible
+implementation of the methods is strongly desirable. This is
+especially required because, while it is feasible for someone to write
+a code implementing a specific finite element discretisation of a PDE,
+the code quickly becomes complicated and unsustainable when
+investigating different orders or families of finite elements, or
+different formulations of the equation. For geophysical fluid dynamics
+applications there are additional considerations relating to the
+geometry and types of waves present in the flow. For example, the
+scales of motion in the horizontal and vertical directions differ by
+several orders of magnitude. (also: boundaries, vertical staggering
+for theta space)
 
@@
 different equations commonly used for modelling geophysical fluids in
-order to demonstrate the similarities between the equations and justify
+order to demonstrate the similaries between the equations and justify
 the use of a unified codebase. The advantages of this unified approach
@@
 The finite element discretisation is formed by choosing appropriate
-representations, or function spaces, for the prognostic variables,
-multiplying by test functions from these function spaces and
-integrating over the domain. In addition to this, we integrate by
-parts when the inter-element continuity of the basis functions is not
-sufficient to define the result of applying the differential operator
-directly.
+function spaces for the prognostic variables, multiplying by test
+functions from these function spaces and integrating over the
+domain. In addition to this, we integrate by parts when the
+inter-element continuity of the basis functions is not sufficient to
+define the result of applying the differential operator directly.
 
@@
 the elements, the choice of stable transport discretisation is
-different for each prognostic variable. The spatial order of the
-finite element spaces also imposes requirements on the choice of
-transport scheme. In particular, when the lowest order space
-$\mathbb{V}_p$ is chosen to be DG0, we need to use the recovered
-transport scheme described in \citet{bendall2019recovered} to achieve
-second order convergence overall. Here we describe the default
-transport schemes used in Gusto for higher order function spaces and
-refer the reader to \citet{bendall2019recovered} for details of the
-recovered schemes, which are also implemented in Gusto.
-
-These schemes do not preserve the monotonicity of the continuous
-equation, generating spurious over and undershoots. When monotonicity
-is important, for example, to avoid generating spurious and unphysical
-negative values of moisture variables, we can apply a limiter as part
-of the transport scheme.
-
-\JScomment{I think that we should at least mention that we can do horizonal and vertical splitting for stability - maybe don't need to write out all of the forms though!}
+different for each prognostic variable.
 
@@
 The aim of this section is to present a selection of results that
-demonstrate the tools available in Gusto, such as the choice of grids,
+showcase the tools available in Gusto, such as the choice of grids,
 HDiv element families, order of finite element spaces, different
-spatial discretisation options, timestepping schemes and physics. The
-numerical properties of these configurations (convergence, stability,
-accuracy and efficiency) have either been demonstrated elsewhere, or
-will be investigated more thoroughly in future publications.
+spatial discretisation options and timestepping schemes.
+As these options can be combined in so many ways, it is only possible to present results from a tiny fraction configurations.
+We have chosen to use a selection from the hierarchy of standard test cases that span the stages of development of dynamical cores.
+Although we do include some results here that demonstrate the numerical properties of certain configurations, the purpose of this section is not to provide a rigorous verification of any particular choice of discretisation.
+Such results are demonstrated elsewhere, for instance \cite{bendall2020compatible} or \cite{hartney2025exploring}.
+The settings used for each result from this section are detailed in Appendix \ref{sec:appendix}.
 
@@
 
-\subsection{Terminator `toy' chemistry}
-This test combines transport and a `toy' chemistry model to
-demonstrate physics-dynamics coupling in Gusto (where here chemistry
-is the `physics').
+\subsection{Terminator Toy chemistry}
+The first results use the ``Terminator Toy'' test of \cite{lauritzen2015terminator}, which involves the transport of two coupled tracers $X$ and $X_2$ on the surface of a sphere.
+The chemicals interact such that the sum $2m_X+m_{X_2}$ should be preserved, where $m_X$ and $m_{X_2}$ are the mixing ratios of the individual chemicals.
+When positivity or monotonicity limiters are applied to the numerical transport scheme, this may break the correlation between the two species which will be highlighted by this test.
+This test is also a demonstration of simple ``physics-dynamics coupling'', as the conversion term between the species can be handled separately from the transport terms.
 
@@
 
-\subsubsection{Held Suarez}
-
 \section{Summary}
@@
 \appendix
-\section{}    %% Appendix A
+\section{Test Case Settings}  \label{sec:appendix}  %% Appendix A
 

### 2025-08-04 results/plotting/plot_galewsky_jet.py

@@ -26,7 +26,7 @@ plot_stem = f'{abspath(dirname(__file__))}/../figures/galewsky_jet'
 # ---------------------------------------------------------------------------- #
 # Plot details
 # ---------------------------------------------------------------------------- #
-subplots_x = 2
+subplots_x = 1
 subplots_y = 2
 field_name = 'RelativeVorticity'
 time_idx = -1

### 2025-08-04 results/plotting/plot_terminator_toy.py

@@ -11,24 +11,24 @@ from tomplot import (
     tomplot_field_title, extract_gusto_coords, extract_gusto_field,
 )
 
-test = 'terminator_toy_rf4_dt450'
+test = 'terminator_toy'
 
 # ---------------------------------------------------------------------------- #
 # Directory for results and plots
 # ---------------------------------------------------------------------------- #
 # When copying this example these paths need editing, which will usually involve
 # removing the abspath part to set directory paths relative to this file
-results_file_name = f'{abspath(dirname(__file__))}/{test}/field_output.nc'
-plot_stem = f'{abspath(dirname(__file__))}'
+results_file_name = f'{abspath(dirname(__file__))}/../results/{test}/field_output.nc'
+plot_stem = f'{abspath(dirname(__file__))}/../figures/{test}'
 
 # ---------------------------------------------------------------------------- #
 # Plot details
 # ---------------------------------------------------------------------------- #
 subplots_x = 2
 subplots_y = 3
-field_names = ['X', 'X',
-               'X2', 'X2',
-               'X_plus_X2_plus_X2', 'X_plus_X2_plus_X2']
+field_names = ['Y', 'Y',
+               'Y2', 'Y2',
+               'Y_plus_Y2_plus_Y2', 'Y_plus_Y2_plus_Y2']
 time_idxs = [0, -1,
              0, -1,
              0, -1]
@@ -44,16 +44,16 @@ projection = ccrs.Robinson()
 
 Xt_contours = np.linspace(-1e-13, 1e-13, 11)
 Xt_colour_scheme = 'BrBG'
-Xt_field_label = r'$X_T - X_T^0$ (kg kg$^{-1}$)'
+Xt_field_label = r'$m_{X_T} - m_{X_T^0}$ (kg kg$^{-1}$)'
 
 X_contours = np.linspace(0.0, 4.e-6, 15)
 X_colour_scheme = 'PuRd'
-X_field_label = r'$X$ (kg kg$^{-1}$)'
+X_field_label = r'$m_X$ (kg kg$^{-1}$)'
 
 X2_contours_first = np.linspace(0.0, 2.000000000001e-6, 15)
 X2_contours = np.linspace(0.0, 2.0e-6, 15)
 X2_colour_scheme = 'PuBu'
-X2_field_label = r'$X_2$ (kg kg$^{-1}$)'
+X2_field_label = r'$m_{X_2}$ (kg kg$^{-1}$)'
 
 contour_method = 'tricontour'
 
[…]

### 2025-08-04 results/test_case_scripts/terminator_toy.py

@@ -70,13 +70,13 @@ def terminator_toy(
 
     # Define the two species as tracers
     X = ActiveTracer(
-        name='X', space='DG',
+        name='Y', space='DG',
         variable_type=TracerVariableType.mixing_ratio,
         transport_eqn=TransportEquationType.advective
     )
 
     X2 = ActiveTracer(
-        name='X2', space='DG',
+        name='Y2', space='DG',
         variable_type=TracerVariableType.mixing_ratio,
         transport_eqn=TransportEquationType.advective
     )
@@ -93,8 +93,8 @@ def terminator_toy(
     )
 
     # Define intermediate sums to be able to use the TracerDensity diagnostic
-    X_plus_X2 = Sum('X', 'X2')
-    X_plus_X2_plus_X2 = Sum('X_plus_X2', 'X2')
+    X_plus_X2 = Sum('Y', 'Y2')
+    X_plus_X2_plus_X2 = Sum('Y_plus_Y2', 'Y2')
     diagnostic_fields = [X_plus_X2, X_plus_X2_plus_X2]
     io = IO(domain, output, diagnostic_fields=diagnostic_fields)
 
@@ -102,14 +102,14 @@ def terminator_toy(
     # Define limiters for the interacting species
     limiter_space = domain.spaces('DG')
     sublimiters = {
-        'X': DG1Limiter(limiter_space),
-        'X2': DG1Limiter(limiter_space)
+        'Y': DG1Limiter(limiter_space),
+        'Y2': DG1Limiter(limiter_space)
     }
     MixedLimiter = MixedFSLimiter(eqn, sublimiters)
 
     transport_scheme = SSPRK3(domain, limiter=MixedLimiter)
     transport_method = [
-        DGUpwind(eqn, 'X'), DGUpwind(eqn, 'X2')
+        DGUpwind(eqn, 'Y'), DGUpwind(eqn, 'Y2')
     ]
 
     # Physics scheme --------------------------------------------------------- #
@@ -123,7 +123,7 @@ def terminator_toy(
     terminator_stepper = BackwardEuler(domain)
 
     physics_schemes = [
-        (TerminatorToy(eqn, k1=k1, k2=k2, species1_name='X', species2_name='X2'),
+        (TerminatorToy(eqn, k1=k1, k2=k2, species1_name='Y', species2_name='Y2'),
          terminator_stepper)
     ]
 
@@ -173,8 +173,8 @@ def terminator_toy(
[…]

### 2025-08-04 source/paper.tex
edits and make advection diffusion example P1

--- 2025-08-04 source/paper.tex
@@
 
-Comparing the equation sets given in the previous sections, we see
+Comparing the equation sets given in the previous section, we see
 several similarities. For example, the prognostic equation for the
@@
 integrating over the domain. In addition to this, we integrate by
-parts when the inter-element continuity of the basis functions is not
+parts when the interelement continuity of the basis functions is not
 sufficient to define the result of applying the differential operator
@@
 
-In line with the theory outlined in section \ref{}, we always choose
-the velocity $\MM{u} \in \mathbb{V}_u$ with $\mathbb{V}_u \subset
-H(\text{div})$. The field whose gradient appears in the `pressure'
-gradient term is chosen to be in the space $\mathbb{V}_p$ with
-$\mathbb{V}_p \subset L_2$, or, in the case of the more complex
-nonlinear pressure gradient term for the compressible Euler equations,
-it is a function of a variable in this space. The compressible
-Boussinesq and compressible Euler equations have a third prognostic
-variable, either buoyancy or potential temperature, which is chosen to
-be in the space $\mathbb{V}_\theta$ which has degrees of freedom that
-are collocated with those of the vertical velocity, mimicking the
-Charney-Phillips staggering as discussed in **. These choices are
-governed by the properties required of discretisation of the
-linearised equation sets (as discussed in section ** and other
-refs). To solve the nonlinear equation sets we need to introduce
-stable and accurate discretisations of the transport terms and the
-nonlinear pressure gradient term in the compressible Euler equation
-set. In addition, some test cases require viscous and diffusive terms
-to generate converged solutions \ref{}. In the following subsections
-we illustrate a non-exhaustive (**probably!) selection of the options
-available in Gusto.
+In line with the theory outlined in the introduction, we always choose
+the velocity $\MM{u} \in \mathbb{V}^{HDiv}$ with $\mathbb{V}^{HDiv}
+\subset H(\text{div})$. **mention bcs here! The field whose gradient
+appears in the `pressure' gradient term is chosen to be in the space
+$\mathbb{V}^{L^2}$ with $\mathbb{V}^{L^2} \subset L_2$, or, in the
+case of the more complex nonlinear pressure gradient term for the
+compressible Euler equations, it is a function of a variable in this
+space. The compressible Boussinesq and compressible Euler equations
+have a third prognostic variable, either buoyancy or potential
+temperature, which is chosen to be in the space $\mathbb{V}_\theta$
+which has degrees of freedom that are collocated with those of the
+vertical velocity, mimicking the Charney-Phillips staggering as
+discussed in **. These choices are governed by the properties required
+of discretisation of the linearised equation sets (as discussed in
+section ** and other refs). To solve the nonlinear equation sets we
+need to introduce stable and accurate discretisations of the transport
+terms and the nonlinear pressure gradient term in the compressible
+Euler equation set. In addition, some test cases require viscous and
+diffusive terms to generate converged solutions \ref{}. In the
+following subsections we illustrate a non-exhaustive selection of the
+options available in Gusto.
 
@@
 the pressure (or depth, for shallow water) field is discretised in
-$\mathbb{V}_p \subset L_2$, meaning that it is discontinuous between
+$\mathbb{V}^{L^2} \subset L_2$, meaning that it is discontinuous between
 cells, the gradient is not defined globally and the term must be
@@
 element facets and on exterior boundaries we set no normal flow
-boundary conditions. However, the pressure gradient term in the
+boundary conditions**. However, the pressure gradient term in the
 compressible Euler equations, $c_p\theta\nabla\Pi$, is nonlinear and
@@
 transfers the differential operator to $\MM{w}\theta$ which is
-discontinuous in the horizontal. Following \ref{} we integrate by parts
+discontinuous in the horizontal direction. Following \ref{} we integrate by parts
 in each element $e$, giving
@@
 where $\Gamma_v$ denotes the vertical element facets, across which
-$\theta$ is discontinuous. Here we have chosen, as in \citet{}, to take
-the average value of $\Pi$ across the facets, $\avg{\Pi} =
-(\Pi^++\Pi^-)/2$, where the $+$ and $-$ indicate each side of the
-facet. The brackets $\jump{q} = q^+-q^-$ denote the jump in $q$
-across the facet and the integral over the horizontal facets vanishes
-since both $\theta$ and $\MM{w}\cdot\MM{n}$ are continuous across
-these facets so the jump is zero.
+$\theta$, and therefore $\Pi$, are discontinuous. Here we have chosen,
+as in \citet{}, to take the average value of $\Pi$ across the facets,
+$\avg{\Pi} = (\Pi^++\Pi^-)/2$, where the $+$ and $-$ indicate each
+side of the facet. The brackets $\jump{q} = q^+-q^-$ denote the jump
+in $q$ across the facet and the integral over the horizontal facets
+vanishes since both $\theta$ and $\MM{w}\cdot\MM{n}$ are continuous
+across these facets so the jump is zero.
 
@@
 diffusive terms for stability, but sometimes they are required to
-produce converged solutions, ie. so that refining the grid does not
+produce converged solutions so that refining the grid does not
 lead to smaller and smaller scale motions. Here we present the symmetric
@@
 be applied to each field. ** something about SIQN structure. Terms
-related to the physics are solved for in the appropriate part of the
-iteration loop, with physics processes that occur on `fast' timescales
-evaluated more frequently and `slower' physics processes evaluated
-just once at the end of the timestep. The choice of where to evaluate
-the physics processes in the iteration is also based on the
-computational cost and the best way to do this is still an open
-research question (**ref).
+involved in physics parameterisations are evaluated in the appropriate
+part of the iteration loop, with physics processes that occur on
+`fast' timescales evaluated more frequently and `slower' physics
+processes evaluated just once at the end of the timestep. The choice
+of where to evaluate the physics processes in the iteration is also
+based on the computational cost and the best way to do this is still
+an open research question (**ref).
 
@@
 will describe the objects required to define labels and filters before
-illustrating this with a simple example(s?).
+illustrating this with some simple examples.
 
@@
 
-\subsection{Example}
+\subsection{Example 1: IMEX-Euler applied to the 1d advection diffuction equation}
 
@@
 compatible finite element discretisation, we use the space of piecewise
-quadratic continuous Lagrange polynomials, denoted by $P_2$, for the
+linear continuous Lagrange polynomials, denoted by $P_1$, for the
 finite element discretisation as this greatly simplifies the
@@
 \begin{equation}\label{eq:weak-advdiff}
-  \int_\Omega vq_t \diff x + c \int_\Omega v q_x \diff x + \kappa \int_\Omega v_x q_x \diff x = 0, \quad \forall v \in P_2,
+  \int_\Omega vq_t \diff x + c \int_\Omega v q_x \diff x + \kappa \int_\Omega v_x q_x \diff x = 0, \quad \forall v \in P_1,
 \end{equation}
-where $v\in P_2$ is the test function and the diffusion term has been
+where $v\in P_1$ is the test function and the diffusion term has been
 integrated by parts.
@@
 Listing \ref{lst:define-residual} presents the code to define this
-spatial discretisation of the residual with the $P_2$ function space
+spatial discretisation of the residual with the $P_1$ function space
 denoted by \texttt{V}. Here we omit the code that sets up the mesh,
@@
 exploit information stored in the label value. We use this in two
-important situations in Gusto. The first situation is when we require
-a linearisation of the residual, for example when applying a Newton
-method to the full nonlinear equations \JScomment{should we clarify
-  that we mean a specific linearisation because Firedrake can do
-  Newton methods without FML?}, or to test a new algorithm or model
-setup on linear equations. The second situation is when we need to
-evaluate source terms that come from physics schemes. We can
-illustrate both situations with simple extensions to the above
-example.
-
-First consider the 1D viscous Burgers' equation
-\begin{equation}
-  u_t + uu_x + \nu u_{xx} = 0.
-\end{equation}
+important situations in Gusto. The first situation is when we need to
+evaluate source terms that come from physics schemes. The second
+situation is when we require a linearisation of the residual, for
+example when applying a quasi-Newton method to solve the full
+nonlinear equations or when testing a new algorithm on the linear
+equations. We illustrate both situations in the following two
+examples.
+
+\subsection{Example 2: Sequential splitting for an idealised chemistry-transport problem}
+
+\subsection{Example 3: Quasi-Newton }
 

### 2025-08-04 source/paper.tex
first draft of results section (without all results)

--- 2025-08-04 source/paper.tex
@@
 
-To include:
-\begin{itemize}
-\item terminator toy (Tim)
-\item shallow water on different grids and with different HDiv
-  families (Nell) e.g. Williamson 5 cubed sphere PV and Galewsky
-  icosahedral sphere PV, high res (ref 6)
-\item convergence with different RK schemes / multistage schemes using
-  Williamson 2 or 6, L2 and Linf errors for D and u (2 plots, one L2,
-  one Linf) (Alex)
-\item vertical slice Euler - Straka, different orders, illustrates
-  recovery for diffusion (Tom)
-\item 3D moist bubble with rain (Tom/Alex)
-\item spherical gravity wave, Boussinesq (Jemma)
-\item spherical baroclinic wave, Euler - whatever people usually plot
-  (horizontal slice pressure contours, temperature shading) (Dan)
-\item Held Suarez??
-\end{itemize}
-
 \subsection{Terminator Toy chemistry}
 The first results use the ``Terminator Toy'' test of \cite{lauritzen2015terminator}, which involves the transport of two coupled tracers $X$ and $X_2$ on the surface of a sphere.
-The chemicals interact such that the sum $2m_X+m_{X_2}$ should be preserved, where $m_X$ and $m_{X_2}$ are the mixing ratios of the individual chemicals.
+The chemicals interact such that the sum $m_T=m_X+2m_{X_2}$ should be preserved, where $m_X$ and $m_{X_2}$ are the mixing ratios of the individual chemicals.
 When positivity or monotonicity limiters are applied to the numerical transport scheme, this may break the correlation between the two species which will be highlighted by this test.
-This test is also a demonstration of simple ``physics-dynamics coupling'', as the conversion term between the species can be handled separately from the transport terms.
+This test is also a demonstration of simple ``physics-dynamics coupling'', as the conversion term between the species can be handled separately from the transport terms. \\
+\\
+For this test, we use the DG upwind scheme described in Section \ref{sec:transport} to transport the species on an icosahedral sphere mesh, and apply the vertex-based limiter of \cite{kuzmin2010vertex}.
+The term described chemical reaction is separated from the using a simple sequential splitting, \textcolor{red}{as described in Section \?}. \\
+\\
+Figure \ref{fig:terminator} shows the mixing ratios of the two individual species at the initial time and following a single rotation around the sphere by a deformational flow.
+The change in the sum of the species is also displayed, with differences from zero remaining close to machine precision, which demonstrates that this transport scheme does preserve the correlation between the tracers.
+
+\begin{figure}[htp!]
+\centering
+\includegraphics[width=0.9\textwidth]{terminator_toy.png}
+\caption{The evolution of the chemical species from the ``Terminator Toy'' test of \cite{lauritzen2015terminator}.
+The fields at the initial time are shown on the left, and at the final time ($t=12$ days) which follows a single revolution around the sphere by a deformational flow.
+The bottom row shows the change in the total $m_T=m_X+2m_{X_2}$ from the initial time, which remains very close to zero.}
+\label{fig:terminator}
+\end{figure}
 
@@
 \citet{williamson1992standard} provides a common starting point for
-testing new discretisations. This test suite is usually augmented by
+testing new discretisations. This test suite is often augmented by
 the unstable barotropic jet described in
-\citet{galewsky2004initial}. Here we present convergence results for a
-variety of timestepping schemes applied to the steady state
-geostrophic balanced flow (test case 2 from
-\citet{williamson1992standard}) and a demonstration of two different
-grids and HDiv element families for shallow water flow over a mountain
-(test case 5 from \citet{williamson1992standard}) and the unstable
-barotropic jet \citep{galewsky2004initial}.
-
-\subsubsection{Convergence results}
-
-\subsubsection{Grids and finite element families}
-
-\subsection{Compressible Boussinesq equations}
+\citet{galewsky2004initial}. \\
+\\
+Here we present two sets of convergence results for a variety of configurations measured using the steady state geostrophic balanced flow (test case 2 from
+\citet{williamson1992standard}) with the linear shallow water equations.
+Figure \ref{fig:sw_convergence} shows the spatial convergence for a number of choices of finite element for the prognostic variables, using a Trapezium Rule discretisation (also known as Crank-Nicolson).
+It also shows the temporal convergence for a range of timestepping schemes using the 1st order Raviart-Thomas elements on quadrilateral cells. 
+\textcolor{red}{Once the test has been completed, come back to edit what we say here about convergence!} \\
+\\
+We show results from the unstable jet of \citep{galewsky2004initial} in Figure \ref{fig:galewsky}, using the non-linear shallow water equations.
+These results use the Semi-Implicit Quasi-Newton timestepper described in Section \ref{???}, and compare the RT elements on a cubed-sphere mesh with the BDM elements on an icosahedral-sphere.
+
+\begin{figure}[htp!]
+\centering
+\includegraphics[width=0.9\textwidth]{}
+\caption{Convergence results using the linear shallow water equations with the steady geostrophic flow of \cite{williamson1992standard}.
+(Left) the spatial convergence of a number of orders and families of finite element, all using the same timestepper.
+(Right) the temporal convergence of different timesteppers, all using the same spatial discretisation.}
+\label{fig:sw_convergence}
+\end{figure}
+
+\begin{figure}[htp!]
+\centering
+\includegraphics[width=0.9\textwidth]{galewsky_jet.png}
+\caption{A view of the relative vorticity around the North pole at day 6 of a simulation of the Galewsky jet test case.
+(Left) the RT1 elements on a cubed-sphere mesh.
+(Right) the BDM1 elements on an icosahedral-sphere mesh.
+Both cases use the Semi-Implicit Quasi-Newton timestepper.
+The edges of the panels of the cube and icosahedron are highlighted.}
+\label{fig:galewsky}
+\end{figure}
 
 \subsection{Compressible Euler equations}
-
-\subsubsection{Dense bubble in a vertical slice}
-
-\subsubsection{3D moist rising bubble}
-
-\subsubsection{Baroclinic wave on the sphere}
+We have also implemented many of the standard test cases for the non-hydrostatic compressible Euler equations that are used in modern NWP models.
+In this section we show results from a small selection of these, which demonstrate different domains in which these methods work.
+Throughout this section we use the Semi-Implicit Quasi-Newton timestepper.
+ \\
+\\
+The first test is the falling density current of \cite{straka1993numerical} in a vertical slice. This test involves the addition of diffusion to both the potential temperature and velocity equations.
+Figure \ref{fig:straka} shows the final potential temperature equation for simulations using the lowest-order and next-to-lowest order finite elements on the vertical slice.
+The lowest-order methods require the transport and diffusion terms to use the ``recovered'' scheme of \citet{bendall2019recovered} to attain second-order accuracy. 
+\\
+\begin{figure}[htp!]
+\centering
+\includegraphics[width=0.8\textwidth]{straka.png}
+\caption{The potential temperature field after 900 s from the falling density current test of \cite{straka1993numerical}.
+(Top) with the lowest-order finite elements, and (bottom) with the next-to-lowest-order finite elements.
+}
+\label{fig:straka}
+\end{figure}
+\\
+To demonstrate the use of moist physics parametrisations, we use a moist three-dimensional rising thermal simulation in an extruded doubly-periodic plane.
+The initial conditions and moisture parametrisation are those of \citep{bendall2020compatible}, which is an extension of the dry case of \citep{kelly2012continuous} and \citep{melvin2019mixed}.
+Cross-sections of the initial and final equivalent potential temperature field are shown in Figure \ref{fig:3d_bubble}.
+As in the two-dimensional test of \citep{bryan2002benchmark}, the atmosphere is saturated and cloudy everywhere, so that the moist processes are active in the whole domain.
+\\
+\begin{figure}[htp!]
+\centering
+\includegraphics[width=0.9\textwidth]{moist_bryan_fritsch.png}
+\caption{Vertical cross-sections of the (left) initial and (right) final equivalent potential temperature field from the three-dimensional moist rising thermal test.
+}
+\label{fig:3d_bubble}
+\end{figure}
+\\
+Our final test case is the deep baroclinic wave of \citep{ullrich2014proposed} on the sphere.
+An initial state in thermal wind balance is perturbed to produce a mid-latitude baroclinic instability, providing a flow that is typical of those found in Numerical Weather Prediction.
+The surface pressure and temperature fields after 10 days are shown in Figure \ref{fig:baroclinic}, which compare well with those in \citep{melvin2024mixed} and demonstrate that our methods can replicate results from the literature.
+
+\begin{figure}[htp!]
+\centering
+\includegraphics[width=0.9\textwidth]{galewsky_jet.png}
+\caption{The surface pressure and temperature after 10 days of a deep baroclinic wave simulation on the cubed-sphere.}
+\label{fig:baroclinic}
+\end{figure}
 

### 2025-08-04 source/paper.tex
some transport terms

--- 2025-08-04 source/paper.tex
@@
 sufficient to define the result of applying the differential operator
-directly.
+directly. For the domains considered in this paper, the only boundary
+conditions we use are those of no normal flow, $\MM{u}\cdot\MM{n}=0$
+where $\MM{n}$ is the outward pointing normal to the boundary. This
+means that the test function for the velocity must also be chosen from
+a function space that has this boundary condition.
 
 In line with the theory outlined in the introduction, we always choose
-the velocity $\MM{u} \in \mathbb{V}^{HDiv}$ with $\mathbb{V}^{HDiv}
-\subset H(\text{div})$. **mention bcs here! The field whose gradient
-appears in the `pressure' gradient term is chosen to be in the space
+the velocity $\MM{u} \in \ring{\mathbb{V}}^{HDiv}$ with
+$\ring{\mathbb{V}}^{HDiv} \subset \mathbb{V}^{HDiv} \subset
+H(\text{div})$ and, as mentioned above, the $\ring{}$ indicates that
+the functions in $\ring{\mathbb{V}}^{HDiv}$ satisfy the no normal flow
+boundary conditions. The field whose gradient appears in the
+`pressure' gradient term is chosen to be in the space
 $\mathbb{V}^{L^2}$ with $\mathbb{V}^{L^2} \subset L_2$, or, in the
@@
 \begin{equation}
-\int_\Omega\MM{w}\cdot\nabla p \diff x = -\int_\Omega\nabla\cdot\MM{w}p\diff x, \quad \forall \MM{w} \in \mathbb{V}_u,
+\int_\Omega\MM{w}\cdot\nabla p \diff x = -\int_\Omega\nabla\cdot\MM{w}p\diff x, \quad \forall \MM{w} \in \ring{\mathbb{V}}^{HDiv},
 \end{equation}
@@
 element facets and on exterior boundaries we set no normal flow
-boundary conditions**. However, the pressure gradient term in the
+boundary conditions. However, the pressure gradient term in the
 compressible Euler equations, $c_p\theta\nabla\Pi$, is nonlinear and
@@
 still required since $\Pi$ is evaluated as a function of the density
-$\rho \in \mathbb{V}_\rho$ and $\theta \in \mathbb{V}_\theta$ so is
+$\rho \in \mathbb{V}^{L^2}$ and $\theta \in \mathbb{V}_\theta$ so is
 discontinuous between elements, but then integrating by parts
@@
 \begin{equation}
-  c_p\int_e\MM{w}\cdot\theta\nabla\Pi\diff x = -c_p\int_e\nabla\cdot(\MM{w}\theta)\Pi\diff x + c_p\int_{\partial e}\hat{\theta}\MM{w}\cdot\MM{n}\hat{\Pi}\diff S,
+  c_p\int_e\MM{w}\cdot\theta\nabla\Pi\diff x = -c_p\int_e\nabla\cdot(\MM{w}\theta)\Pi\diff x + c_p\int_{\partial e}\hat{\theta}\MM{w}\cdot\MM{n}\hat{\Pi}\diff S, \quad \forall \MM{w} \in \ring{\mathbb{V}}^{HDiv},
 \end{equation}
@@
 \begin{equation}
-  c_p\int_\Omega\MM{w}\cdot\theta\nabla\Pi\diff x = -c_p\int_\Omega\nabla\cdot(\MM{w}\theta)\Pi\diff x + c_p\int_{\Gamma_v}\jump{\theta\MM{w}\cdot\MM{n}}\avg{\Pi} \diff S,
+  c_p\int_\Omega\MM{w}\cdot\theta\nabla\Pi\diff x = -c_p\int_\Omega\nabla\cdot(\MM{w}\theta)\Pi\diff x + c_p\int_{\Gamma_v}\jump{\theta\MM{w}\cdot\MM{n}}\avg{\Pi} \diff S, \quad \forall \MM{w} \in \ring{\mathbb{V}}^{HDiv},
 \end{equation}
@@
 Since the prognostic variables are discretised using different finite
-element function spaces $\mathbb{V}_u$, $\mathbb{V}_p$ and
+element function spaces $\mathbb{V}^{HDiv}$, $\mathbb{V}^{L^2}$ and
 $\mathbb{V}_\theta$, each with different continuity properties between
@@
 transport scheme. In particular, when the lowest order space
-$\mathbb{V}_p$ is chosen to be DG0, we need to use the recovered
-transport scheme described in \citet{bendall2019recovered} to achieve
-second order convergence overall. Here we describe the default
-transport schemes used in Gusto for higher order function spaces and
-refer the reader to \citet{bendall2019recovered} for details of the
-recovered schemes, which are also implemented in Gusto.
+$\mathbb{V}^{L2}$ is chosen to be piecewise constant, we need to use
+the recovered transport scheme described in
+\citet{bendall2019recovered} to achieve second order convergence
+overall. Here we describe the default transport schemes used in Gusto
+for the next-to-lowest order sequence of function spaces (i.e. where
+the lowest order space $\mathbb{V}^{L2}$ is chosen to be piecewise
+linear and discontinuous between elements) and refer the reader to
+\citet{bendall2019recovered} for details of the recovered schemes,
+which are also implemented in Gusto.
+
+\subsubsection{Transport of $\mathbb{V}^{L^2}$ fields}
+The shallow water and compressible Euler equation sets both contain a
+prognostic equation that is the continuity form of the transport
+equation for a variable represented by a function that is fully
+discontinuous between elements. This means that the divergence term is
+not well defined. A stable discretiation of this equation can be
+constructed by integrating that term by parts and taking the upwind
+values on the element facets. Using the compressible Euler equation
+notation, this gives, in each element $e$ with boundary $\partial e$:
+
+\begin{equation}
+  \int_e \phi\nabla\cdot(\rho\MM{u}) \diff x = -\int_e \nabla\phi \cdot (\rho\MM{u}) \diff x + \int_{\partial e} \phi\tilde{\rho}\MM{u}\cdot\MM{n} \diff S, \quad \forall \phi \in \mathbb{V}^{L^2}.
+\end{equation}
+Summing over the domain gives
+\begin{equation}
+  \int_\Omega \phi\nabla\cdot(\rho\MM{u}) \diff x = -\int_\Omega \nabla\phi \cdot (\rho\MM{u}) \diff x + \int_\Gamma \jump{\phi\MM{u}}\tilde{\rho} \diff S, \quad \forall \phi \in \mathbb{V}^{L^2},
+\end{equation}
+where $\Omega$ is the domain, $\Gamma$ denotes the element facets,
+$\tilde{\rho}$ denotes the value of $\rho$ on the upwind side of the
+facet and the brackets $\jump{q} = q^+-q^-$ denote the jump in $q$
+across the facet.
+
+\subsubsection{Transport of $\mathbb{V}_\theta$ fields}
+The compressible Boussinesq and compressible Euler equation sets both
+contain a prognostic equation that is the advective form of the
+transport equation for a variable that is represented by a function
+that is continuous in the vertical direction but discontinuous in the
+horizontal direction. This means that the gradient of the field
+required in the transport term is not well defined. A stable
+discretisation of this equation can be constructed by using a
+combination of the above upwinding strategy in the discontinuous
+horizontal direction and a streamline-upwind Petrov-Galerkin (SUPG)
+method in the continuous vertical dirction. The SUPG method replaces
+the finite element test function $\gamma$ with $\gamma + \tau
+\MM{u}\cdot\nabla\gamma$ where $\tau$ is a stabilisation parameter,
+although here we just require the vertical part of the gradient term. This
+has the effect of suppressing spurious oscillation associated with the
+continuous finite element approximation of the transport term. The
+stabilisation parameter $\tau$ depends on the spatial resolution and
+** and is set to be ** by default in Gusto, following \citet{}.
+
+We first integrate by parts in each column $C$ of the mesh, taking the
+upwind value of the field, indicated by a $\tilde{}$, on the vertical
+facets of the mesh, denoted by $\Gamma_v$. Using the notation of the
+compressible Euler equations, this gives
+
+\begin{align}
+  \int_C \gamma\MM{u}\cdot\nabla\theta\diff x &= -\int_C \nabla\cdot(\gamma\MM{u})\theta \diff x + \int_{\Gamma_v}\gamma \MM{u}\cdot\MM{n}\tilde{\theta}\diff S, \quad \forall \gamma \in \mathbb{V}_\theta, \\
+  &= \int_C \gamma\MM{u}\cdot\nabla\theta \diff x + \int_{\Gamma_v}\gamma \MM{u}\cdot\MM{n}(\tilde{\theta}-\theta_{int})\diff S, \quad \forall \gamma \in \mathbb{V}_\theta,
+\end{align}
+where we have integrated by parts a second time to move the
+differential operator from $\gamma$ back to $\theta$ since we need to
+replace the test function $\gamma$ with $\gamma+\tau w\gamma_z$, where
+$w=\MM{u}\cdot\hat{\MM{z}}$ is the vertical component of the
+velocity. This gives a second boundary term that now contains the
+value of $\theta$ in the interior of the column, denoted by
+$\theta_{int}$. Summing over the domain and applying the SUPG method
+in the vertical gives
+\begin{equation}
+  \int_C \gamma\MM{u}\cdot\nabla\theta\diff x =\int_\Omega(\gamma+\tau w\gamma_z)\MM{u}\cdot\nabla\theta \diff x + \int_{\Gamma_v}\jump{(\gamma+\tau w\gamma_z) \MM{u}}\tilde{\theta}-\jump{(\gamma+\tau w\gamma_z) \MM{u}\theta}\diff S, \quad \forall \gamma \in \mathbb{V}_\theta.
+\end{equation}
+Note that, as explained in \citet{Colin}, the test function should
+also be replaced in the time derivative term for consistency **
+
+\subsubsection{Transport of $\mathbb{V}^{HDiv}$ fields}
 

### 2025-08-05 source/paper.tex
a bit more typing about transport

@@
 \subsubsection{Transport of $\mathbb{V}^{HDiv}$ fields}
-
-These schemes do not preserve the monotonicity of the continuous
-equation, generating spurious over and undershoots. When monotonicity
-is important, for example, to avoid generating spurious and unphysical
-negative values of moisture variables, we can apply a limiter as part
-of the transport scheme.
-
-\JScomment{I think that we should at least mention that we can do horizonal and vertical splitting for stability - maybe don't need to write out all of the forms though!}
+All three of the equation sets described in section \ref{} include a
+prognostic equation for the velocity field
+$\MM{u}\in\mathbb{V}^{HDiv}$ that contains a nonlinear transport term
+$\MM{u}\cdot\nabla\MM{u}$. This term can be discretised directly using
+a vector version of the scheme in section ** or, as in \citet{} it can
+be replaced by the vector invariant form and then discretised. In
+Gusto we offer both options and we present the vector invariant scheme
+below.
+
+The schemes described above do not preserve monotonicity unlike the
+continuous equation, generating spurious over and undershoots. When
+monotonicity is important, for example, to avoid generating spurious
+and unphysical negative values of moisture variables, we can apply a
+limiter as part of the transport scheme. These limiters are described
+in \ref{}.
+
+It is common in weather and climate models for the horizontal and
+vertical directions to be discretised differently due to the different
+scales of motion in each direction. Typically the vertical part of the
+transport term is treated implicitly so that the timestep is not
+restricted by the vertical Courant number. We provide the option to
+split the transport terms into horizontal and vertical components, as
+described in \ref{}.
 

### 2025-08-05 source/paper.tex
add figures

--- 2025-08-04 source/paper.tex
@@
 The initial conditions and moisture parametrisation are those of \citep{bendall2020compatible}, which is an extension of the dry case of \citep{kelly2012continuous} and \citep{melvin2019mixed}.
-Cross-sections of the initial and final equivalent potential temperature field are shown in Figure \ref{fig:3d_bubble}.
+A cross-sections of the final equivalent potential temperature field is shown in Figure \ref{fig:3d_bubble}.
 As in the two-dimensional test of \citep{bryan2002benchmark}, the atmosphere is saturated and cloudy everywhere, so that the moist processes are active in the whole domain.
@@
 \centering
-\includegraphics[width=0.9\textwidth]{moist_bryan_fritsch.png}
-\caption{Vertical cross-sections of the (left) initial and (right) final equivalent potential temperature field from the three-dimensional moist rising thermal test.
+\includegraphics[width=0.48\textwidth]{moist_3d_bubble.png}
+\caption{Vertical cross-section of the final equivalent potential temperature field from the three-dimensional moist rising thermal test.
 }

### 2025-08-05 source/paper.tex
Jemma edits and comments on results

@@
 spatial discretisation options and timestepping schemes.
-As these options can be combined in so many ways, it is only possible to present results from a tiny fraction configurations.
+As these options can be combined in so many ways, it is only possible to present results from a tiny fraction of the configurations.
 We have chosen to use a selection from the hierarchy of standard test cases that span the stages of development of dynamical cores.
@@
 For this test, we use the DG upwind scheme described in Section \ref{sec:transport} to transport the species on an icosahedral sphere mesh, and apply the vertex-based limiter of \cite{kuzmin2010vertex}.
-The term described chemical reaction is separated from the using a simple sequential splitting, \textcolor{red}{as described in Section \?}. \\
+The term describing the chemical reaction is separated from the transport term using a simple sequential splitting, \textcolor{red}{as described in Section \?}. \\
 \\
@@
 Here we present two sets of convergence results for a variety of configurations measured using the steady state geostrophic balanced flow (test case 2 from
-\citet{williamson1992standard}) with the linear shallow water equations.
+\citet{williamson1992standard}) with the linear shallow water equations. \JScomment{I think Hilary should also be cited here as she uses the linear version}
 Figure \ref{fig:sw_convergence} shows the spatial convergence for a number of choices of finite element for the prognostic variables, using a Trapezium Rule discretisation (also known as Crank-Nicolson).
@@
 \\
-We show results from the unstable jet of \citep{galewsky2004initial} in Figure \ref{fig:galewsky}, using the non-linear shallow water equations.
+We show results from the unstable jet of \citet{galewsky2004initial} in Figure \ref{fig:galewsky}, using the non-linear shallow water equations.
 These results use the Semi-Implicit Quasi-Newton timestepper described in Section \ref{???}, and compare the RT elements on a cubed-sphere mesh with the BDM elements on an icosahedral-sphere.
+\JScomment{both first order?}
 
@@
 \subsection{Compressible Euler equations}
-We have also implemented many of the standard test cases for the non-hydrostatic compressible Euler equations that are used in modern NWP models.
+We have also implemented many of the standard test cases for the non-hydrostatic compressible Euler equations that are used in modern NWP \JScomment{I think I've been saying `weather and climate' instead of NWP - happy to change but this is a note to remind us to be consistent!'} models.
 In this section we show results from a small selection of these, which demonstrate different domains in which these methods work.
@@
 \\
-The first test is the falling density current of \cite{straka1993numerical} in a vertical slice. This test involves the addition of diffusion to both the potential temperature and velocity equations.
+The first test is the falling density current of \citet{straka1993numerical} in a vertical slice. This test involves the addition of diffusion to both the potential temperature and velocity equations.
 Figure \ref{fig:straka} shows the final potential temperature equation for simulations using the lowest-order and next-to-lowest order finite elements on the vertical slice.
@@
 To demonstrate the use of moist physics parametrisations, we use a moist three-dimensional rising thermal simulation in an extruded doubly-periodic plane.
-The initial conditions and moisture parametrisation are those of \citep{bendall2020compatible}, which is an extension of the dry case of \citep{kelly2012continuous} and \citep{melvin2019mixed}.
+The initial conditions and moisture parametrisation are those of \citet{bendall2020compatible}, which is an extension of the dry case of \citet{kelly2012continuous} and \citet{melvin2019mixed}.
 A cross-sections of the final equivalent potential temperature field is shown in Figure \ref{fig:3d_bubble}.
-As in the two-dimensional test of \citep{bryan2002benchmark}, the atmosphere is saturated and cloudy everywhere, so that the moist processes are active in the whole domain.
+As in the two-dimensional test of \citet{bryan2002benchmark}, the atmosphere is saturated and cloudy everywhere, so that the moist processes are active in the whole domain.
 \\
@@
 \\
-Our final test case is the deep baroclinic wave of \citep{ullrich2014proposed} on the sphere.
-An initial state in thermal wind balance is perturbed to produce a mid-latitude baroclinic instability, providing a flow that is typical of those found in Numerical Weather Prediction.
-The surface pressure and temperature fields after 10 days are shown in Figure \ref{fig:baroclinic}, which compare well with those in \citep{melvin2024mixed} and demonstrate that our methods can replicate results from the literature.
+Our final test case is the deep baroclinic wave of \citet{ullrich2014proposed} on the sphere.
+An initial state in thermal wind balance is perturbed to produce a mid-latitude baroclinic instability, providing a flow that is typical of those found in Numerical Weather Prediction \JScomment{not sure I approve of these capitals...}.
+The surface pressure and temperature fields after 10 days are shown in Figure \ref{fig:baroclinic}, which compare well with those in \citet{melvin2024mixed} and demonstrate that our methods can replicate results from the literature.
 

### 2025-08-05 source/paper.tex

--- 2025-08-05 source/paper.tex
@@
 \subsubsection{Transport of $\mathbb{V}^{HDiv}$ fields}
-All three of the equation sets described in section \ref{} include a
-prognostic equation for the velocity field
-$\MM{u}\in\mathbb{V}^{HDiv}$ that contains a nonlinear transport term
-$\MM{u}\cdot\nabla\MM{u}$. This term can be discretised directly using
-a vector version of the scheme in section ** or, as in \citet{} it can
-be replaced by the vector invariant form and then discretised. In
-Gusto we offer both options and we present the vector invariant scheme
-below.
-
-The schemes described above do not preserve monotonicity unlike the
-continuous equation, generating spurious over and undershoots. When
-monotonicity is important, for example, to avoid generating spurious
-and unphysical negative values of moisture variables, we can apply a
-limiter as part of the transport scheme. These limiters are described
-in \ref{}.
-
-It is common in weather and climate models for the horizontal and
-vertical directions to be discretised differently due to the different
-scales of motion in each direction. Typically the vertical part of the
-transport term is treated implicitly so that the timestep is not
-restricted by the vertical Courant number. We provide the option to
-split the transport terms into horizontal and vertical components, as
-described in \ref{}.
+
+These schemes do not preserve the monotonicity of the continuous
+equation, generating spurious over and undershoots. When monotonicity
+is important, for example, to avoid generating spurious and unphysical
+negative values of moisture variables, we can apply a limiter as part
+of the transport scheme.
+
+\JScomment{I think that we should at least mention that we can do horizonal and vertical splitting for stability - maybe don't need to write out all of the forms though!}
 
@@
 The aim of this section is to present a selection of results that
-demonstrate the tools available in Gusto, such as the choice of grids,
+showcase the tools available in Gusto, such as the choice of grids,
 HDiv element families, order of finite element spaces, different
-spatial discretisation options, timestepping schemes and physics. The
-numerical properties of these configurations (convergence, stability,
-accuracy and efficiency) have either been demonstrated elsewhere, or
-will be investigated more thoroughly in future publications.
-
-To include:
-\begin{itemize}
-\item terminator toy (Tim)
-\item shallow water on different grids and with different HDiv
-  families (Nell) e.g. Williamson 5 cubed sphere PV and Galewsky
-  icosahedral sphere PV, high res (ref 6)
-\item convergence with different RK schemes / multistage schemes using
-  Williamson 2 or 6, L2 and Linf errors for D and u (2 plots, one L2,
-  one Linf) (Alex)
-\item vertical slice Euler - Straka, different orders, illustrates
-  recovery for diffusion (Tom)
-\item 3D moist bubble with rain (Tom/Alex)
-\item spherical gravity wave, Boussinesq (Jemma)
-\item spherical baroclinic wave, Euler - whatever people usually plot
-  (horizontal slice pressure contours, temperature shading) (Dan)
-\item Held Suarez??
-\end{itemize}
-
-\subsection{Terminator `toy' chemistry}
-This test combines transport and a `toy' chemistry model to
-demonstrate physics-dynamics coupling in Gusto (where here chemistry
-is the `physics').
+spatial discretisation options and timestepping schemes.
+As these options can be combined in so many ways, it is only possible to present results from a tiny fraction of the configurations.
+We have chosen to use a selection from the hierarchy of standard test cases that span the stages of development of dynamical cores.
+Although we do include some results here that demonstrate the numerical properties of certain configurations, the purpose of this section is not to provide a rigorous verification of any particular choice of discretisation.
+Such results are demonstrated elsewhere, for instance \cite{bendall2020compatible} or \cite{hartney2025exploring}.
+The settings used for each result from this section are detailed in Appendix \ref{sec:appendix}.
+
+\subsection{Terminator Toy chemistry}
+The first results use the ``Terminator Toy'' test of \cite{lauritzen2015terminator}, which involves the transport of two coupled tracers $X$ and $X_2$ on the surface of a sphere.
+The chemicals interact such that the sum $m_T=m_X+2m_{X_2}$ should be preserved, where $m_X$ and $m_{X_2}$ are the mixing ratios of the individual chemicals.
+When positivity or monotonicity limiters are applied to the numerical transport scheme, this may break the correlation between the two species which will be highlighted by this test.
+This test is also a demonstration of simple ``physics-dynamics coupling'', as the conversion term between the species can be handled separately from the transport terms. \\
+\\
+For this test, we use the DG upwind scheme described in Section \ref{sec:transport} to transport the species on an icosahedral sphere mesh, and apply the vertex-based limiter of \cite{kuzmin2010vertex}.
+The term describing the chemical reaction is separated from the transport term using a simple sequential splitting, \textcolor{red}{as described in Section \?}. \\
+\\
+Figure \ref{fig:terminator} shows the mixing ratios of the two individual species at the initial time and following a single rotation around the sphere by a deformational flow.
+The change in the sum of the species is also displayed, with differences from zero remaining close to machine precision, which demonstrates that this transport scheme does preserve the correlation between the tracers.
+
+\begin{figure}[htp!]
+\centering
+\includegraphics[width=0.9\textwidth]{terminator_toy.png}
+\caption{The evolution of the chemical species from the ``Terminator Toy'' test of \cite{lauritzen2015terminator}.
+The fields at the initial time are shown on the left, and at the final time ($t=12$ days) which follows a single revolution around the sphere by a deformational flow.
+The bottom row shows the change in the total $m_T=m_X+2m_{X_2}$ from the initial time, which remains very close to zero.}
+\label{fig:terminator}
+\end{figure}
 
@@
 \citet{williamson1992standard} provides a common starting point for
-testing new discretisations. This test suite is usually augmented by
+testing new discretisations. This test suite is often augmented by
 the unstable barotropic jet described in
-\citet{galewsky2004initial}. Here we present convergence results for a
-variety of timestepping schemes applied to the steady state
-geostrophic balanced flow (test case 2 from
-\citet{williamson1992standard}) and a demonstration of two different
-grids and HDiv element families for shallow water flow over a mountain
-(test case 5 from \citet{williamson1992standard}) and the unstable
-barotropic jet \citep{galewsky2004initial}.
-
-\subsubsection{Convergence results}
-
-\subsubsection{Grids and finite element families}
-
-\subsection{Compressible Boussinesq equations}
+\citet{galewsky2004initial}. \\
+\\
+Here we present two sets of convergence results for a variety of configurations measured using the steady state geostrophic balanced flow (test case 2 from
+\citet{williamson1992standard}) with the linear shallow water equations. \JScomment{I think Hilary should also be cited here as she uses the linear version}
+Figure \ref{fig:sw_convergence} shows the spatial convergence for a number of choices of finite element for the prognostic variables, using a Trapezium Rule discretisation (also known as Crank-Nicolson).
+It also shows the temporal convergence for a range of timestepping schemes using the 1st order Raviart-Thomas elements on quadrilateral cells. 
+\textcolor{red}{Once the test has been completed, come back to edit what we say here about convergence!} \\
+\\
+We show results from the unstable jet of \citet{galewsky2004initial} in Figure \ref{fig:galewsky}, using the non-linear shallow water equations.
+These results use the Semi-Implicit Quasi-Newton timestepper described in Section \ref{???}, and compare the RT elements on a cubed-sphere mesh with the BDM elements on an icosahedral-sphere.
+\JScomment{both first order?}
+
+\begin{figure}[htp!]
+\centering
+\includegraphics[width=0.9\textwidth]{}
+\caption{Convergence results using the linear shallow water equations with the steady geostrophic flow of \cite{williamson1992standard}.
+(Left) the spatial convergence of a number of orders and families of finite element, all using the same timestepper.
+(Right) the temporal convergence of different timesteppers, all using the same spatial discretisation.}
+\label{fig:sw_convergence}
+\end{figure}
+
+\begin{figure}[htp!]
+\centering
+\includegraphics[width=0.9\textwidth]{galewsky_jet.png}
+\caption{A view of the relative vorticity around the North pole at day 6 of a simulation of the Galewsky jet test case.
+(Left) the RT1 elements on a cubed-sphere mesh.
+(Right) the BDM1 elements on an icosahedral-sphere mesh.
+Both cases use the Semi-Implicit Quasi-Newton timestepper.
+The edges of the panels of the cube and icosahedron are highlighted.}
+\label{fig:galewsky}
+\end{figure}
 
 \subsection{Compressible Euler equations}
-
-\subsubsection{Dense bubble in a vertical slice}
-
-\subsubsection{3D moist rising bubble}
-
-\subsubsection{Baroclinic wave on the sphere}
-
-\subsubsection{Held Suarez}
+We have also implemented many of the standard test cases for the non-hydrostatic compressible Euler equations that are used in modern NWP \JScomment{I think I've been saying `weather and climate' instead of NWP - happy to change but this is a note to remind us to be consistent!'} models.
+In this section we show results from a small selection of these, which demonstrate different domains in which these methods work.
+Throughout this section we use the Semi-Implicit Quasi-Newton timestepper.
+ \\
+\\
+The first test is the falling density current of \citet{straka1993numerical} in a vertical slice. This test involves the addition of diffusion to both the potential temperature and velocity equations.
+Figure \ref{fig:straka} shows the final potential temperature equation for simulations using the lowest-order and next-to-lowest order finite elements on the vertical slice.
+The lowest-order methods require the transport and diffusion terms to use the ``recovered'' scheme of \citet{bendall2019recovered} to attain second-order accuracy. 
+\\
+\begin{figure}[htp!]
+\centering
+\includegraphics[width=0.8\textwidth]{straka.png}
+\caption{The potential temperature field after 900 s from the falling density current test of \cite{straka1993numerical}.
+(Top) with the lowest-order finite elements, and (bottom) with the next-to-lowest-order finite elements.
+}
+\label{fig:straka}
+\end{figure}
+\\
+To demonstrate the use of moist physics parametrisations, we use a moist three-dimensional rising thermal simulation in an extruded doubly-periodic plane.
+The initial conditions and moisture parametrisation are those of \citet{bendall2020compatible}, which is an extension of the dry case of \citet{kelly2012continuous} and \citet{melvin2019mixed}.
+A cross-sections of the final equivalent potential temperature field is shown in Figure \ref{fig:3d_bubble}.
+As in the two-dimensional test of \citet{bryan2002benchmark}, the atmosphere is saturated and cloudy everywhere, so that the moist processes are active in the whole domain.
+\\
+\begin{figure}[htp!]
+\centering
+\includegraphics[width=0.48\textwidth]{moist_3d_bubble.png}
+\caption{Vertical cross-section of the final equivalent potential temperature field from the three-dimensional moist rising thermal test.
+}
+\label{fig:3d_bubble}
+\end{figure}
+\\
+Our final test case is the deep baroclinic wave of \citet{ullrich2014proposed} on the sphere.
+An initial state in thermal wind balance is perturbed to produce a mid-latitude baroclinic instability, providing a flow that is typical of those found in Numerical Weather Prediction \JScomment{not sure I approve of these capitals...}.
+The surface pressure and temperature fields after 10 days are shown in Figure \ref{fig:baroclinic}, which compare well with those in \citet{melvin2024mixed} and demonstrate that our methods can replicate results from the literature.
+
+\begin{figure}[htp!]
+\centering
+\includegraphics[width=0.9\textwidth]{galewsky_jet.png}
+\caption{The surface pressure and temperature after 10 days of a deep baroclinic wave simulation on the cubed-sphere.}
+\label{fig:baroclinic}
+\end{figure}
 

### 2025-08-05 source/paper.tex
fix tiny error

@@
 The first test is the falling density current of \citet{straka1993numerical} in a vertical slice. This test involves the addition of diffusion to both the potential temperature and velocity equations.
-Figure \ref{fig:straka} shows the final potential temperature equation for simulations using the lowest-order and next-to-lowest order finite elements on the vertical slice.
+Figure \ref{fig:straka} shows the final potential temperature field for simulations using the lowest-order and next-to-lowest order finite elements on the vertical slice.
 The lowest-order methods require the transport and diffusion terms to use the ``recovered'' scheme of \citet{bendall2019recovered} to attain second-order accuracy. 

### 2025-08-05 source/paper.tex
small changes e.g. spelling

@@
 grids of a spherical manifold will necessarily have singular points
-that lead to a degredation in parallel scaling. An example of this is
+that lead to a degradation in parallel scaling. An example of this is
 the commonly-used latitude-longitude grid where the singular points
@@
 % supercomputer architectures
-% are run on massivly parallel supercomputers that require the co
+% are run on massively parallel supercomputers that require the co
 % Spatial domain decomposition methods share the computation
@@
 range of different equation sets relevant for weather and climate
-modelling. In section \ref{sec: governing} we will outline three
+modelling. In section \ref{sec: equations} we will outline three
 different equation sets commonly used for modelling geophysical fluids
@@
 types of terms (e.g. pressure gradient, transport, diffusion) seen in
-section \ref{sec: governing}. In section \ref{sec: design} we describe
+section \ref{sec: equations}. In section \ref{sec: design} we describe
 the code structure of Gusto followed by a detailed example (section
@@
 \section{Governing Equations}
-\label{sec: governing}
+\label{sec: equations}
 The aim of this section is to introduce three governing equation sets
@@
 and climate models, along with enough detail about the finite element
-discretiations provided by Gusto to demonstrate the similarities
+discretisations provided by Gusto to demonstrate the similarities
 between them. These similarities guide the design of the code
@@
 \subsection{Transport terms}
+\label{sec: transport}
 Since the prognostic variables are discretised using different finite
@@
 \subsubsection{Transport of $\mathbb{V}^{L^2}$ fields}
+\label{subsec: DG transport}
 The shallow water and compressible Euler equation sets both contain a
@@
 discontinuous between elements. This means that the divergence term is
-not well defined. A stable discretiation of this equation can be
+not well defined. A stable discretisation of this equation can be
 constructed by integrating that term by parts and taking the upwind
@@
 \subsubsection{Transport of $\mathbb{V}_\theta$ fields}
+\label{subsec: theta transport}
 The compressible Boussinesq and compressible Euler equation sets both
@@
 horizontal direction and a streamline-upwind Petrov-Galerkin (SUPG)
-method in the continuous vertical dirction. The SUPG method replaces
+method in the continuous vertical direction. The SUPG method replaces
 the finite element test function $\gamma$ with $\gamma + \tau
@@
 although here we just require the vertical part of the gradient term. This
-has the effect of suppressing spurious oscillation associated with the
+has the effect of suppressing spurious oscillations associated with the
 continuous finite element approximation of the transport term. The
@@
 \begin{equation}
-  \int_C \gamma\MM{u}\cdot\nabla\theta\diff x =\int_\Omega(\gamma+\tau w\gamma_z)\MM{u}\cdot\nabla\theta \diff x + \int_{\Gamma_v}\jump{(\gamma+\tau w\gamma_z) \MM{u}}\tilde{\theta}-\jump{(\gamma+\tau w\gamma_z) \MM{u}\theta}\diff S, \quad \forall \gamma \in \mathbb{V}_\theta.
+  \int_\Omega \gamma\MM{u}\cdot\nabla\theta\diff x =\int_\Omega(\gamma+\tau w\gamma_z)\MM{u}\cdot\nabla\theta \diff x + \int_{\Gamma_v}\jump{(\gamma+\tau w\gamma_z) \MM{u}}\tilde{\theta}-\jump{(\gamma+\tau w\gamma_z) \MM{u}\theta}\diff S, \quad \forall \gamma \in \mathbb{V}_\theta.
 \end{equation}
@@
 \subsubsection{Transport of $\mathbb{V}^{HDiv}$ fields}
-All three of the equation sets described in section \ref{} include a
-prognostic equation for the velocity field
+All three of the equation sets described in section \ref{sec:
+  equations} include a prognostic equation for the velocity field
 $\MM{u}\in\mathbb{V}^{HDiv}$ that contains a nonlinear transport term
 $\MM{u}\cdot\nabla\MM{u}$. This term can be discretised directly using
-a vector version of the scheme in section ** or, as in \citet{} it can
-be replaced by the vector invariant form and then discretised. In
-Gusto we offer both options and we present the vector invariant scheme
-below.
+a vector version of the scheme in section \ref{subsec: theta
+  transport} or, as shown in \citet{}, it can be replaced by the vector
+invariant form and then discretised. In Gusto we offer both options
+and we present the vector invariant scheme below.
 
@@
 equation sets are each defined in a python class, making use of the
-common finite element forms for the transport and difffusion terms to
+common finite element forms for the transport and diffusion terms to
 set up the residual (defined in section \ref{}) in weak form. At this
@@
 
-\subsection{Example 1: IMEX-Euler applied to the 1d advection diffuction equation}
+\subsection{Example 1: IMEX-Euler applied to the 1d advection diffusion equation}
 
@@
 in the equation based on the presence or absence of certain
-labels. The full strength of FML becomes even more apparant when we
+labels. The full strength of FML becomes even more apparent when we
 exploit information stored in the label value. We use this in two
@@
 The aim of this section is to present a selection of results that
-showcase the tools available in Gusto, such as the choice of grids,
+demonstrate the tools available in Gusto, such as the choice of grids,
 HDiv element families, order of finite element spaces, different
-spatial discretisation options and timestepping schemes.
-As these options can be combined in so many ways, it is only possible to present results from a tiny fraction of the configurations.
-We have chosen to use a selection from the hierarchy of standard test cases that span the stages of development of dynamical cores.
-Although we do include some results here that demonstrate the numerical properties of certain configurations, the purpose of this section is not to provide a rigorous verification of any particular choice of discretisation.
-Such results are demonstrated elsewhere, for instance \cite{bendall2020compatible} or \cite{hartney2025exploring}.
-The settings used for each result from this section are detailed in Appendix \ref{sec:appendix}.
-
-\subsection{Terminator Toy chemistry}
-The first results use the ``Terminator Toy'' test of \cite{lauritzen2015terminator}, which involves the transport of two coupled tracers $X$ and $X_2$ on the surface of a sphere.
-The chemicals interact such that the sum $m_T=m_X+2m_{X_2}$ should be preserved, where $m_X$ and $m_{X_2}$ are the mixing ratios of the individual chemicals.
-When positivity or monotonicity limiters are applied to the numerical transport scheme, this may break the correlation between the two species which will be highlighted by this test.
-This test is also a demonstration of simple ``physics-dynamics coupling'', as the conversion term between the species can be handled separately from the transport terms. \\
-\\
-For this test, we use the DG upwind scheme described in Section \ref{sec:transport} to transport the species on an icosahedral sphere mesh, and apply the vertex-based limiter of \cite{kuzmin2010vertex}.
-The term describing the chemical reaction is separated from the transport term using a simple sequential splitting, \textcolor{red}{as described in Section \?}. \\
-\\
-Figure \ref{fig:terminator} shows the mixing ratios of the two individual species at the initial time and following a single rotation around the sphere by a deformational flow.
-The change in the sum of the species is also displayed, with differences from zero remaining close to machine precision, which demonstrates that this transport scheme does preserve the correlation between the tracers.
-
-\begin{figure}[htp!]
-\centering
-\includegraphics[width=0.9\textwidth]{terminator_toy.png}
-\caption{The evolution of the chemical species from the ``Terminator Toy'' test of \cite{lauritzen2015terminator}.
-The fields at the initial time are shown on the left, and at the final time ($t=12$ days) which follows a single revolution around the sphere by a deformational flow.
-The bottom row shows the change in the total $m_T=m_X+2m_{X_2}$ from the initial time, which remains very close to zero.}
-\label{fig:terminator}
-\end{figure}
+spatial discretisation options, timestepping schemes and physics. The
+numerical properties of these configurations (convergence, stability,
+accuracy and efficiency) have either been demonstrated elsewhere, or
+will be investigated more thoroughly in future publications.
+
+To include:
+\begin{itemize}
+\item terminator toy (Tim)
+\item shallow water on different grids and with different HDiv
+  families (Nell) e.g. Williamson 5 cubed sphere PV and Galewsky
+  icosahedral sphere PV, high res (ref 6)
+\item convergence with different RK schemes / multistage schemes using
+  Williamson 2 or 6, L2 and Linf errors for D and u (2 plots, one L2,
+  one Linf) (Alex)
+\item vertical slice Euler - Straka, different orders, illustrates
+  recovery for diffusion (Tom)
+\item 3D moist bubble with rain (Tom/Alex)
+\item spherical gravity wave, Boussinesq (Jemma)
+\item spherical baroclinic wave, Euler - whatever people usually plot
+  (horizontal slice pressure contours, temperature shading) (Dan)
+\item Held Suarez??
+\end{itemize}
+
+\subsection{Terminator `toy' chemistry}
+This test combines transport and a `toy' chemistry model to
+demonstrate physics-dynamics coupling in Gusto (where here chemistry
+is the `physics').
 
@@
 \citet{williamson1992standard} provides a common starting point for
-testing new discretisations. This test suite is often augmented by
+testing new discretisations. This test suite is usually augmented by
 the unstable barotropic jet described in
-\citet{galewsky2004initial}. \\
-\\
-Here we present two sets of convergence results for a variety of configurations measured using the steady state geostrophic balanced flow (test case 2 from
-\citet{williamson1992standard}) with the linear shallow water equations. \JScomment{I think Hilary should also be cited here as she uses the linear version}
-Figure \ref{fig:sw_convergence} shows the spatial convergence for a number of choices of finite element for the prognostic variables, using a Trapezium Rule discretisation (also known as Crank-Nicolson).
-It also shows the temporal convergence for a range of timestepping schemes using the 1st order Raviart-Thomas elements on quadrilateral cells. 
-\textcolor{red}{Once the test has been completed, come back to edit what we say here about convergence!} \\
-\\
-We show results from the unstable jet of \citet{galewsky2004initial} in Figure \ref{fig:galewsky}, using the non-linear shallow water equations.
-These results use the Semi-Implicit Quasi-Newton timestepper described in Section \ref{???}, and compare the RT elements on a cubed-sphere mesh with the BDM elements on an icosahedral-sphere.
-\JScomment{both first order?}
-
-\begin{figure}[htp!]
-\centering
-\includegraphics[width=0.9\textwidth]{}
-\caption{Convergence results using the linear shallow water equations with the steady geostrophic flow of \cite{williamson1992standard}.
-(Left) the spatial convergence of a number of orders and families of finite element, all using the same timestepper.
-(Right) the temporal convergence of different timesteppers, all using the same spatial discretisation.}
-\label{fig:sw_convergence}
-\end{figure}
-
-\begin{figure}[htp!]
-\centering
-\includegraphics[width=0.9\textwidth]{galewsky_jet.png}
-\caption{A view of the relative vorticity around the North pole at day 6 of a simulation of the Galewsky jet test case.
-(Left) the RT1 elements on a cubed-sphere mesh.
-(Right) the BDM1 elements on an icosahedral-sphere mesh.
-Both cases use the Semi-Implicit Quasi-Newton timestepper.
-The edges of the panels of the cube and icosahedron are highlighted.}
-\label{fig:galewsky}
-\end{figure}
+\citet{galewsky2004initial}. Here we present convergence results for a
+variety of timestepping schemes applied to the steady state
+geostrophic balanced flow (test case 2 from
+\citet{williamson1992standard}) and a demonstration of two different
+grids and HDiv element families for shallow water flow over a mountain
+(test case 5 from \citet{williamson1992standard}) and the unstable
+barotropic jet \citep{galewsky2004initial}.
+
+\subsubsection{Convergence results}
+
+\subsubsection{Grids and finite element families}
+
+\subsection{Compressible Boussinesq equations}
 
 \subsection{Compressible Euler equations}
-We have also implemented many of the standard test cases for the non-hydrostatic compressible Euler equations that are used in modern NWP \JScomment{I think I've been saying `weather and climate' instead of NWP - happy to change but this is a note to remind us to be consistent!'} models.
-In this section we show results from a small selection of these, which demonstrate different domains in which these methods work.
-Throughout this section we use the Semi-Implicit Quasi-Newton timestepper.
- \\
-\\
-The first test is the falling density current of \citet{straka1993numerical} in a vertical slice. This test involves the addition of diffusion to both the potential temperature and velocity equations.
-Figure \ref{fig:straka} shows the final potential temperature field for simulations using the lowest-order and next-to-lowest order finite elements on the vertical slice.
-The lowest-order methods require the transport and diffusion terms to use the ``recovered'' scheme of \citet{bendall2019recovered} to attain second-order accuracy. 
-\\
-\begin{figure}[htp!]
-\centering
-\includegraphics[width=0.8\textwidth]{straka.png}
-\caption{The potential temperature field after 900 s from the falling density current test of \cite{straka1993numerical}.
-(Top) with the lowest-order finite elements, and (bottom) with the next-to-lowest-order finite elements.
-}
-\label{fig:straka}
-\end{figure}
-\\
-To demonstrate the use of moist physics parametrisations, we use a moist three-dimensional rising thermal simulation in an extruded doubly-periodic plane.
-The initial conditions and moisture parametrisation are those of \citet{bendall2020compatible}, which is an extension of the dry case of \citet{kelly2012continuous} and \citet{melvin2019mixed}.
-A cross-sections of the final equivalent potential temperature field is shown in Figure \ref{fig:3d_bubble}.
-As in the two-dimensional test of \citet{bryan2002benchmark}, the atmosphere is saturated and cloudy everywhere, so that the moist processes are active in the whole domain.
-\\
-\begin{figure}[htp!]
-\centering
-\includegraphics[width=0.48\textwidth]{moist_3d_bubble.png}
-\caption{Vertical cross-section of the final equivalent potential temperature field from the three-dimensional moist rising thermal test.
-}
-\label{fig:3d_bubble}
-\end{figure}
-\\
-Our final test case is the deep baroclinic wave of \citet{ullrich2014proposed} on the sphere.
-An initial state in thermal wind balance is perturbed to produce a mid-latitude baroclinic instability, providing a flow that is typical of those found in Numerical Weather Prediction \JScomment{not sure I approve of these capitals...}.
-The surface pressure and temperature fields after 10 days are shown in Figure \ref{fig:baroclinic}, which compare well with those in \citet{melvin2024mixed} and demonstrate that our methods can replicate results from the literature.
-
-\begin{figure}[htp!]
-\centering
-\includegraphics[width=0.9\textwidth]{galewsky_jet.png}
-\caption{The surface pressure and temperature after 10 days of a deep baroclinic wave simulation on the cubed-sphere.}
-\label{fig:baroclinic}
-\end{figure}
+
+\subsubsection{Dense bubble in a vertical slice}
+
+\subsubsection{3D moist rising bubble}
+
+\subsubsection{Baroclinic wave on the sphere}
+
+\subsubsection{Held Suarez}
 
@@
 \appendix
-\section{Test Case Settings}  \label{sec:appendix}  %% Appendix A
+\section{}    %% Appendix A
 

### 2025-08-05 source/paper.tex
finish transport section

@@
 
+The starting point for this discretisation is to replace the
+$\MM{u}\cdot\nabla\MM{u}$ term with $(\nabla\times\MM{u})\times\MM{u}
++ \frac{1}{2}\nabla(\MM{u}\cdot\MM{u})$. Now, multiplying by a test
+function and integrating by parts over an element $e$ with boundary
+$\partial e$ gives
+\begin{equation}
+  \int_e \MM{w}\cdot(\nabla\times\MM{u})\times\MM{u}
+  + \frac{1}{2}\MM{w}\cdot\nabla(\MM{u}\cdot\MM{u}) \diff x = \int_e\nabla\times(\MM{u}\times\MM{w})\cdot\MM{u}\diff x - \int_{\partial e}\MM{n}\times(\MM{u}\times\MM{w})\cdot\MM{u} \diff S - \frac{1}{2}\int_e\nabla\cdot\MM{w}\MM{u}\cdot\MM{u}\diff x, \quad \forall \MM{w} \in \mathbb{V}^{HDiv}.
+\end{equation}
+Sunming over the domain and taking the upwind value of $\MM{u}$,
+denoted by $\tilde{\MM{u}}$, on the element boundaries gives
+\begin{equation}
+  \int_\Omega \MM{w}\cdot(\nabla\times\MM{u})\times\MM{u}
+  + \frac{1}{2}\MM{w}\cdot\nabla(\MM{u}\cdot\MM{u}) \diff x = \int_\Omega\nabla\times(\MM{u}\times\MM{w})\cdot\MM{u}\diff x - \int_{\Gamma}\jump{\MM{n}\times(\MM{u}\times\MM{w})}\cdot\tilde{\MM{u}} \diff S - \frac{1}{2}\int_\Omega\nabla\cdot\MM{w}\MM{u}\cdot\MM{u}\diff x, \quad \forall \MM{w} \in \mathbb{V}^{HDiv},
+\end{equation}
+where $\tilde{\MM{u}}$ denotes the upwind value of $\MM{u}$ on the facet.
+
 The schemes described above do not preserve monotonicity unlike the

### 2025-08-05 source/paper.tex
finish off transport section

@@
 \Author[2]{Thomas M.}{Bendall}
-\Author[3]{So many}{Others}
+\Author[3]{Colin J.}{Cotter}
+\Author[4]{Timothy}{Andrews}
+\Author[5]{Jack}{Betteridge}
+\Author[1, 2]{Alex}{Brown}
+\Author[6]{Thomas}{Gibson}
+\Author[1]{Nell}{Hartney}
+\Author[3]{Josh}{Hope-Collins}
+\Author[1]{Dan}{Witt}
 
@@
 \affil[2]{Met Office}
+\affil[3]{Imperial College}
+\affil[4]{University of Michigan}
+\affil[5]{}
+\affil[6]{}
 
@@
 the integration by parts because these vanish on interior element
-boundaries since the normal component of $\MM{w}$ is continuous across
-element facets and on exterior boundaries we set no normal flow
-boundary conditions. However, the pressure gradient term in the
+boundaries (since the normal component of $\MM{w}$ is continuous
+across element facets) and on exterior boundaries we set no normal
+flow boundary conditions. However, the pressure gradient term in the
 compressible Euler equations, $c_p\theta\nabla\Pi$, is nonlinear and
@@
 transfers the differential operator to $\MM{w}\theta$ which is
-discontinuous in the horizontal direction. Following \ref{} we integrate by parts
-in each element $e$, giving
+discontinuous in the horizontal direction. Following
+\citet{natale2016compatible} we integrate by parts in each element $e$,
+giving
 
@@
 $\theta$, and therefore $\Pi$, are discontinuous. Here we have chosen,
-as in \citet{}, to take the average value of $\Pi$ across the facets,
-$\avg{\Pi} = (\Pi^++\Pi^-)/2$, where the $+$ and $-$ indicate each
-side of the facet. The brackets $\jump{q} = q^+-q^-$ denote the jump
-in $q$ across the facet and the integral over the horizontal facets
-vanishes since both $\theta$ and $\MM{w}\cdot\MM{n}$ are continuous
-across these facets so the jump is zero.
+as in \citet{natale2016compatible}, to take the average value of $\Pi$
+across the facets, $\avg{\Pi} = (\Pi^++\Pi^-)/2$, where the $+$ and
+$-$ indicate each side of the facet. The brackets $\jump{q} = q^+-q^-$
+denote the jump in $q$ across the facet and the integral over the
+horizontal facets vanishes since both $\theta$ and $\MM{w}\cdot\MM{n}$
+are continuous across these facets so the jump is zero.
 
@@
 transport scheme. In particular, when the lowest order space
-$\mathbb{V}^{L2}$ is chosen to be piecewise constant, we need to use
+$\mathbb{V}^{L^2}$ is chosen to be piecewise constant, we need to use
 the recovered transport scheme described in
@@
 for the next-to-lowest order sequence of function spaces (i.e. where
-the lowest order space $\mathbb{V}^{L2}$ is chosen to be piecewise
+the lowest order space $\mathbb{V}^{L^2}$ is chosen to be piecewise
 linear and discontinuous between elements) and refer the reader to
@@
 discontinuous between elements. This means that the divergence term is
-not well defined. A stable discretisation of this equation can be
+not globally defined. A stable discretisation of this equation can be
 constructed by integrating that term by parts and taking the upwind
 values on the element facets. Using the compressible Euler equation
-notation, this gives, in each element $e$ with boundary $\partial e$:
+notation this gives, in each element $e$ with boundary $\partial e$,
 
 \begin{equation}
-  \int_e \phi\nabla\cdot(\rho\MM{u}) \diff x = -\int_e \nabla\phi \cdot (\rho\MM{u}) \diff x + \int_{\partial e} \phi\tilde{\rho}\MM{u}\cdot\MM{n} \diff S, \quad \forall \phi \in \mathbb{V}^{L^2}.
+  \int_e \phi\nabla\cdot(\rho\MM{u}) \diff x = -\int_e \nabla\phi \cdot (\rho\MM{u}) \diff x + \int_{\partial e} \phi\hat{\rho}\MM{u}\cdot\MM{n} \diff S, \quad \forall \phi \in \mathbb{V}^{L^2},
 \end{equation}
-Summing over the domain gives
+where $\hat{\rho}$ denotes the value of $\rho$ on the element
+boundary. Summing over the domain gives
 \begin{equation}
@@
 facet and the brackets $\jump{q} = q^+-q^-$ denote the jump in $q$
-across the facet.
+across the facet, where the $+$ and $-$ indicate each side of the
+facet.
 
@@
 horizontal direction. This means that the gradient of the field
-required in the transport term is not well defined. A stable
-discretisation of this equation can be constructed by using a
-combination of the above upwinding strategy in the discontinuous
-horizontal direction and a streamline-upwind Petrov-Galerkin (SUPG)
-method in the continuous vertical direction. The SUPG method replaces
-the finite element test function $\gamma$ with $\gamma + \tau
-\MM{u}\cdot\nabla\gamma$ where $\tau$ is a stabilisation parameter,
-although here we just require the vertical part of the gradient term. This
-has the effect of suppressing spurious oscillations associated with the
-continuous finite element approximation of the transport term. The
-stabilisation parameter $\tau$ depends on the spatial resolution and
-** and is set to be ** by default in Gusto, following \citet{}.
+required in the transport term is not globally defined. As described
+in \citet{yamazaki2017vertical}, a stable discretisation of this
+equation can be constructed by using a combination of the above
+upwinding strategy in the discontinuous horizontal direction and a
+streamline-upwind Petrov-Galerkin (SUPG) method
+\citep{brooks1982streamline} in the continuous vertical direction. The
+SUPG method replaces the finite element test function $\gamma$ with
+$\gamma + \tau \MM{u}\cdot\nabla\gamma$ where $\tau$ is a
+stabilisation parameter, although here we just require the vertical
+part of the gradient term. This has the effect of suppressing spurious
+oscillations associated with the continuous finite element
+approximation of the transport term. The stabilisation parameter
+$\tau$ depends is set to be $\Delta t/\sqrt{15}$ by default in Gusto,
+following \citet{raymond1976selective}.
 
 We first integrate by parts in each column $C$ of the mesh, taking the
-upwind value of the field, indicated by a $\tilde{}$, on the vertical
-facets of the mesh, denoted by $\Gamma_v$. Using the notation of the
-compressible Euler equations, this gives
+upwind value of the field $\tilde{\theta}$ on the vertical
+facets of the mesh, denoted by $\Gamma_v$, giving
 
@@
 \end{equation}
-Note that, as explained in \citet{Colin}, the test function should
-also be replaced in the time derivative term for consistency **
+Note that, as explained in \citet{cotter2023compatible}, the test
+function should also be replaced in the time derivative term for
+consistency.
 
@@
 a vector version of the scheme in section \ref{subsec: theta
-  transport} or, as shown in \citet{}, it can be replaced by the vector
+  transport} or, as in \citet{natale2018variational} and
+\citet{gibson2019compatible}, it can be replaced by the vector
 invariant form and then discretised. In Gusto we offer both options
@@
   \int_\Omega \MM{w}\cdot(\nabla\times\MM{u})\times\MM{u}
-  + \frac{1}{2}\MM{w}\cdot\nabla(\MM{u}\cdot\MM{u}) \diff x = \int_\Omega\nabla\times(\MM{u}\times\MM{w})\cdot\MM{u}\diff x - \int_{\Gamma}\jump{\MM{n}\times(\MM{u}\times\MM{w})}\cdot\tilde{\MM{u}} \diff S - \frac{1}{2}\int_\Omega\nabla\cdot\MM{w}\MM{u}\cdot\MM{u}\diff x, \quad \forall \MM{w} \in \mathbb{V}^{HDiv},
+  + \frac{1}{2}\MM{w}\cdot\nabla(\MM{u}\cdot\MM{u}) \diff x = \int_\Omega\nabla\times(\MM{u}\times\MM{w})\cdot\MM{u}\diff x - \int_{\Gamma}\jump{\MM{n}\times(\MM{u}\times\MM{w})}\cdot\tilde{\MM{u}} \diff S - \frac{1}{2}\int_\Omega\nabla\cdot\MM{w}\MM{u}\cdot\MM{u}\diff x, \quad \forall \MM{w} \in \mathbb{V}^{HDiv}.
 \end{equation}
-where $\tilde{\MM{u}}$ denotes the upwind value of $\MM{u}$ on the facet.
 
@@
 limiter as part of the transport scheme. These limiters are described
-in \ref{}.
+in \citet{bendall2020compatible}.
 
@@
 transport term is treated implicitly so that the timestep is not
-restricted by the vertical Courant number. We provide the option to
-split the transport terms into horizontal and vertical components, as
-described in \ref{}.
+restricted by the vertical Courant number (e.g. as in
+\citet{weller2013runge}). We provide the option to split the transport
+terms into horizontal and vertical components, as described in
+\citet{brown2025fast}.
 
@@
 
-\authorcontribution{TEXT} %% this section is mandatory
+\authorcontribution{JS and TMB conceptualisation, investigation, methodology, project administration, software, supervision, validation and writing.
+JS and CJC funding acquisition, TA, AB, NH, DW software and validation, JB, TG, JH-C, CW software  } %% this section is mandatory
 

### 2025-08-05 source/paper.tex
reorganise results and add table with test case details

--- 2025-08-05 source/paper.tex
@@
 grids of a spherical manifold will necessarily have singular points
-that lead to a degradation in parallel scaling. An example of this is
+that lead to a degredation in parallel scaling. An example of this is
 the commonly-used latitude-longitude grid where the singular points
@@
 % supercomputer architectures
-% are run on massively parallel supercomputers that require the co
+% are run on massivly parallel supercomputers that require the co
 % Spatial domain decomposition methods share the computation
@@
 range of different equation sets relevant for weather and climate
-modelling. In section \ref{sec: equations} we will outline three
+modelling. In section \ref{sec: governing} we will outline three
 different equation sets commonly used for modelling geophysical fluids
@@
 types of terms (e.g. pressure gradient, transport, diffusion) seen in
-section \ref{sec: equations}. In section \ref{sec: design} we describe
+section \ref{sec: governing}. In section \ref{sec: design} we describe
 the code structure of Gusto followed by a detailed example (section
@@
 \section{Governing Equations}
-\label{sec: equations}
+\label{sec: governing}
 The aim of this section is to introduce three governing equation sets
@@
 and climate models, along with enough detail about the finite element
-discretisations provided by Gusto to demonstrate the similarities
+discretiations provided by Gusto to demonstrate the similarities
 between them. These similarities guide the design of the code
@@
 \subsection{Transport terms}
-\label{sec: transport}
 Since the prognostic variables are discretised using different finite
@@
 \subsubsection{Transport of $\mathbb{V}^{L^2}$ fields}
-\label{subsec: DG transport}
 The shallow water and compressible Euler equation sets both contain a
@@
 discontinuous between elements. This means that the divergence term is
-not well defined. A stable discretisation of this equation can be
+not well defined. A stable discretiation of this equation can be
 constructed by integrating that term by parts and taking the upwind
@@
 \subsubsection{Transport of $\mathbb{V}_\theta$ fields}
-\label{subsec: theta transport}
 The compressible Boussinesq and compressible Euler equation sets both
@@
 horizontal direction and a streamline-upwind Petrov-Galerkin (SUPG)
-method in the continuous vertical direction. The SUPG method replaces
+method in the continuous vertical dirction. The SUPG method replaces
 the finite element test function $\gamma$ with $\gamma + \tau
@@
 although here we just require the vertical part of the gradient term. This
-has the effect of suppressing spurious oscillations associated with the
+has the effect of suppressing spurious oscillation associated with the
 continuous finite element approximation of the transport term. The
@@
 \begin{equation}
-  \int_\Omega \gamma\MM{u}\cdot\nabla\theta\diff x =\int_\Omega(\gamma+\tau w\gamma_z)\MM{u}\cdot\nabla\theta \diff x + \int_{\Gamma_v}\jump{(\gamma+\tau w\gamma_z) \MM{u}}\tilde{\theta}-\jump{(\gamma+\tau w\gamma_z) \MM{u}\theta}\diff S, \quad \forall \gamma \in \mathbb{V}_\theta.
+  \int_C \gamma\MM{u}\cdot\nabla\theta\diff x =\int_\Omega(\gamma+\tau w\gamma_z)\MM{u}\cdot\nabla\theta \diff x + \int_{\Gamma_v}\jump{(\gamma+\tau w\gamma_z) \MM{u}}\tilde{\theta}-\jump{(\gamma+\tau w\gamma_z) \MM{u}\theta}\diff S, \quad \forall \gamma \in \mathbb{V}_\theta.
 \end{equation}
@@
 \subsubsection{Transport of $\mathbb{V}^{HDiv}$ fields}
-All three of the equation sets described in section \ref{sec:
-  equations} include a prognostic equation for the velocity field
+All three of the equation sets described in section \ref{} include a
+prognostic equation for the velocity field
 $\MM{u}\in\mathbb{V}^{HDiv}$ that contains a nonlinear transport term
 $\MM{u}\cdot\nabla\MM{u}$. This term can be discretised directly using
-a vector version of the scheme in section \ref{subsec: theta
-  transport} or, as shown in \citet{}, it can be replaced by the vector
-invariant form and then discretised. In Gusto we offer both options
-and we present the vector invariant scheme below.
+a vector version of the scheme in section ** or, as in \citet{} it can
+be replaced by the vector invariant form and then discretised. In
+Gusto we offer both options and we present the vector invariant scheme
+below.
 
@@
 equation sets are each defined in a python class, making use of the
-common finite element forms for the transport and diffusion terms to
+common finite element forms for the transport and difffusion terms to
 set up the residual (defined in section \ref{}) in weak form. At this
@@
 
-\subsection{Example 1: IMEX-Euler applied to the 1d advection diffusion equation}
+\subsection{Example 1: IMEX-Euler applied to the 1d advection diffuction equation}
 
@@
 in the equation based on the presence or absence of certain
-labels. The full strength of FML becomes even more apparent when we
+labels. The full strength of FML becomes even more apparant when we
 exploit information stored in the label value. We use this in two
@@
 
-\subsection{Example 2: Sequential splitting for an idealised chemistry-transport problem}
-
-\subsection{Example 3: Quasi-Newton }
+\subsection{Example 2: Sequential splitting for an idealised chemistry-transport problem} \label{sec:seq splitting}
+
+\subsection{Example 3: Quasi-Newton } \label{sec:siqn}
 
@@
 The aim of this section is to present a selection of results that
-demonstrate the tools available in Gusto, such as the choice of grids,
+showcase the tools available in Gusto, such as the choice of grids,
 HDiv element families, order of finite element spaces, different
-spatial discretisation options, timestepping schemes and physics. The
-numerical properties of these configurations (convergence, stability,
-accuracy and efficiency) have either been demonstrated elsewhere, or
-will be investigated more thoroughly in future publications.
-
-To include:
-\begin{itemize}
-\item terminator toy (Tim)
-\item shallow water on different grids and with different HDiv
-  families (Nell) e.g. Williamson 5 cubed sphere PV and Galewsky
-  icosahedral sphere PV, high res (ref 6)
-\item convergence with different RK schemes / multistage schemes using
-  Williamson 2 or 6, L2 and Linf errors for D and u (2 plots, one L2,
-  one Linf) (Alex)
-\item vertical slice Euler - Straka, different orders, illustrates
-  recovery for diffusion (Tom)
-\item 3D moist bubble with rain (Tom/Alex)
-\item spherical gravity wave, Boussinesq (Jemma)
-\item spherical baroclinic wave, Euler - whatever people usually plot
-  (horizontal slice pressure contours, temperature shading) (Dan)
-\item Held Suarez??
-\end{itemize}
-
-\subsection{Terminator `toy' chemistry}
-This test combines transport and a `toy' chemistry model to
-demonstrate physics-dynamics coupling in Gusto (where here chemistry
-is the `physics').
+spatial discretisation options and timestepping schemes.
+As these options can be combined in so many ways, it is only possible to present results from a tiny fraction of the configurations.
+We have chosen to use a selection from the hierarchy of standard test cases that span the stages of development of dynamical cores.
+Although we do include some results here that demonstrate the numerical properties of certain configurations, the purpose of this section is not to provide a rigorous verification of any particular choice of discretisation.
+Such results are demonstrated elsewhere, for instance \citet{bendall2020compatible} or \citet{hartney2025exploring}.
+The settings used for each result from this section are detailed in Appendix \ref{sec:appendix}.
+
+\subsection{Terminator Toy chemistry}
+The first results use the ``Terminator Toy'' test of \citet{lauritzen2015terminator}, which involves the transport of two coupled tracers $X$ and $X_2$ on the surface of a sphere.
+The chemicals interact such that the sum $m_T=m_X+2m_{X_2}$ should be preserved, where $m_X$ and $m_{X_2}$ are the mixing ratios of the individual chemicals.
+When positivity or monotonicity limiters are applied to the numerical transport scheme, this may break the correlation between the two species which will be highlighted by this test.
+This test is also a demonstration of simple ``physics-dynamics coupling'', as the conversion term between the species can be handled separately from the transport terms. \\
+\\
+For this test, we use the DG upwind scheme described in Section \ref{sec:transport} to transport the species on an icosahedral sphere mesh, and apply the vertex-based limiter of \citet{kuzmin2010vertex}.
+The term describing the chemical reaction is separated from the transport term using a simple sequential splitting, as described in Section \ref{sec:seq splitting}. \\
+\\
+Figure \ref{fig:terminator} shows the mixing ratios of the two individual species at the initial time and following a single rotation around the sphere by a deformational flow.
+The change in the sum of the species is also displayed, with differences from zero remaining close to machine precision, which demonstrates that this transport scheme does preserve the correlation between the tracers.
+
+\begin{figure}[htp!]
+\centering
+\includegraphics[width=0.9\textwidth]{figures/terminator_toy.png}
+\caption{The evolution of the chemical species from the ``Terminator Toy'' test of \citet{lauritzen2015terminator}.
+The fields at the initial time are shown on the left, and at the final time ($t=12$ days) which follows a single revolution around the sphere by a deformational flow.
+The bottom row shows the change in the total $m_T=m_X+2m_{X_2}$ from the initial time, which remains very close to zero.}
+\label{fig:terminator}
+\end{figure}
 
@@
 \citet{williamson1992standard} provides a common starting point for
-testing new discretisations. This test suite is usually augmented by
+testing new discretisations. This test suite is often augmented by
 the unstable barotropic jet described in
-\citet{galewsky2004initial}. Here we present convergence results for a
-variety of timestepping schemes applied to the steady state
-geostrophic balanced flow (test case 2 from
-\citet{williamson1992standard}) and a demonstration of two different
-grids and HDiv element families for shallow water flow over a mountain
-(test case 5 from \citet{williamson1992standard}) and the unstable
-barotropic jet \citep{galewsky2004initial}.
-
-\subsubsection{Convergence results}
-
-\subsubsection{Grids and finite element families}
-
-\subsection{Compressible Boussinesq equations}
+\citet{galewsky2004initial}. \\
+\\
+Here we present two sets of convergence results for a variety of configurations measured using the steady state geostrophic balanced flow (test case 2 from
+\citet{williamson1992standard}) with the linear shallow water equations, as used in \citet{weller2012computational}.
+Figure \ref{fig:sw_convergence} shows the spatial convergence for a number of choices of finite element for the prognostic variables, using a Trapezium Rule discretisation (also known as Crank-Nicolson).
+It also shows the temporal convergence for a range of timestepping schemes using the 1st order Raviart-Thomas elements on quadrilateral cells. 
+\textcolor{red}{Once the test has been completed, come back to edit what we say here about convergence!} \\
+\\
+We show results from the unstable jet of \citet{galewsky2004initial} in Figure \ref{fig:galewsky}, using the non-linear shallow water equations.
+These results use the Semi-Implicit Quasi-Newton timestepper described in Section \ref{sec:siqn}, and compare the first-order RT elements on a cubed-sphere mesh with the first-order BDM elements on an icosahedral-sphere.
+
+\begin{figure}[htp!]
+\centering
+\includegraphics[width=0.9\textwidth]{}
+\caption{Convergence results using the linear shallow water equations with the steady geostrophic flow of \citet{williamson1992standard}.
+(Left) the spatial convergence of a number of orders and families of finite element, all using the same timestepper.
+(Right) the temporal convergence of different timesteppers, all using the same spatial discretisation.}
+\label{fig:sw_convergence}
+\end{figure}
+
+\begin{figure}[htp!]
+\centering
+\includegraphics[width=0.9\textwidth]{figures/galewsky_jet.png}
+\caption{A view of the relative vorticity around the North pole at day 6 of a simulation of the Galewsky jet test case.
+(Left) the RT1 elements on a cubed-sphere mesh.
+(Right) the BDM1 elements on an icosahedral-sphere mesh.
+Both cases use the Semi-Implicit Quasi-Newton timestepper.
+The edges of the panels of the cube and icosahedron are highlighted.}
+\label{fig:galewsky}
+\end{figure}
 
 \subsection{Compressible Euler equations}
-
-\subsubsection{Dense bubble in a vertical slice}
-
-\subsubsection{3D moist rising bubble}
-
-\subsubsection{Baroclinic wave on the sphere}
-
-\subsubsection{Held Suarez}
+We have also implemented many of the standard test cases for the non-hydrostatic compressible Euler equations that are used in modern weather and climate models.
+In this section we show results from a small selection of these, which demonstrate different domains in which these methods work.
+Throughout this section we use the Semi-Implicit Quasi-Newton timestepper.
+ \\
+\\
+The first test is the falling density current of \citet{straka1993numerical} in a vertical slice. This test involves the addition of diffusion to both the potential temperature and velocity equations.
+Figure \ref{fig:straka} shows the final potential temperature field for simulations using the lowest-order and next-to-lowest order finite elements on the vertical slice.
+The lowest-order methods require the transport and diffusion terms to use the ``recovered'' scheme of \citet{bendall2019recovered} to attain second-order accuracy. 
+\\
+\begin{figure}[htp!]
+\centering
+\includegraphics[width=0.8\textwidth]{figures/straka.png}
+\caption{The potential temperature field after 900 s from the falling density current test of \citet{straka1993numerical}.
+(Top) with the lowest-order finite elements, and (bottom) with the next-to-lowest-order finite elements.
+}
+\label{fig:straka}
+\end{figure}
+\\
+To demonstrate the use of moist physics parametrisations, we use a moist three-dimensional rising thermal simulation in an extruded doubly-periodic plane.
+The initial conditions are those of \citet{bendall2020compatible}, which is an extension of the dry case of \citet{kelly2012continuous} and \citet{melvin2019mixed}.
+A cross-sections of the final equivalent potential temperature field is shown in Figure \ref{fig:3d_bubble}.
+As in the two-dimensional test of \citet{bryan2002benchmark}, the atmosphere is saturated and cloudy everywhere, so that the moist processes are active in the whole domain.
+A simple saturation adjustment parametrisation is used, with the details described in \citet{bendall2020compatible}.
+\\
+\begin{figure}[htp!]
+\centering
+\includegraphics[width=0.48\textwidth]{figures/moist_3d_bubble.png}
+\caption{Vertical cross-section of the final equivalent potential temperature field from the three-dimensional moist rising thermal test.
+}
+\label{fig:3d_bubble}
+\end{figure}
+\\
+Our final test case is the deep baroclinic wave of \citet{ullrich2014proposed} on the sphere.
+An initial state in thermal wind balance is perturbed to produce a mid-latitude baroclinic instability, providing a flow that is typical of those found in numerical weather prediction.
+The surface pressure and temperature fields after 10 days are shown in Figure \ref{fig:baroclinic}, which compare well with those in \citet{melvin2024mixed} and demonstrate that our methods can replicate results from the literature.
+
+\begin{figure}[htp!]
+\centering
+\includegraphics[width=0.9\textwidth]{figures/baroclinic_wave_lonlat.png}
+\caption{The surface pressure and temperature after 10 days of a deep baroclinic wave simulation on the cubed-sphere.}
+\label{fig:baroclinic}
+\end{figure}
 
@@
 \appendix
-\section{}    %% Appendix A
+\section{Test Case Settings}  \label{sec:appendix}  %% Appendix A
+Here we summarise the configuration details for the test cases used to produce the results in Section \ref{sec: results}.
+
+\begin{sidewaystable*}[tbp!]
+\caption{Details of the configurations used for the test cases in Section \ref{sec: results}.}
+\begin{tabular}{l|ccccccc}
+\tophline
+Test Case & Equation Set & Timestepper & Finite Elements & \begin{tabular}{@{}c@{}}Horizontal \\ Mesh\end{tabular} & \begin{tabular}{@{}c@{}}Number \\ of  \ Layers\end{tabular} & Time step & Final Time \\
+\middlehline
+Terminator Toy & \begin{tabular}{@{}c@{}}Tracer transport \\ with chemistry\end{tabular} & Sequential Splitting & BDM1 & I16 & None & 450 s & 12 days \\
+Steady Geostrophic Flow & Linear Shallow Water & Multiple & BDM1 & I32 & None & ?? & 1 day \\
+Steady Geostrophic Flow & Linear Shallow Water & Trapezium Rule & Multiple & ?? & None & ?? & 1 day \\
+Unstable Jet & Shallow Water & SIQN & BDM1 and RT1 & I64 and C100 & None & 150 s & 6 days \\
+Density Current & Compressible Euler & SIQN & RT0 and RT1 & 100 m and 200 m & 64 and 32 & 1 s & 900 s \\
+Rising Thermal & Moist Compressible Euler & SIQN & RT1 & $\Delta x = \Delta y = \frac{10}{81}$ km & 81 & 2 s & 1000 s \\
+Baroclinic Wave & Compressible Euler & SIQN & RT1 & C32 & 15 & 900 s & 10 days
+\bottomhline
+\end{tabular}
+\belowtable{Notation: RT$x$ refers to the de Rham complex that uses the Raviart-Thomas elements on quadrilateral cells for the HDiv function space, and results in an $x$-th order polynomials for the $L_2$ space.
+Similarly, BDM$x$ refers to the Brezzi-Douglas-Marini de Rham complex on triangular cells.
+SIQN denotes the Semi-Implicit Quasi-Newton timestepper described in Section \ref{sec:siqn}.
+When describing horizontal meshes of the sphere, I$y$ and C$y$ correspond to icosehedral-sphere and cubed-sphere meshes with $y$ cells per edge of the icosahedron or cube panel.}% Table Footnotes
+\end{sidewaystable*}
 

### 2025-08-05 results/plotting/plot_baroclinic_fancy.py

@@ -1,0 +1,135 @@
+"""
+Plots the Baroclinic wave test case.
+
+This plots:
+(a) Surface Pressure , (b) Temperature
+"""
+from os.path import abspath, dirname
+from matplotlib.colors import ListedColormap
+import matplotlib.pyplot as plt
+import matplotlib.colors as colors
+from netCDF4 import Dataset
+import cartopy.crs as ccrs
+import numpy as np
+from tomplot import (
+    set_tomplot_style, tomplot_cmap, plot_contoured_field,
+    add_colorbar_ax, tomplot_field_title, tomplot_contours,
+    extract_gusto_coords, extract_gusto_field, reshape_gusto_data,
+    regrid_horizontal_slice
+)
+
+# ---------------------------------------------------------------------------- #
+# Directory for results and plots
+# ---------------------------------------------------------------------------- #
+# When copying this example these paths need editing, which will usually involve
+# removing the abspath part to set directory paths relative to this file
+results_dir = f'/media/thomas/1670402270400B47/results/baroclinic_wave'
+plot_stem = f'{abspath(dirname(__file__))}/../../source/figures/'
+
+# ---------------------------------------------------------------------------- #
+# Final plot details
+# ---------------------------------------------------------------------------- #
+field_names = ['Pressure_Vt', 'Temperature']
+slice_along = 'z'
+colours = ['PRGn', 'Reds']
+centre_titles = ['Surface pressure',  'Temperature']
+field_labels = [r'$P \ / $hPa', r'$T \ / $K']
+
+# We need to regrid onto lon-lat grid -- specify that here
+lon_1d = np.linspace(-180.0, 180.0, 80)
+lat_1d = np.linspace(-90, 90, 80)
+coords_lon, coords_lat = np.meshgrid(lon_1d, lat_1d, indexing='ij')
+
+spherical_centre = (120, 60)
+projection = ccrs.Orthographic(spherical_centre[0], spherical_centre[1])
+
+contours = [
+    np.linspace(94000, 102000, 17),
+    np.linspace(230, 330, 21)
+]
+remove_contours = [100000.0, None]
+cmap_rescales = ['top', None]
+
+# ---------------------------------------------------------------------------- #
+# General options
+# ---------------------------------------------------------------------------- #
+contour_method = 'contour'
+domain_limit = {'X' : (-180, 180), 'Y' : (-90, 90)}
+
+time_idx = 10*4
[…]

### 2025-08-05 results/plotting/plot_baroclinic_lonlat.py

@@ -1,0 +1,129 @@
+"""
+Plots the Baroclinic wave test case.
+
+This plots:
+(a) Surface Pressure , (b) Temperature
+"""
+from os.path import abspath, dirname
+from matplotlib.colors import ListedColormap
+import matplotlib.pyplot as plt
+import matplotlib.colors as colors
+from netCDF4 import Dataset
+import numpy as np
+from tomplot import (
+    set_tomplot_style, tomplot_cmap, plot_contoured_field,
+    add_colorbar_ax, tomplot_field_title, tomplot_contours,
+    extract_gusto_coords, extract_gusto_field, reshape_gusto_data,
+    area_restriction
+)
+
+# ---------------------------------------------------------------------------- #
+# Directory for results and plots
+# ---------------------------------------------------------------------------- #
+# When copying this example these paths need editing, which will usually involve
+# removing the abspath part to set directory paths relative to this file
+results_dir = f'/media/thomas/1670402270400B47/results/baroclinic_wave'
+plot_stem = f'{abspath(dirname(__file__))}/../../source/figures/'
+
+# ---------------------------------------------------------------------------- #
+# Final plot details
+# ---------------------------------------------------------------------------- #
+field_names = ['Pressure_Vt', 'Temperature']
+slice_along = 'z'
+colour_schemes = ['PiYG', 'RdBu_r']
+centre_titles = ['Surface pressure',  'Temperature']
+field_labels = [r'$P \ / $hPa', r'$T \ / $K']
+
+contours = [
+    np.linspace(94000, 102000, 17),
+    np.linspace(230, 330, 11)
+]
+norms = [colors.TwoSlopeNorm(vcenter=100000, vmin=min(contours[0]), vmax=max(contours[0])), None]
+cmaps = ['PRGn', 'Reds']
+
+# ---------------------------------------------------------------------------- #
+# General options
+# ---------------------------------------------------------------------------- #
+contour_method = 'tricontour'
+domain_limit = {'X' : (-180, 180), 'Y' : (-90, 90)}
+
+time_idx = 10*4
+# Things that are likely the same for all plots --------------------------------
+set_tomplot_style()
+level = 0
+
+y_ticks = [-90, 90]
+y_tick_labels = ['90S', '90N']
+xticks = [-180, 180]
+xtick_labels = ['180W', '180E']
+
[…]

### 2025-08-05 results/plotting/plot_williamson_5.py

@@ -1,99 +1,0 @@
-"""
-Plot from the Terminator Toy test case.
-"""
-from os.path import abspath, dirname
-import cartopy.crs as ccrs
-import matplotlib.pyplot as plt
-import numpy as np
-from netCDF4 import Dataset
-from tomplot import (
-    set_tomplot_style, tomplot_cmap, plot_contoured_field, add_colorbar_fig,
-    tomplot_field_title, extract_gusto_coords, extract_gusto_field,
-)
-
-results_dirs = ['williamson_5_ssprk3', 'williamson_5_semiimplicit',
-                'williamson_5_ssprk3', 'williamson_5_semiimplicit']
-
-# ---------------------------------------------------------------------------- #
-# Directory for results and plots
-# ---------------------------------------------------------------------------- #
-results_stem = f'{abspath(dirname(__file__))}/../data/'  # where it should eventually be
-results_stem = f'{abspath(dirname(__file__))}/../results/'  # where it currently is
-plot_stem = f'{abspath(dirname(__file__))}/../figures/williamson_5'
-
-# ---------------------------------------------------------------------------- #
-# Plot details
-# ---------------------------------------------------------------------------- #
-subplots_x = 2
-subplots_y = 2
-field_name = 'RelativeVorticity'
-time_idxs = [3, 3, -1, -1]
-subtitles = ['Explicit, 15 days', 'Semi-Implicit, 15 days',
-             'Explicit, 50 days', 'Semi-Implicit, 50 days']
-
-# ---------------------------------------------------------------------------- #
-# General options
-# ---------------------------------------------------------------------------- #
-
-projection = ccrs.Robinson()
-contours = np.linspace(-1e-4, 1e-4, 17)
-colour_scheme = 'RdBu_r'
-field_label = r'$\zeta$ (s$^{-1}$)'
-contour_method = 'tricontour'
-
-# Things that are likely the same for all plots --------------------------------
-set_tomplot_style()
-
-# ---------------------------------------------------------------------------- #
-# PLOTTING
-# ---------------------------------------------------------------------------- #
-
-fig = plt.figure(figsize=(15, 10))
-
-# Save filled contour data for adding colourbars at end
-all_cf = []
-
-for i, (subtitle, results_dir, time_idx) in enumerate(zip(subtitles, results_dirs, time_idxs)):
-
-    results_file_name = f'{results_stem}{results_dir}/field_output.nc'
[…]

### 2025-08-05 results/test_case_scripts/williamson_5.py

@@ -1,185 +1,0 @@
-"""
-Test Case 5 (flow over a mountain) of Williamson et al, 1992:
-``A standard test set for numerical approximations to the shallow water
-equations in spherical geometry'', JCP.
-"""
-
-from argparse import ArgumentParser, ArgumentDefaultsHelpFormatter
-from firedrake import (
-    SpatialCoordinate, as_vector, pi, sqrt, min_value, Function
-)
-from gusto import (
-    Domain, IO, OutputParameters, SemiImplicitQuasiNewton, SSPRK3, DGUpwind,
-    ShallowWaterParameters, ShallowWaterEquations, Sum,
-    lonlatr_from_xyz, GeneralCubedSphereMesh, ZonalComponent,
-    MeridionalComponent, RelativeVorticity, Timestepper
-)
-
-williamson_5_defaults = {
-    'ncells_per_edge': 64,     # number of cells per cubed sphere edge
-    'dt': 300.0,               # 5 minutes
-    'tmax': 50.*24.*60.*60.,   # 50 days
-    'dumpfreq': 1440,          # once per 5 days with default options
-    'dirname': 'williamson_5',
-    'semi_implicit': False
-}
-
-
-def williamson_5(
-        ncells_per_edge=williamson_5_defaults['ncells_per_edge'],
-        dt=williamson_5_defaults['dt'],
-        tmax=williamson_5_defaults['tmax'],
-        dumpfreq=williamson_5_defaults['dumpfreq'],
-        dirname=williamson_5_defaults['dirname'],
-        semi_implicit=williamson_5_defaults['semi_implicit']
-):
-
-    # ------------------------------------------------------------------------ #
-    # Parameters for test case
-    # ------------------------------------------------------------------------ #
-
-    radius = 6371220.           # planetary radius (m)
-    mean_depth = 5960           # reference depth (m)
-    g = 9.80616                 # acceleration due to gravity (m/s^2)
-    u_max = 20.                 # max amplitude of the zonal wind (m/s)
-    mountain_height = 2000.     # height of mountain (m)
-    R0 = pi/9.                  # radius of mountain (rad)
-    lamda_c = -pi/2.            # longitudinal centre of mountain (rad)
-    phi_c = pi/6.               # latitudinal centre of mountain (rad)
-
-    # ------------------------------------------------------------------------ #
-    # Our settings for this set up
-    # ------------------------------------------------------------------------ #
-
[…]

### 2025-08-06 source/paper.tex
add pseudocode for timesteppers

--- 2025-08-05 source/paper.tex
@@
 \subsection{Viscous and diffusive terms}
-
+\label{sec: diffusion}
 For the simulations presented here we do not require viscous or
@@
 common finite element forms for the transport and diffusion terms to
-set up the residual (defined in section \ref{}) in weak form. At this
-point, no particular discretisation of the transport or diffusion
-forms is specified. This happens in the \texttt{SpatialMethod} class
-which can be used to apply the methods described in section \ref{} to
-the relevant terms. Any physics source terms, such as **, are added by
-calling the \texttt{Physics} class that defines them. The
-\texttt{TimeStepper} class contains the definition of the timestepping
-procedure and the \texttt{run} method that loops over the
-timesteps. It also sets up diagnostic fields and calls the IO as
-required. The simplest timeloop class will just apply one time
-discretisation method to the equation set but it is common in weather
-and climate prediction models to split the different types of terms
-and apply a time discretisation separately to each type of term. For
-example, the transport and forcing terms might be solved for
+set up the residual (i.e. the left hand side of the equations defined
+in section \ref{sec: equations}) in weak form. At this point, no
+particular discretisation of the transport or diffusion forms is
+specified. This happens in the \texttt{SpatialMethod} class which can
+be used to apply the methods described in sections \ref{sec:
+  transport} and \ref{sec: diffusion} to the relevant terms. Any
+physics source terms, such as those required for moist processes or
+tubulence parameterisation, are added by calling the \texttt{Physics}
+class that defines them. The \texttt{TimeStepper} class contains the
+definition of the timestepping procedure and the \texttt{run} method
+that loops over the timesteps. It also sets up diagnostic fields and
+calls the IO as required. The simplest timeloop class will just apply
+one time discretisation method to the equation set but it is common in
+weather and climate prediction models to split the different types of
+terms and apply a time discretisation separately to each type of
+term. For example, the transport and forcing terms might be solved for
 separately, such as in **, which enables explicit transport schemes to
@@
 `fast' timescales evaluated more frequently and `slower' physics
-processes evaluated just once at the end of the timestep. The choice
-of where to evaluate the physics processes in the iteration is also
-based on the computational cost and the best way to do this is still
-an open research question (**ref).
-
-\begin{itemize}
-\item what about initialisation and diagnostics?
-\end{itemize}
+processes evaluated just once at the end of the timestep. Since the
+test cases presented later only use simple moist physics
+parameterisations, we just include the option to solve for these at
+the end of the timestep.
 
 \begin{algorithm} 
-  \caption{Pseudocode for SIQN timestep loop}\label{alg:SIQN}
-  \begin{algorithmic}
-    \STATE
+  \caption{Basic timestepping scheme}
+  \label{alg: basic}
+  \begin{algorithmic}[1]
+    \STATE Replace transport and diffusion terms with specified spatial discretisation scheme.
+    \STATE Setup IO and diagnostics and write out initial conditions.
+    \WHILE{$t < t_{max}$}
+      \STATE Apply timestepping scheme to equation set: $\chi^{n+1} = \mathcal{T}(\chi^n, \chi^{n+1})$
+      \STATE Advance timestep: $\chi^n = \chi^{n+1}$
+      \IF{ndt \% nts == 0}
+        \STATE Update timeseries diagnostics file.
+      \ENDIF
+      \IF{ndt \% ndiag == 0}
+        \STATE Compute and write out diagnostic fields.
+      \ENDIF
+      \IF{ndt \% nchk == 0}
+        \STATE Write out checkpoint file.
+      \ENDIF
+    \ENDWHILE
   \end{algorithmic} 
@@
 
-\begin{itemize}
-\item generic example(s) - no FML, just pseudocode for basic
-  timestepper, SIQN and split physics
-\item lead in to next section... facilitated by FML i.e. timesteppers don't care what equations they're given because forms are labelled
-\end{itemize}
+\begin{algorithm} 
+  \caption{Split physics timestepping scheme}
+  \label{alg: split physics}
+  \begin{algorithmic}[1]
+    \STATE Dynamics step: $\chi^{n+1} = \mathcal{D}(\chi^n)$
+    \FOR {$\mathcal{P}$ in physics schemes}
+      \STATE Apply physics scheme: $\chi^{n+1} = \chi^{n+1} + \mathcal{P}(\chi^{n+1})$
+    \ENDFOR
+    \STATE Advance timestep: $\chi^n = \chi^{n+1}$
+  \end{algorithmic} 
+\end{algorithm}
+
+\begin{algorithm} 
+  \caption{Semi-implicit quasi-Newton timestepping scheme}
+  \label{alg: SIQN}
+  \begin{algorithmic}[1]
+    \STATE Set: $\chi^{n+1} = \chi^{n}$
+    \STATE Explicit forcing: $\chi_{f, e} = \chi^n + \frac{1}{2}\Delta t \mathcal{F}(\chi^n)$
+    \FOR {$OUTER$}
+      \STATE Update advecting velocity: $\bar{\MM{u}} = \frac{1}{2} (\MM{u}^{n+1} + \MM{u}^n)$
+      \STATE Explicit transport: $\chi_T = \mathcal{A}_{\bar{\MM{u}}}(\chi_{f,e})$
+      \FOR {$INNER$}
+        \STATE Implicit forcing: $\chi_{f,i} = \chi_T + \frac{1}{2}\Delta t \mathcal{F}(\chi^{n+1})$
+        \STATE Write residual: $\chi_{r} = \chi_{f,i} - \chi^{n+1}$
+        \STATE Solve: $\mathcal{S}(\chi') = \chi_{r}$ for $\chi'$
+        \STATE Increment: $\chi^{n+1} = \chi^{n+1} + \chi'$
+      \ENDFOR
+    \ENDFOR
+    \STATE Explicit physics: $\chi^{n+1} = \mathcal{P}(\chi^{n+1})$
+    \STATE Advance timestep: $\chi^n = \chi^{n+1}$
+  \end{algorithmic} 
+\end{algorithm}
 

### 2025-08-06 source/paper.tex

--- 2025-08-05 source/paper.tex
@@
 equation sets are each defined in a python class, making use of the
-common finite element forms for the transport and difffusion terms to
-set up the residual (defined in section \ref{}) in weak form. At this
-point, no particular discretisation of the transport or diffusion
-forms is specified. This happens in the \texttt{SpatialMethod} class
-which can be used to apply the methods described in section \ref{} to
-the relevant terms. Any physics source terms, such as **, are added by
-calling the \texttt{Physics} class that defines them. The
-\texttt{TimeStepper} class contains the definition of the timestepping
-procedure and the \texttt{run} method that loops over the
-timesteps. It also sets up diagnostic fields and calls the IO as
-required. The simplest timeloop class will just apply one time
-discretisation method to the equation set but it is common in weather
-and climate prediction models to split the different types of terms
-and apply a time discretisation separately to each type of term. For
-example, the transport and forcing terms might be solved for
+common finite element forms for the transport and diffusion terms to
+set up the residual (i.e. the left hand side of the equations defined
+in section \ref{sec: equations}) in weak form. At this point, no
+particular discretisation of the transport or diffusion forms is
+specified. This happens in the \texttt{SpatialMethod} class which can
+be used to apply the methods described in sections \ref{sec:
+  transport} and \ref{sec: diffusion} to the relevant terms. Any
+physics source terms, such as those required for moist processes or
+tubulence parameterisation, are added by calling the \texttt{Physics}
+class that defines them. The \texttt{TimeStepper} class contains the
+definition of the timestepping procedure and the \texttt{run} method
+that loops over the timesteps. It also sets up diagnostic fields and
+calls the IO as required. The simplest timeloop class will just apply
+one time discretisation method to the equation set but it is common in
+weather and climate prediction models to split the different types of
+terms and apply a time discretisation separately to each type of
+term. For example, the transport and forcing terms might be solved for
 separately, such as in **, which enables explicit transport schemes to
@@
 \centering
-\includegraphics[width=0.8\textwidth]{figures/straka.png}
+\includegraphics[width=0.6\textwidth]{figures/straka.png}
 \caption{The potential temperature field after 900 s from the falling density current test of \citet{straka1993numerical}.

### 2025-08-06 source/paper.tex
fix issues from merge

@@
 the integration by parts because these vanish on interior element
-boundaries since the normal component of $\MM{w}$ is continuous across
-element facets and on exterior boundaries we set no normal flow
+boundaries (since the normal component of $\MM{w}$ is continuous across
+element facets) and on exterior boundaries we set no normal flow
 boundary conditions. However, the pressure gradient term in the
@@
 $\theta$, and therefore $\Pi$, are discontinuous. Here we have chosen,
-as in \citet{}, to take the average value of $\Pi$ across the facets,
-$\avg{\Pi} = (\Pi^++\Pi^-)/2$, where the $+$ and $-$ indicate each
-side of the facet. The brackets $\jump{q} = q^+-q^-$ denote the jump
-in $q$ across the facet and the integral over the horizontal facets
-vanishes since both $\theta$ and $\MM{w}\cdot\MM{n}$ are continuous
-across these facets so the jump is zero.
+as in \citet{natale2016compatible}, to take the average value of $\Pi$
+across the facets, $\avg{\Pi} = (\Pi^++\Pi^-)/2$, where the $+$ and
+$-$ indicate each side of the facet. The brackets $\jump{q} = q^+-q^-$
+denote the jump in $q$ across the facet and the integral over the
+horizontal facets vanishes since both $\theta$ and $\MM{w}\cdot\MM{n}$
+are continuous across these facets so the jump is zero.
 
 \subsection{Transport terms}
+\label{sec: transport}
 Since the prognostic variables are discretised using different finite
@@
 discontinuous between elements. This means that the divergence term is
-not well defined. A stable discretiation of this equation can be
+not globally defined. A stable discretisation of this equation can be
 constructed by integrating that term by parts and taking the upwind
@@
 \begin{equation}
-  \int_e \phi\nabla\cdot(\rho\MM{u}) \diff x = -\int_e \nabla\phi \cdot (\rho\MM{u}) \diff x + \int_{\partial e} \phi\tilde{\rho}\MM{u}\cdot\MM{n} \diff S, \quad \forall \phi \in \mathbb{V}^{L^2}.
+  \int_e \phi\nabla\cdot(\rho\MM{u}) \diff x = -\int_e \nabla\phi \cdot (\rho\MM{u}) \diff x + \int_{\partial e} \phi\hat{\rho}\MM{u}\cdot\MM{n} \diff S, \quad \forall \phi \in \mathbb{V}^{L^2}.
 \end{equation}
-Summing over the domain gives
+where $\hat{\rho}$ denotes the value of $\rho$ on the element
+boundary. Summing over the domain gives
 \begin{equation}
@@
 facet and the brackets $\jump{q} = q^+-q^-$ denote the jump in $q$
-across the facet.
+across the facet, where the $+$ and $-$ indicate each side of the
+facet.
 
 \subsubsection{Transport of $\mathbb{V}_\theta$ fields}
+\label{subsec: theta transport}
 The compressible Boussinesq and compressible Euler equation sets both
@@
 horizontal direction. This means that the gradient of the field
-required in the transport term is not well defined. A stable
-discretisation of this equation can be constructed by using a
-combination of the above upwinding strategy in the discontinuous
-horizontal direction and a streamline-upwind Petrov-Galerkin (SUPG)
-method in the continuous vertical dirction. The SUPG method replaces
-the finite element test function $\gamma$ with $\gamma + \tau
-\MM{u}\cdot\nabla\gamma$ where $\tau$ is a stabilisation parameter,
-although here we just require the vertical part of the gradient term. This
-has the effect of suppressing spurious oscillation associated with the
-continuous finite element approximation of the transport term. The
-stabilisation parameter $\tau$ depends on the spatial resolution and
-** and is set to be ** by default in Gusto, following \citet{}.
+required in the transport term is not well defined. As described in
+\citet{yamazaki2017vertical}, a stable discretisation of this equation
+can be constructed by using a combination of the above upwinding
+strategy in the discontinuous horizontal direction and a
+streamline-upwind Petrov-Galerkin (SUPG) method
+\citep{brooks1982streamline} in the continuous vertical dirction. The
+SUPG method replaces the finite element test function $\gamma$ with
+$\gamma + \tau \MM{u}\cdot\nabla\gamma$ where $\tau$ is a
+stabilisation parameter, although here we just require the vertical
+part of the gradient term. This has the effect of suppressing spurious
+oscillations associated with the continuous finite element
+approximation of the transport term. The stabilisation parameter
+$\tau$ is set to be $\Delta t/\sqrt{15}$ by default in Gusto,
+following \citet{raymond1976selective}.
 
 We first integrate by parts in each column $C$ of the mesh, taking the
-upwind value of the field, indicated by a $\tilde{}$, on the vertical
-facets of the mesh, denoted by $\Gamma_v$. Using the notation of the
-compressible Euler equations, this gives
+upwind value of the transported field $\tilde{\theta}$ on the vertical
+facets of the mesh, denoted by $\Gamma_v$, giving
 
@@
 \begin{equation}
-  \int_C \gamma\MM{u}\cdot\nabla\theta\diff x =\int_\Omega(\gamma+\tau w\gamma_z)\MM{u}\cdot\nabla\theta \diff x + \int_{\Gamma_v}\jump{(\gamma+\tau w\gamma_z) \MM{u}}\tilde{\theta}-\jump{(\gamma+\tau w\gamma_z) \MM{u}\theta}\diff S, \quad \forall \gamma \in \mathbb{V}_\theta.
+  \int_\Omega \gamma\MM{u}\cdot\nabla\theta\diff x =\int_\Omega(\gamma+\tau w\gamma_z)\MM{u}\cdot\nabla\theta \diff x + \int_{\Gamma_v}\jump{(\gamma+\tau w\gamma_z) \MM{u}}\tilde{\theta}-\jump{(\gamma+\tau w\gamma_z) \MM{u}\theta}\diff S, \quad \forall \gamma \in \mathbb{V}_\theta.
 \end{equation}
-Note that, as explained in \citet{Colin}, the test function should
-also be replaced in the time derivative term for consistency **
+Note that, as explained in \citet{cotter2023compatible}, the test
+function should also be replaced in the time derivative term for
+consistency.
 
 \subsubsection{Transport of $\mathbb{V}^{HDiv}$ fields}
-All three of the equation sets described in section \ref{} include a
-prognostic equation for the velocity field
+All three of the equation sets described in section \ref{sec:
+  equations} include a prognostic equation for the velocity field
 $\MM{u}\in\mathbb{V}^{HDiv}$ that contains a nonlinear transport term
 $\MM{u}\cdot\nabla\MM{u}$. This term can be discretised directly using
-a vector version of the scheme in section ** or, as in \citet{} it can
-be replaced by the vector invariant form and then discretised. In
-Gusto we offer both options and we present the vector invariant scheme
-below.
+a vector version of the scheme in section \ref{subsec: theta
+  transport} or, as in \citet{natale2018variational} and
+\citet{gibson2019compatible}, it can be replaced by the vector
+invariant form and then discretised. In Gusto we offer both options
+and we present the vector invariant scheme below.
+
+The starting point for this discretisation is to replace the
+$\MM{u}\cdot\nabla\MM{u}$ term with $(\nabla\times\MM{u})\times\MM{u}
++ \frac{1}{2}\nabla(\MM{u}\cdot\MM{u})$. Now, multiplying by a test
+function and integrating by parts over an element $e$ with boundary
+$\partial e$ gives
+\begin{equation}
+  \int_e \MM{w}\cdot(\nabla\times\MM{u})\times\MM{u}
+  + \frac{1}{2}\MM{w}\cdot\nabla(\MM{u}\cdot\MM{u}) \diff x = \int_e\nabla\times(\MM{u}\times\MM{w})\cdot\MM{u}\diff x - \int_{\partial e}\MM{n}\times(\MM{u}\times\MM{w})\cdot\MM{u} \diff S - \frac{1}{2}\int_e\nabla\cdot\MM{w}\MM{u}\cdot\MM{u}\diff x, \quad \forall \MM{w} \in \mathbb{V}^{HDiv}.
+\end{equation}
+Sunming over the domain and taking the upwind value of $\MM{u}$,
+denoted by $\tilde{\MM{u}}$, on the element boundaries gives
+\begin{equation}
+  \int_\Omega \MM{w}\cdot(\nabla\times\MM{u})\times\MM{u}
+  + \frac{1}{2}\MM{w}\cdot\nabla(\MM{u}\cdot\MM{u}) \diff x = \int_\Omega\nabla\times(\MM{u}\times\MM{w})\cdot\MM{u}\diff x - \int_{\Gamma}\jump{\MM{n}\times(\MM{u}\times\MM{w})}\cdot\tilde{\MM{u}} \diff S - \frac{1}{2}\int_\Omega\nabla\cdot\MM{w}\MM{u}\cdot\MM{u}\diff x, \quad \forall \MM{w} \in \mathbb{V}^{HDiv}.
+\end{equation}
 
 The schemes described above do not preserve monotonicity unlike the
-continuous equation, generating spurious over and undershoots. When
+continuous equation, generating spurious overshoots and undershoots. When
 monotonicity is important, for example, to avoid generating spurious
@@
 
-\authorcontribution{TEXT} %% this section is mandatory
+\authorcontribution{JS and TMB conceptualisation, investigation, methodology, project administration, software, supervision, validation and writing.
+JS and CJC funding acquisition, CJC conceptualisation, investigation, methodology, project administration, software, supervision, TA, AB, NH, DW software and validation, JB, TG, JH-C software.} %% this section is mandatory
 

### 2025-08-06 source/paper.tex

@@
 the integration by parts because these vanish on interior element
-boundaries (since the normal component of $\MM{w}$ is continuous across
-element facets) and on exterior boundaries we set no normal flow
-boundary conditions. However, the pressure gradient term in the
+boundaries (since the normal component of $\MM{w}$ is continuous
+across element facets) and on exterior boundaries we set no normal
+flow boundary conditions. However, the pressure gradient term in the
 compressible Euler equations, $c_p\theta\nabla\Pi$, is nonlinear and
@@
 discontinuous in the horizontal direction. Following
-\citet{natale2016compatible} we integrate by parts in each element $e$,
-giving
+\citet{natale2016compatible} we integrate by parts in each element
+$e$, giving
 
@@
 \begin{equation}
-  \int_e \phi\nabla\cdot(\rho\MM{u}) \diff x = -\int_e \nabla\phi \cdot (\rho\MM{u}) \diff x + \int_{\partial e} \phi\hat{\rho}\MM{u}\cdot\MM{n} \diff S, \quad \forall \phi \in \mathbb{V}^{L^2}.
+  \int_e \phi\nabla\cdot(\rho\MM{u}) \diff x = -\int_e \nabla\phi \cdot (\rho\MM{u}) \diff x + \int_{\partial e} \phi\hat{\rho}\MM{u}\cdot\MM{n} \diff S, \quad \forall \phi \in \mathbb{V}^{L^2},
 \end{equation}
@@
 horizontal direction. This means that the gradient of the field
-required in the transport term is not well defined. As described in
-\citet{yamazaki2017vertical}, a stable discretisation of this equation
-can be constructed by using a combination of the above upwinding
-strategy in the discontinuous horizontal direction and a
+required in the transport term is not globally defined. As described
+in \citet{yamazaki2017vertical}, a stable discretisation of this
+equation can be constructed by using a combination of the above
+upwinding strategy in the discontinuous horizontal direction and a
 streamline-upwind Petrov-Galerkin (SUPG) method
-\citep{brooks1982streamline} in the continuous vertical dirction. The
+\citep{brooks1982streamline} in the continuous vertical direction. The
 SUPG method replaces the finite element test function $\gamma$ with

### 2025-08-06 source/paper.tex
small changes to compile and fix referencing

@@
 Euler equation set. In addition, some test cases require viscous and
-diffusive terms to generate converged solutions \ref{}. In the
-following subsections we illustrate a non-exhaustive selection of the
-options available in Gusto.
+diffusive terms to generate converged solutions (for example
+\citet{straka1993numerical}). In the following subsections we
+illustrate a non-exhaustive selection of the options available in
+Gusto.
 
@@
 \\
-For this test, we use the DG upwind scheme described in Section \ref{sec:transport} to transport the species on an icosahedral sphere mesh, and apply the vertex-based limiter of \citet{kuzmin2010vertex}.
+For this test, we use the DG upwind scheme described in Section \ref{subsec: DG transport} to transport the species on an icosahedral sphere mesh, and apply the vertex-based limiter of \citet{kuzmin2010vertex}.
 The term describing the chemical reaction is separated from the transport term using a simple sequential splitting, as described in Section \ref{sec:seq splitting}. \\
@@
 Rising Thermal & Moist Compressible Euler & SIQN & RT1 & $\Delta x = \Delta y = \frac{10}{81}$ km & 81 & 2 s & 1000 s \\
-Baroclinic Wave & Compressible Euler & SIQN & RT1 & C32 & 15 & 900 s & 10 days
+Baroclinic Wave & Compressible Euler & SIQN & RT1 & C32 & 15 & 900 s & 10 days \\
 \bottomhline

### 2025-08-06 source/paper.tex
finish viscous / diffusion section

@@
 diffusive terms for stability, but sometimes they are required to
-produce converged solutions so that refining the grid does not
-lead to smaller and smaller scale motions. Here we present the symmetric
-interior penalty discretisation method used for these terms in Gusto.
+produce converged solutions so that refining the grid does not lead to
+smaller and smaller scale motions. Viscous or diffusive terms involve
+the laplacian of the prognostic field, which in all cases is not
+globally defined for the finite element function spaces we use in this
+paper. Here we present the symmetric interior penalty discretisation
+method \citep{cockburn2007note} applied to the viscous term that we
+add to the prognostic velocity equation when required. This method can
+be adapted to apply to scalar fields when required.
+
+First we multiply the term $\nu\nabla^2\MM{u}$ by a test function
+$\MM{w}\in\mathbb{V}^{HDiv}$, where $\nu$ is the viscosity
+coefficient, assumed to be constant. Integrating by parts introduces
+surface terms which, after applying the symmetric interior penalty
+method \citep{cockburn2007note} become:
 
 \begin{equation}
-  \nu\int_\Omega\MM{w}\cdot\nabla^2\MM{u}\diff x = \nu\int_\Omega\nabla\MM{w}:\nabla\MM{u}\diff x
+ \nu\int_\Omega\nabla\MM{w}:\nabla\MM{u}\diff x - \int_\Gamma\jump{\MM{v}\otimes\MM{n}}:\{\nabla\MM{u}\}\diff S - \int_\Gamma\jump{\MM{u}\otimes\MM{n}}:\{\nabla\MM{v}\}\diff S + \mu\int_\Gamma \jump{\MM{u}\otimes\MM{n}}:\jump{\MM{v}\otimes\MM{n}}\diff S,
 \end{equation}
+where $:$ denotes double contraction, $\avg{q} = (q^++q^-)/2$ is the
+average value of $q$ across the facet, $\jump{q} = q^+-q^-$ is the
+jump in $q$ across the facet (where in both cases the $+$ and $-$
+indicate each side of the facet), and $\mu$ is the penalty parameter,
+typically chosen to be $10/h$ where $h$ is a measure of the gridsize.
 

### 2025-08-07 source/paper.tex
almost finished abstract

@@
 \begin{abstract}
-TEXT
+Gusto is a python code library for building dynamical cores that are
+the heart of weather and climate prediction models. Gusto is built on
+top of Firedrake which provides the software infrastructure required
+to solve partial differential equations using the finite element
+method. The compressible Euler equations solved by the dynamical core
+have a particular mathematical structure that should be preserved by
+the spatial discretisation. Recent research has shown that one way to
+do this is using compatible finite element methods which also have the
+advantage that this structure is preserved even on non-orthogonal
+grids, enabling better parallel scaling through spatial domain
+decomposition. This was the motivation for the development of the Met
+Office new dynamical core, LFRic. However, there are different
+families of compatible finite element method that can be defined on
+different grids, such as the cubed sphere or icosahedral sphere
+grids. For each family the different prognostic fields are discretised
+using finite element spaces that have different orders and
+interelement continuity. These function spaces can be difficult to
+construct, in particular on the sphere where the Piola transform is
+required to ensure that the horizontal wind vector is tangential to
+the sphere. In addition, when simulating flows that have a much larger
+horizontal scale than vertical scale, it is advantageous to have
+meshes that are aligned in vertical columns; Firedrake provides such
+meshes and Gusto provides the correct finite element families for the
+compatible finite element method and a range of algorithms for stable
+and accurate timestepping fields discretised on these finite element
+spaces.
+
+Gusto is intended as a tool for rapid prototyping of new algorithms,
+based on compatible finite element methods, for dynamical cores. In
+this context, the development of new algorithms typically follows a
+path where they are tested on simpler equation sets such as the
+shallow water equations or the Boussinesq equations. These equation
+sets share important features of the mathematical structure of the
+compressible Euler equation set and so provide a good indication of
+whether new algorithms will work in a dynamical core, but they provide
+a simpler setting for investigating the properties of the
+algorithms. Therefore Gusto provides compatible finite element spatial
+discretisations of these equation sets and then the timestepping
+algorithms are written in such a way that they can be applied to any
+of these equation sets. ** write a bit more about what this enables?
+** talk about physics?
+
+In this paper we describe the algorithms implemented in Gusto and
+demonstrate their accuracy (and efficiency?) when applied to the
+hierarchy of test problems commonly used in the numerical weather
+prediction community.
 \end{abstract}
@@
 smaller and smaller scale motions. Viscous or diffusive terms involve
-the laplacian of the prognostic field, which in all cases is not
+the laplacian of the prognostic field, which for all fields is not
 globally defined for the finite element function spaces we use in this
-paper. Here we present the symmetric interior penalty discretisation
-method \citep{cockburn2007note} applied to the viscous term that we
-add to the prognostic velocity equation when required. This method can
-be adapted to apply to scalar fields when required.
+paper. In Gusto we use the symmetric interior penalty discretisation
+method \citep{cockburn2007note} which we descirbe below applied to the
+viscous term added to the prognostic velocity equation when
+required. This method can be adapted to apply to scalar fields.
 

### 2025-08-07 source/paper.tex
some description of the timestepping algorithms

@@
 terms and apply a time discretisation separately to each type of
-term. For example, the transport and forcing terms might be solved for
-separately, such as in **, which enables explicit transport schemes to
-be applied to each field. ** something about SIQN structure. Terms
-involved in physics parameterisations are evaluated in the appropriate
-part of the iteration loop, with physics processes that occur on
-`fast' timescales evaluated more frequently and `slower' physics
-processes evaluated just once at the end of the timestep. Since the
-test cases presented later only use simple moist physics
-parameterisations, we just include the option to solve for these at
-the end of the timestep.
+term. For example, in \citet{} the transport and forcing terms are
+solved separately, enabling subcycling of transport schemes and a
+larger stable Courant number. Terms involved in physics
+parameterisations are evaluated in the appropriate part of the
+iteration loop, with physics processes that occur on `fast' timescales
+evaluated more frequently and `slower' physics processes evaluated
+just once at the end of the timestep. Since the test cases presented
+later only use simple moist physics parameterisations, we only
+describe the option to solve for these at the end of the timestep.
+
+In Algorithm \ref{alg: basic} we show the basic structure of the
+timestepping loop in Gusto. The first step is to modify the equation
+residual to apply the spatial discretisation (for example, the DG
+upwind or SUPG discretisations described in section \ref{sec:
+  transport}). Before entering the timestepping loop, we set up the
+IO, including any diagnostics, and write out the initial
+conditions. In this basic timestepper, the timestepping loops just
+applies the timestepping scheme to the equation set. Any output is
+then decised based on the frequencies specified by the user. For
+example, we output a timeseries of diagnostics, such as minimum and
+maximum values of the fields, every \texttt{nts} timesteps; we output
+the diagnostic fields every \texttt{ndiag} timesteps and we write out
+checkpointing files every \texttt{nchk} timesteps. Checkpointing files
+can be used to initialise a new simulation or to compare to previous
+output.
+
+Algorithms \ref{alg: split physics} and \ref{alg: SIQN} we illustrate
+two more complex timestepping algorithms. These replace lines 4-5 in
+Algorithm \ref{alg: basic}.
 
@@
       \ENDIF
+      \STATE $t \gets t + \Delta t$
     \ENDWHILE
